---
title: "Additional Practice Labs"
---

These labs transfer the core workflow to datasets with different structures.
They are designed for class, homework, or revision rather than as worked
solutions. Begin each lab by writing the source, grain, denominator, and claim
limit before analyzing.

## Lab A: Hong Kong graduate salary averages

```{r}
library(tidyverse)

salary <- read_csv(
  "data/hksalary.csv",
  show_col_types = FALSE
) |>
  rename(
    academic_year = `Academic Year`,
    level = `Level of Study`,
    programme = `Broad Academic Programme Category`,
    average_salary_hkd_thousands = `Average Annual Salary (HK$'000)`
  ) |>
  mutate(
    year_start = as.integer(str_sub(academic_year, 1, 4))
  )

glimpse(salary)
```

### Grain and limitation

One row represents an **already aggregated average salary** for one academic
year, level of study, and broad programme category. The file does not contain
individual graduates or the number of graduates behind each average.

Therefore, this is not a defensible overall annual mean:

```{r}
#| eval: false
salary |>
  summarise(mean_of_means = mean(average_salary_hkd_thousands), .by = year_start)
```

The code gives every programme-level cell equal weight, even if the underlying
numbers of graduates differ greatly. Without counts, a weighted overall average
cannot be reconstructed.

### Tasks

1. Check whether `year_start + level + programme` is unique.
2. Choose one level and compare the same programme categories over time.
3. Identify programme categories that enter or leave the data.
4. Plot like-for-like changes for no more than five programmes.
5. Write a headline that refers to programme-level published averages, not the
   “average Hong Kong graduate.”
6. List the additional data needed to calculate an overall graduate-weighted
   average.

### Required checks

- no duplicated analytical keys;
- same units in every year;
- same programme labels at both comparison endpoints; and
- explicit treatment of missing programme-years.

## Lab B: World Bank life expectancy

The file is in wide form: one row per country or aggregate area, with one column
for each year.

```{r}
life_wide <- read_csv(
  "data/life_expectancy.csv",
  show_col_types = FALSE
)

life_long <- life_wide |>
  select(-matches("^\\.\\.\\.")) |>
  pivot_longer(
    cols = matches("^[0-9]{4}$"),
    names_to = "year",
    values_to = "life_expectancy"
  ) |>
  transmute(
    country = `Country Name`,
    country_code = `Country Code`,
    indicator = `Indicator Name`,
    year = as.integer(year),
    life_expectancy
  )

glimpse(life_long)
```

### Grain and limitation

After reshaping, one row represents one published country-or-aggregate estimate
for one year. The file includes regional and income-group aggregates alongside
countries. Pooling every row as if it were a country would mix incompatible
geographies and double-count populations.

For a beginner comparison, define a small list of actual places explicitly:

```{r}
selected_places <- c(
  "Hong Kong SAR, China",
  "Japan",
  "Singapore",
  "Korea, Rep."
)

life_selected <- life_long |>
  filter(country %in% selected_places)
```

### Tasks

1. Confirm that each selected country-year is unique.
2. Count missing years by country before filtering them out.
3. Compare 2000 and 2022 using absolute change in years.
4. Create a line chart with a common y-axis.
5. Find the largest one-year rise or fall and verify it against the World Bank
   source before describing it as a real event.
6. Explain why pooling all country-years into one correlation or regression
   would confuse differences between countries with changes within countries.

### Required checks

- no aggregate areas in the selected comparison;
- no duplicate country-years;
- the same indicator code across all rows;
- missingness reported by country and year; and
- endpoint values checked against the publisher.

## Submission template

For either lab, submit:

1. **Question** — one sentence;
2. **Source** — publisher, URL, reporting period, and access date;
3. **Grain** — one sentence before and after aggregation;
4. **Method** — the transformations and denominator;
5. **Checks** — at least three machine-readable tests;
6. **Evidence** — one table and one chart; and
7. **Claim limit** — one paragraph on what the file cannot establish.
