---
title: "Practice Labs"
---

## How to use these labs

This optional appendix applies the core tidyverse workflow to new datasets. It
provides starting points and reporting questions rather than complete
solutions. Choose one lab for class, homework, or revision after completing the
main chapters.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `13-practice-labs.Rmd`. Download
[`hksalary.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/hksalary.csv)
and [`life_expectancy.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/life_expectancy.csv),
then save both files in `data/`. You may complete either lab or both.
:::

## Lab A: Hong Kong graduate salary averages

The source columns are long, so rename them after importing:

```{r}
library(tidyverse)

salary <- read_csv("data/hksalary.csv") |>
  rename(
    academic_year = `Academic Year`,
    level = `Level of Study`,
    programme = `Broad Academic Programme Category`,
    average_salary = `Average Annual Salary (HK$'000)`
  )

glimpse(salary)
```

### Understand the rows

One row represents a published **average salary** for one academic year, study
level, and broad programme category. The file does not contain individual
graduates or the number of graduates behind each average.

This means that averaging all rows would create an unweighted mean of published
programme averages. It would not be the average salary of all graduates.

### Tasks

1. Count the rows for each academic year.
2. Choose one study level.
3. Compare the same programme categories over time.
4. Find the five largest published programme averages in the latest year.
5. Create one line or bar chart with an exact measure in the title.
6. List the additional data needed to calculate an overall graduate-weighted
   average.

### Reporting check

Write a headline that refers to published programme-level averages, not “the
average Hong Kong graduate.”

## Lab B: World Bank life expectancy

This file begins in wide form: one row per country or aggregate area, with a
separate column for every year.

```{r}
life_wide <- read_csv("data/life_expectancy.csv")

glimpse(life_wide)
```

Use `pivot_longer()` to put the years into one column. `parse_integer()` turns
the year labels into numbers:

```{r}
life_long <- life_wide |>
  pivot_longer(
    cols = `1960`:`2023`,
    names_to = "year",
    values_to = "life_expectancy"
  ) |>
  rename(
    country = `Country Name`,
    country_code = `Country Code`,
    indicator = `Indicator Name`
  ) |>
  mutate(year = parse_integer(year))

glimpse(life_long)
```

### Understand the rows

After reshaping, one row represents one published country-or-aggregate estimate
in one year. The file contains regional and income-group aggregates as well as
countries, so treating every row as a country would mix different geographies.

For a first comparison, choose a short list of actual places:

```{r}
selected_places <- c(
  "Hong Kong SAR, China",
  "Japan",
  "Singapore",
  "Korea, Rep."
)

life_selected <- life_long |>
  filter(country %in% selected_places)
```

### Tasks

1. Count the available years for each selected country.
2. Check missing values before removing any rows.
3. Compare life expectancy in 2000 and 2022.
4. Calculate the absolute change in years.
5. Create a line chart using the same y-axis for every country.
6. Check the most surprising value against the World Bank source.

### Reporting check

Explain why a change over time is descriptive. The file alone cannot establish
which policy or event caused it.

## Submission template

For either lab, submit:

1. **Question** — one sentence;
2. **Source** — publisher, URL, period, and access date;
3. **Rows** — what one row represents;
4. **Method** — the main filters, groups, and calculation;
5. **Evidence** — one table and one chart; and
6. **Limit** — one thing the data cannot establish.

## Takeaways

| Function or practice | Why it matters |
|---|---|
| `rename()` | Makes long source column names easier to use |
| `pivot_longer()` | Turns many year columns into a tidy year variable |
| `parse_integer()` | Converts year text into whole numbers |
| Fixed comparison set | Keeps the same places or programmes in each period |
| Source check | Confirms that a surprising result is not an import or coding error |

::: {.callout-tip title="Want more control?"}
The official [tidyr reference](https://tidyr.tidyverse.org/reference/) explains
additional reshaping controls. The essential `pivot_longer()` arguments shown
in this lab are enough for the supplied file.
:::
