---
title: "Analyze Change Across Ten Years"
---

## Learning objectives

By the end of this chapter, you should be able to:

- calculate year-to-year and start-to-end changes;
- choose a stable comparison set;
- distinguish market-value change from purchases and performance;
- identify classification and identity breaks; and
- report a trend with appropriate caveats.

## Newsroom question

> How did the scale and geographic distribution of NBIM's reported equity
> holdings change across the rolling ten-year teaching period?

```{r}
library(tidyverse)
library(scales)

gpfg_panel <- readRDS(
  "data/processed/gpfg_equities_last_10_years.rds"
)

panel_years <- sort(unique(gpfg_panel$year))
start_year <- min(panel_years)
end_year <- max(panel_years)

stopifnot(length(panel_years) == 10)
stopifnot(all(diff(panel_years) == 1))
```

## Start with annual coverage and totals

```{r}
annual <- gpfg_panel |>
  summarise(
    holding_records = n(),
    distinct_company_names = n_distinct(company),
    market_value_nok = sum(market_value_nok),
    .by = year
  ) |>
  arrange(year) |>
  mutate(
    annual_change_nok = market_value_nok - lag(market_value_nok),
    annual_percent_change = market_value_nok / lag(market_value_nok) - 1
  )

annual
```

`lag()` compares a row with the preceding row, so the table must be ordered by
year first. The first annual change is missing because the preceding year is
outside the teaching period.

The market value is nominal NOK. A change combines several possible forces:

- security prices;
- exchange rates;
- purchases and sales;
- companies entering or leaving the portfolio;
- corporate events; and
- source classifications or reporting coverage.

The holdings file does not separate these components. Do not describe a rise
in market value as “the amount Norway invested during the year.”

## Calculate start-to-end change

```{r}
start_end <- annual |>
  summarise(
    start_year = start_year,
    end_year = end_year,
    start_value_nok = market_value_nok[year == start_year],
    end_value_nok = market_value_nok[year == end_year],
    absolute_change_nok = end_value_nok - start_value_nok,
    percent_change = end_value_nok / start_value_nok - 1
  )

start_end
```

This is a comparison of two year-end snapshots. It is not a compounded return
calculation.

## Choose a fixed comparison set

If we select the top countries separately in every year, the membership changes
and the lines answer inconsistent questions. Instead, define the comparison
set using the latest snapshot and then trace those same countries backward.

```{r}
top_countries_latest <- gpfg_panel |>
  filter(year == end_year) |>
  summarise(
    market_value_nok = sum(market_value_nok),
    .by = country
  ) |>
  slice_max(market_value_nok, n = 6, with_ties = FALSE) |>
  pull(country)

country_trends <- gpfg_panel |>
  filter(country %in% top_countries_latest) |>
  summarise(
    market_value_nok = sum(market_value_nok),
    holding_records = n(),
    .by = c(year, country)
  ) |>
  arrange(country, year)

country_trends
```

Check completeness:

```{r}
country_trends |>
  count(country, name = "years_present")
```

If a country has fewer than ten years, decide whether the absence is a true
zero, a missing record, a renamed category, or a change in investment coverage.

## Compare shares as well as values

Absolute market value can grow while a country's share falls. Join the annual
denominator to the country table:

```{r}
country_trends <- country_trends |>
  left_join(
    annual |>
      select(year, annual_total_nok = market_value_nok),
    by = "year",
    relationship = "many-to-one"
  ) |>
  mutate(share_of_annual_total = market_value_nok / annual_total_nok)

country_trends
```

The value trend asks “how many NOK?” The share trend asks “how large a portion
of the reported equity portfolio?” A strong story often needs both.

## Audit the industry break

```{r}
gpfg_panel |>
  filter(year %in% c(2020, 2021)) |>
  distinct(year, industry) |>
  arrange(year, industry)
```

The source changes from ten industry labels in 2020 to eleven in 2021.
Consumer Goods and Consumer Services become Consumer Staples and Consumer
Discretionary; Oil & Gas becomes Energy; and Real Estate becomes a separate
category.

Do not draw a continuous industry line across this break without either:

1. building and documenting a concordance to broader categories; or
2. presenting the pre-2021 and 2021-onward years as separate classification periods.

For a beginner assignment, the second option is easier to defend.

## Audit company-name persistence

```{r}
company_persistence <- gpfg_panel |>
  distinct(year, company) |>
  count(company, name = "years_present")

company_persistence |>
  count(years_present, name = "company_names") |>
  arrange(years_present)
```

Only a minority of published names appear in all ten years. That pattern mixes
real portfolio entry and exit with renaming and corporate events. A company
trend is therefore an advanced verification task, not a simple join on text.

## Reporting a defensible trend

A concise methods paragraph could read:

> We analyzed NBIM's year-end listed-equity holdings for `r start_year`–`r end_year` and summed
> nominal market values in Norwegian kroner by reporting year and investment
> market. The values describe holdings at each year-end and cannot be separated
> into price, currency, purchase, and sale effects. Company-level matching was
> not used because published names are not stable identifiers.

## Independent task

Choose one country from the fixed comparison set and produce a short trend
memo. Include:

1. annual NOK value and share of total;
2. start-to-end absolute and percentage change;
3. the largest annual rise and fall;
4. a check that all ten years are present;
5. one external reporting question that the holdings data cannot answer; and
6. a headline that describes the evidence without implying purchases, returns,
   or causation.
