# Data Analysis II

## Overview 

In this chapter, we will use a different dataset to demonstrate the three types of questions that are often asked in journalistic reporting.

### About the data

This dataset was downloaded from the World Bank website and saved as a CSV file named `life_expectancy.csv`, stored in the `data` folder. The dataset was last updated on March 24, 2025, contains the life expectancy at birth for various countries over the years.

### Learning Objectives

This chapter will demonstrate three types of data analysis:

-  Single Variable Analysis: Analyzing the distribution of global life expectancy.
-  Time-Based Analysis: Tracking changes in China's life expectancy over time.
-  Group Comparisons: Comparing life expectancy across different countries.

## Load data and packages

```{r}
library(tidyverse)
lifex <- read_csv("data/life_expectancy.csv")
```

## Data Wrangling
```{r}
lifex_clean <- lifex |> 
  select(-`Indicator Name`, -`Indicator Code`, -`Country Code`) |>
  pivot_longer(cols = -`Country Name`, 
               names_to = "year", 
               values_to = "life_expectancy") |>
  mutate(year = as.numeric(year)) |>
  select(country = `Country Name`, year, life_expectancy) |>
  drop_na()

glimpse(lifex_clean)
```



## Single Variable Analysis: Understanding Global Life Expectancy Distribution

```{r}
# Summary statistics of life expectancy
lifex_clean |> 
  summarise(
    avg = mean(life_expectancy, na.rm = TRUE),
    median = median(life_expectancy, na.rm = TRUE),
    min = min(life_expectancy, na.rm = TRUE),
    max = max(life_expectancy, na.rm = TRUE)
  )
```

```{r}
# visualize the distribution of life expectancy
hist(lifex_clean$life_expectancy, 
     main = "Global Life Expectancy Distribution",
     xlab = "Life Expectancy",
     col = "skyblue",
     border = "black")
```
## Time-Based Analysis: Tracking Changes in China's Life Expectancy

```{r}
lifex_clean |>
  filter(country == "China") |>
  ggplot(aes(x = year, y = life_expectancy)) +
  geom_line(color = "steelblue") +
  labs(title = "Life Expectancy in China Over Time",
       x = "Year",
       y = "Life Expectancy") +
  theme_minimal()
```
## Group Comparisons: Comparing Life Expectancy Across Countries

```{r}
lifex_clean |>
  filter(year == 2022) |>
  top_n(10, life_expectancy) |>
  ggplot(aes(x = reorder(country, life_expectancy), y = life_expectancy)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(title = "Life Expectancy Across Countries",
       x = "Country",
       y = "Life Expectancy") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))
```

