---
title: "Analysis I"
---

::: {.callout-note title="Part 3: Data Analysis"}
Chapter 4 produced a clean 2025 table. We can already use that single annual
snapshot to describe values, rank holdings, and compare groups. Chapter 6 will
add several years so that we can also investigate change over time.
:::

## Learning objectives

By the end of this chapter, you should be able to:

- turn a broad topic into a precise reporting question;
- inspect a clean table before analysing it;
- distinguish description, ranking, comparison, relationship, and trend
  questions;
- calculate summaries with `summarise()`;
- rank rows with `arrange()`;
- compare groups with `group_by()` and `summarise()`; and
- turn a calculated result into a carefully worded finding.

## Continue with the clean 2025 data

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `05-data-analysis-i.Rmd`. Reuse the
`data/gpfg_2025.csv` file created in Chapter 4. A ready-made
[`gpfg_2025.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2025.csv)
is available if needed.
:::

Start the new document by loading the tidyverse and importing the clean file:

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_2025.csv")
```

One row represents one equity holding reported by NBIM at the end of 2025. A
row is not a purchase, sale, investment return, or payment received by a
company.

Here, a **holding** means the fund's reported ownership stake in a company at
the reporting date. Saying that a company appears in the data does not mean
that it received cash from Norway in 2025.

## A question-to-answer workflow

Use the same sequence for every analysis:

1. State what one row represents.
2. Write a precise question naming the measure, group, and time period.
3. Identify the columns needed to answer it.
4. Check their types, missing values, units, and definitions.
5. Calculate and inspect the result.
6. Verify important or surprising values.
7. Write only what the evidence supports.

The function should follow from the question, not the other way around.

## Know the dataset before asking questions

Recheck the clean table:

```{r}
glimpse(gpfg)
```

The output should show 7,201 rows, 11 columns, convenient names, numeric value
columns, and a real date column.

`distinct(column)` keeps each observed value in a column once.
`arrange(column)` then orders rows from smallest to largest or, for text,
alphabetically. Use both to inspect the available region labels:

```{r}
gpfg |>
  distinct(region) |>
  arrange(region)
```

The result is a short alphabetical list rather than thousands of repeated
region labels.

`count(region)` returns one row per region and creates a count column named
`n`. `arrange(desc(n))` sorts that column in descending order, with the
largest count first:

```{r}
gpfg |>
  count(region) |>
  arrange(desc(n))
```

The result describes the number of rows. It does not yet tell us which region
has the largest total market value. Count and amount answer different
questions.

## Types of questions in data journalism

| Question type | What it asks | Example using the 2025 data |
|---|---|---|
| Description | How much is there, or what is typical? | Across NBIM's 2025 company holdings, what were the total and median market values? |
| Ranking | Which observations are largest or smallest? | Which companies had the largest reported holding values in 2025? |
| Comparison | How do categories differ? | Which investment markets accounted for the largest 2025 totals? |
| Relationship | Do two measures vary together? | Were higher ownership percentages associated with larger holding values? |
| Trend | How did a measure change over time? | How did NBIM's reported equity-holdings total change from 2021 to 2025? |

The clean 2025 file can answer the first four types because it contains many
holdings, categories, and numeric measures. It cannot answer a trend question:
one year provides no earlier or later observation for comparison. Chapter 6
will solve that limitation.

## Ask a descriptive question

> Across the company holdings reported by NBIM at the end of 2025, what were
> the total and median market values?

`summarise()` reduces many rows to calculated summary values. Its basic form
is `summarise(new_name = calculation)`: the name on the left labels the result
column, and the calculation on the right produces its value.

```{r}
holding_summary <- gpfg |>
  summarise(
    total_market_value_nok = sum(market_value_nok),
    average_market_value_nok = mean(market_value_nok),
    median_market_value_nok = median(market_value_nok),
    smallest_market_value_nok = min(market_value_nok),
    largest_market_value_nok = max(market_value_nok)
  )

holding_summary
```

One row in `holding_summary` describes the full 2025 table. Financial values
are often uneven, so compare the mean and median before calling either value
“typical.” A few very large holdings can pull the mean upward.

## Ask a ranking question

> Which ten companies in NBIM's 2025 equity-holdings data had the largest
> reported holding values, measured in NOK?

The question needs the company, market, industry, and NOK value.
`select(column1, column2)` keeps the named columns. `arrange(column)` sorts
from smallest to largest, while `arrange(desc(column))` sorts from largest to
smallest. After sorting, `slice_head(n = 10)` keeps the first ten rows:

```{r}
largest_holdings <- gpfg |>
  select(company, country, industry, market_value_nok) |>
  arrange(desc(market_value_nok)) |>
  slice_head(n = 10)

largest_holdings
```

`slice_head(n = 10)` keeps the first ten rows after sorting. A ranking
identifies large observations; it does not explain why they are large. Verify
important rows against the source before publication.

## Ask a comparison question

> Which investment markets accounted for the largest total market value of
> NBIM's reported equity holdings at the end of 2025?

This question needs one total per country. `group_by(country)` tells dplyr to
treat each country as a separate group; it does not calculate a result by
itself. The following `summarise()` returns one row for each group. Inside it,
`n()` counts rows and `sum()` adds their values:

```{r}
country_summary <- gpfg |>
  group_by(country) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(desc(market_value_nok))

country_summary |>
  slice_head(n = 10)
```

Track the changing row meaning:

- before `summarise()`, one row represents one holding;
- after `summarise()`, one row represents one investment market.

Record count and total value need not produce the same ranking.

### Calculate a share

`mutate()` adds or changes columns while keeping the existing rows. Its basic
form is `mutate(new_column = calculation)`. Use it to add each market's
percentage share of the full reported value:

```{r}
country_summary <- country_summary |>
  mutate(
    share_percent = market_value_nok / sum(market_value_nok) * 100
  )

country_summary |>
  slice_head(n = 10)
```

The denominator is the sum of the country totals. The resulting percentage
answers: “What share of the total market value in NBIM's 2025 equity-holdings
data is associated with this investment market?”

## Prepare a relationship question

> Among companies in NBIM's 2025 equity-holdings data, were higher ownership
> percentages associated with larger reported holding values?

The required columns already exist in the single-year file:

```{r}
relationship_data <- gpfg |>
  select(company, market_value_usd, ownership_pct)

relationship_data |>
  slice_head(n = 5)
```

A relationship is difficult to judge from thousands of printed rows. Chapter
7 will turn these two numeric columns into a scatterplot. Preparing and
checking the relevant table still comes before making the chart.

## From a result to a finding

Compare these statements:

- Too broad: “The United States received the most Norwegian investment.”
- Better: “The United States had the largest reported market value in NBIM's
  2025 year-end equity-holdings file.”

The second statement matches the measure, time, and source. Explanation
requires additional reporting beyond the calculated table.

## What this chapter cannot answer

The 2025 table cannot tell us whether a value increased, decreased, or stayed
stable. A trend requires the same measure at multiple points in time. Chapter
6 will import five annual snapshots, combine their rows, and add that missing
time dimension.

## Practice

1. Use `distinct()` to inspect the industry labels.
2. Use `count()` to find the most frequently occurring industries.
3. Calculate the total, mean, and median USD value of the company holdings.
4. Rank the ten largest company holdings by USD value.
5. Compare total market value and record count across industries.
6. Write one supported finding and one explanatory question requiring more
   evidence.
7. Write one trend question that this single-year file cannot answer.

## Takeaways

| Function or idea | What it does |
|---|---|
| `distinct()` | Shows the unique values present in a column |
| `count()` | Counts rows within categories |
| `summarise()` | Reduces rows to calculated summary values |
| `select()` | Keeps the columns required for a question |
| `arrange()` + `desc()` | Orders rows from largest to smallest |
| `slice_head()` | Keeps the first rows after ordering |
| `group_by()` + `summarise()` | Produces one summary row per group |
| `n()` | Counts rows inside each group |
| `mutate()` | Adds a derived measure such as percentage share |

::: {.callout-tip title="Want more control?"}
The official [dplyr reference](https://dplyr.tidyverse.org/reference/index.html)
contains additional summary and ranking tools. Add one only when a reporting
question requires it.
:::
