---
title: "Find Patterns in a Single Snapshot"
---

## Learning objectives

By the end of this chapter, you should be able to:

- move from a broad question to testable measures;
- describe a skewed financial distribution with appropriate statistics;
- distinguish rankings, shares, and concentration;
- inspect outliers in context; and
- write a claim that matches the evidence.

## Newsroom questions

We will investigate four related questions about the latest-year snapshot:

1. How unequal is the distribution of holding values?
2. Which countries and industries account for the largest shares?
3. Which individual holdings are largest?
4. Does a large market value necessarily mean a large ownership stake?

```{r}
library(tidyverse)
library(scales)

gpfg <- read_csv(
  "data/processed/gpfg_equities_latest.csv",
  show_col_types = FALSE
)
```

The grain remains one reported equity holding at the latest year-end.

## Describe the distribution

Start with totals and robust summaries:

```{r}
holding_summary <- gpfg |>
  summarise(
    holding_records = n(),
    total_value_nok = sum(market_value_nok),
    mean_value_nok = mean(market_value_nok),
    median_value_nok = median(market_value_nok),
    p90_value_nok = quantile(market_value_nok, 0.90),
    maximum_value_nok = max(market_value_nok)
  )

holding_summary
```

In a highly skewed distribution, the mean is pulled upward by a small number of
very large holdings. Report the median and at least one upper percentile rather
than presenting the mean as “typical.”

```{r}
gpfg |>
  summarise(
    p25 = quantile(market_value_nok, 0.25),
    p50 = quantile(market_value_nok, 0.50),
    p75 = quantile(market_value_nok, 0.75),
    p90 = quantile(market_value_nok, 0.90),
    p99 = quantile(market_value_nok, 0.99)
  )
```

Quantiles answer a precise question: the 90th percentile is the value at or
below which 90% of holding records fall. It is not the value of the “average
company.”

## Rank individual holdings

```{r}
largest_holdings <- gpfg |>
  select(company, country, industry, market_value_nok, ownership_pct) |>
  arrange(desc(market_value_nok)) |>
  slice_head(n = 15)

largest_holdings
```

Before publishing a ranking, search the source for several top rows and check
company names, units, and whether a name describes a fund or another listed
vehicle.

Now calculate concentration with an explicit ranking and denominator:

```{r}
ranked_holdings <- gpfg |>
  arrange(desc(market_value_nok)) |>
  mutate(rank = row_number())

total_nok <- sum(ranked_holdings$market_value_nok)

ranked_holdings |>
  summarise(
    top_10_share = sum(market_value_nok[rank <= 10]) / total_nok,
    top_100_share = sum(market_value_nok[rank <= 100]) / total_nok
  )
```

Prefer transparent intermediate objects when the result will support a major
claim.

## Compare countries with a denominator

```{r}
country_summary <- gpfg |>
  summarise(
    market_value_nok = sum(market_value_nok),
    companies = n_distinct(company),
    median_holding_nok = median(market_value_nok),
    .by = country
  ) |>
  mutate(share_of_total = market_value_nok / sum(market_value_nok)) |>
  arrange(desc(market_value_nok)) |>
  mutate(cumulative_share = cumsum(share_of_total))

stopifnot(isTRUE(all.equal(sum(country_summary$share_of_total), 1)))
```

`cumsum()` uses the current row order, so it must come after sorting. Code can
run without an error and still answer the wrong question; order is part of the
method.

This deliberate correction illustrates an important point: code can run and
still answer the wrong question.

## Compare industries

```{r}
industry_summary <- gpfg |>
  summarise(
    market_value_nok = sum(market_value_nok),
    holding_records = n(),
    median_ownership_pct = median(ownership_pct),
    .by = industry
  ) |>
  mutate(share_of_total = market_value_nok / sum(market_value_nok)) |>
  arrange(desc(market_value_nok))

industry_summary
```

Industry share and number of holdings answer different questions. A sector can
have many relatively small holdings or fewer very large ones.

## Market value and ownership are different measures

```{r}
gpfg |>
  summarise(
    pearson = cor(market_value_nok, ownership_pct),
    spearman = cor(market_value_nok, ownership_pct, method = "spearman")
  )
```

Pearson correlation is sensitive to extreme values. Spearman correlation asks
whether higher ranks on one measure tend to accompany higher ranks on the
other. Neither correlation proves that ownership causes market value or that
the fund “prefers” one company.

Inspect exceptions directly:

```{r}
gpfg |>
  filter(ownership_pct >= quantile(ownership_pct, 0.99)) |>
  arrange(desc(ownership_pct)) |>
  select(company, country, industry, market_value_nok, ownership_pct) |>
  slice_head(n = 15)
```

## From observation to claim

Use a claim ladder:

| Level | Example | Evidence needed |
|---|---|---|
| Description | “Country A had the largest reported market value.” | Correct aggregation and source check |
| Comparison | “Country A's share was twice Country B's.” | Shared denominator and comparable categories |
| Trend | “The value rose over time.” | Multiple comparable snapshots |
| Explanation | “The value rose because the fund bought more.” | Transaction or attribution evidence not present here |

The single-year file supports the first two levels. It cannot explain changes over
time, because it contains only one snapshot.

## Independent task

Choose either country or industry as the analytical grain. Produce a short
newsroom memo with:

1. a ranked table;
2. a share-of-total calculation;
3. a median and a 90th percentile;
4. two checks that protect against double-counting; and
5. a three-sentence finding: one descriptive claim, one comparison, and one
   explicitly rejected causal claim.
