---
title: "Analysis II"
---

## Learning objectives

By the end of this chapter, you should be able to:

- explain why trend questions require multiple time points;
- import several annual CSV files;
- compare their names and types before combining them;
- combine tables with the same columns using `bind_rows()`;
- inspect time coverage, record counts, categories, and missing values;
- calculate annual and country-year summaries; and
- interpret changes without confusing holdings with flows or returns.

## Add the missing time dimension

Chapter 5 used the clean 2025 snapshot to answer description, ranking,
comparison, and relationship questions. It could not answer:

> How did the total reported market value of NBIM's equity holdings change
> from 2021 to 2025?

A trend needs the same measure at several dates. We will add four earlier
year-end snapshots and analyse the five-year period from 2021 through 2025.
Five files provide enough time points for a trend while keeping the import
exercise manageable.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `06-data-analysis-ii.Rmd`. Keep the
`gpfg_2025.csv` created in Chapter 4. Download
[`gpfg_2021.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2021.csv),
[`gpfg_2022.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2022.csv),
[`gpfg_2023.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2023.csv),
and [`gpfg_2024.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2024.csv)
into the same `data` folder. A ready-made
[`gpfg_2025.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2025.csv)
is also available if needed.
:::

## Import the annual files

Repeat the local `read_csv()` pattern from Chapter 3:

```{r}
library(tidyverse)

gpfg_2021 <- read_csv("data/gpfg_2021.csv")
gpfg_2022 <- read_csv("data/gpfg_2022.csv")
gpfg_2023 <- read_csv("data/gpfg_2023.csv")
gpfg_2024 <- read_csv("data/gpfg_2024.csv")
gpfg_2025 <- read_csv("data/gpfg_2025.csv")
```

The object names record which annual snapshot each table contains.

## Check the structures before combining

Tables should describe the same variables before we place their rows together.
Compare an early and late file:

```{r}
glimpse(gpfg_2021)
```

```{r}
glimpse(gpfg_2025)
```

The column names and types should match. Each file already contains a `year`
column, so observations will remain distinguishable after combination.

::: {.callout-important title="Do not combine first and inspect later"}
If annual files use different names, types, units, or definitions, stacking
them can create a misleading table. Resolve meaningful differences before
analysing a trend.
:::

## Combine annual rows with `bind_rows()`

The five tables contain the same columns but different annual observations.
`bind_rows()` places their rows underneath one another:

```{r}
gpfg <- bind_rows(
  gpfg_2021,
  gpfg_2022,
  gpfg_2023,
  gpfg_2024,
  gpfg_2025
)
```

One row still represents one reported holding in one year-end snapshot. The
operation adds observations; it does not add new variables.

This is different from a join. Chapter 9 will use a shared key to add
variables from another table.

## Check the combined result

### Are all expected years present?

```{r}
gpfg |>
  count(year)
```

The result should contain five consecutive years. The number of reported
holding records differs by year, so changes in totals should not be interpreted
without considering changes in coverage.

### What does one row represent now?

```{r}
glimpse(gpfg)
```

The same company can appear in several years. A row is therefore not a unique
company across the full period, and the table does not contain transactions.

### Are important values missing?

```{r}
gpfg |>
  filter(is.na(company) | is.na(market_value_nok))
```

The `|` symbol means “or.” An empty result confirms that both fields are
complete in the supplied annual files, although it does not prove that every
published value is correct.

### Do category labels look consistent?

```{r}
gpfg |>
  distinct(region) |>
  arrange(region)
```

For longer historical periods, country and industry definitions may change.
Never assume that identically named columns are perfectly comparable across
time without checking the source documentation.

## Save the five-year table

Later chapters can begin from the completed combination step:

```{r}
write_csv(gpfg, "data/gpfg_5_years.csv")
```

The repository also provides a ready-made
[`gpfg_5_years.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_5_years.csv)
for learners beginning with a visualization chapter.

## Ask an overall trend question

> How did the total reported market value of NBIM's equity holdings change
> from 2021 through 2025?

The question needs `year` and `market_value_nok`. Group the observations by
year and return one summary row for each annual snapshot:

```{r}
annual_summary <- gpfg |>
  group_by(year) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(year)

annual_summary
```

Track the changing row meaning:

- before `summarise()`, one row represents one holding in one year;
- after `summarise()`, one row represents one annual snapshot.

Keeping both record count and market value visible helps us notice that a
change in the total occurs alongside a changing number of reported holdings.

## Combine trend and comparison

We can also ask:

> How did the total reported market value of NBIM's equity holdings in the
> United States, China, and Japan change from 2021 through 2025?

Create a vector containing the three selected markets, then use `%in%` to keep
rows whose country appears in that vector:

```{r}
selected_countries <- c("United States", "China", "Japan")

country_trends <- gpfg |>
  filter(country %in% selected_countries) |>
  group_by(year, country) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(country, year)

country_trends
```

One row now represents one selected investment market in one year. `%in%`
asks whether each country belongs to `selected_countries`.

This table combines a comparison question—differences among markets—with a
trend question—change across years. Chapter 8 will display the three series
with separate colored lines.

## Interpret trends carefully

The annual values are year-end market values of reported holdings. A change
can reflect:

- security prices;
- exchange rates;
- purchases or sales;
- companies entering or leaving the portfolio; or
- changes in reporting coverage or classification.

The table does not isolate those causes. Describe the observed pattern first,
then seek additional evidence before explaining it. Use “year-end market
value,” not “money invested that year” or “investment return.”

::: {.callout-note title="The complete history is optional"}
`data/gpfg_history.csv.gz` contains the full 1998--2025 archive. It is useful
for advanced reporting, but the longer period includes major coverage and
classification changes. The five-year table is the main beginner dataset.
:::

## Questions now available

| Question | Single year needed? | Multiple years needed? |
|---|:---:|:---:|
| What was the median value of an NBIM company holding in 2025? | Yes | No |
| Which investment markets accounted for the largest 2025 totals? | Yes | No |
| Were ownership percentage and holding value associated in 2025? | Yes | No |
| Did the total value of NBIM's reported equity holdings rise or fall? | No | Yes |
| Did the selected investment markets follow similar paths? | No | Yes |

More data do not replace a precise question. Use multiple years only when time
is part of the question.

## Practice

1. Import the five annual files and compare the structures of 2021 and 2025.
2. Combine them and confirm the available years with `count()`.
3. Calculate one row per year containing record count and total USD value.
4. Choose one country and calculate its annual NOK totals.
5. Compare two industries across the five years.
6. Write one supported trend finding and two possible explanations requiring
   additional reporting.

## Takeaways

| Function or idea | What it does |
|---|---|
| `read_csv()` | Imports each annual CSV |
| `glimpse()` | Confirms that annual names and types match |
| `bind_rows()` | Places observations from similar tables underneath one another |
| `count(year)` | Checks year coverage and annual record counts |
| `write_csv()` | Saves the combined result for later chapters |
| `group_by(year)` + `summarise()` | Produces one summary row per year |
| `%in%` | Tests whether a value belongs to a selected vector |
| `group_by(year, country)` | Defines country-year groups for a comparative trend |

Chapter 7 will turn the checked annual summary into a line chart and the 2025
country comparison into a bar chart.
