---
title: "Data Import"
---

::: {.callout-note title="Part 2: Data Import & Cleaning"}
In this part, you will bring data into R and prepare it for analysis. We begin
with one CSV file and add new tidyverse tools gradually.
:::

## Learning objectives

By the end of this chapter, you should be able to:

- explain what a CSV file is;
- explain the relationship between the tidyverse, readr, and `read_csv()`;
- compare package data, URL imports, and local-file imports;
- import a CSV file with a relative path;
- explain the main parts of `glimpse()` output; and
- identify common issues to investigate after an import.

## Why import data?

So far, we have typed small examples directly into R. In reporting, the data
usually come from somewhere else: a government website, an institution, a
researcher, or a file supplied by a source. **Importing** brings that file into
R so we can work with it.

We will use a **CSV**, or comma-separated values file. A CSV stores a table as
plain text. It is widely used, easy to share, and a good format for learning
data analysis.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `02-data-import.Rmd`. Download
[`gpfg.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg.csv)
and save it inside your project's `data` folder. Do not open and resave the
file in spreadsheet software.
:::

## Tidyverse, readr, and `read_csv()`

The **tidyverse** is a collection of packages that work together. One of those
packages is **readr**, which imports rectangular data files. The readr function
for a CSV file is `read_csv()`.

Loading the tidyverse also loads readr:

```{r}
library(tidyverse)
```

Because readr is one of the core tidyverse packages, you do **not** also need
to run `library(readr)`. The single `library(tidyverse)` line makes
`read_csv()` and the other core tidyverse tools available. If a project used
readr without the rest of the tidyverse, it could load only readr with
`library(readr)`, but that is not the workflow used in this book.

The simplest form of the import function is:

```r
read_csv("file_path.csv")
```

The function needs the path to the file. Because a path is text, it goes inside
quotation marks.

::: {.callout-important title="Use `read_csv()`, not `read.csv()`"}
Both functions can import a CSV, but they come from different parts of R:

| Function | Where it comes from | What it returns |
|---|---|---|
| `read_csv()` | readr and the tidyverse | A tibble |
| `read.csv()` | Base R | A traditional data frame |

This book uses `read_csv()` for four reasons:

1. It returns a tibble, the table format used throughout the tidyverse.
2. It reports the column types it detected, giving us immediate information
   about the import.
3. It records values that it could not convert correctly, which we can inspect
   with `problems()`.
4. Its name and behavior fit consistently with other tidyverse functions used
   in the book.

`read.csv()` is valid R code, but it belongs to a different workflow and its
name uses a period instead of an underscore. Using one approach consistently
makes the course easier to follow. Check the function name carefully:
`read_csv()` is the one used here.
:::

## Package data, URLs, and local files

Data can enter R in several ways. Three common possibilities are package data,
a direct URL, and a file stored on your computer.

### Data supplied by a package

Some R packages include small datasets for examples and practice. For example,
the ggplot2 package includes a table called `mpg`. It becomes available when
the tidyverse is loaded:

```r
library(tidyverse)
mpg
```

Package datasets are useful for learning because everyone with the same
package receives the same data. In journalism, however, the main source is
more often a file published by an organization or obtained through reporting.

### Data imported from a URL

`read_csv()` can read a public CSV directly from its web address:

```r
gpfg_online <- read_csv(
  "https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg.csv"
)
```

This can be convenient for a quick exploration or an automated update. It also
means that R must download the file again whenever the code runs.

### Data imported from a local file

For the main course workflow, download the source file once, keep it unchanged
inside the Project, and import the local copy. This is safer for a reproducible
analysis because:

- a website or URL may move, disappear, or become temporarily unavailable;
- the contents at the same URL may be replaced without warning;
- an internet connection may fail when the document is knitted;
- a website may limit or block repeated downloads; and
- the local file preserves the exact version used for the story.

Record the original URL and the date of download in your notes. A local copy is
not a replacement for documenting the source; it is an archived snapshot of
what the source provided at that time.

::: {.callout-important title="The workflow used in this book"}
The book mentions package data and URL imports so you can recognize them. The
guided lessons use local CSV files stored in `data/`, which makes the code more
reliable and allows the same analysis to run again later.
:::

## Relative paths and RStudio Projects

Our file is stored here inside the learner project:

```text
djr/
├── data/
│   └── gpfg.csv
├── 02-data-import.Rmd
└── djr.Rproj
```

Starting from the project folder, the relative path is:

```text
data/gpfg.csv
```

This is why we open the `.Rproj` file before working. Everyone who opens the
same project starts in the same place, so the same relative path works on
different computers.

::: {.callout-tip title="Avoid personal paths"}
Do not use a path such as `/Users/name/Desktop/...` or
`C:/Users/name/Desktop/...` in shared code. It points to one person's
computer. Keep the data inside the project and use a relative path.
:::

## Import the CSV

The teaching file comes from the equity-holdings data published by
[Norges Bank Investment Management](https://www.nbim.no/en/investments/all-investments/).
It is a prepared CSV containing one year of reported holdings.

Import it and save the result as `gpfg`:

```{r}
gpfg <- read_csv("data/gpfg.csv")
```

Read the line from right to left: `read_csv()` reads the file, and `<-` saves
the imported table with the name `gpfg`.

readr may display a message describing the rows, columns, and column types it
found. This is information about the import, not an error.

## Inspect the imported data

Never begin an analysis without looking at what was imported.

### Look at the structure

`glimpse()` shows the number of rows and columns, the column names, their data
types, and a few example values:

```{r}
glimpse(gpfg)
```

Read the output from top to bottom:

- **`Rows: 7,201`** tells us how many observations were imported.
- **`Columns: 13`** tells us how many variables were imported.
- Each line beginning with **`$`** describes one column.
- The column name appears after `$`, such as `company` or `market_value_usd`.
- The short label in angle brackets is the data type. For example, `<chr>`
  means text, `<dbl>` means a number, and `<date>` means a date.
- The values at the end of the line are examples from that column.

For example, a line beginning `$ company <chr>` tells us that `company` is a
text column. A line beginning `$ market_value_usd <dbl>` tells us that
`market_value_usd` is numeric.

`glimpse()` is a quick overview. It shows only a few example values, so it
cannot prove that every value is valid or tell us how many values are missing.

### Preview a few rows

`slice_head()` shows rows from the beginning of a table. The `n` argument says
how many rows to show:

```{r}
gpfg |>
  slice_head(n = 5)
```

The result contains the same columns as `gpfg`, but only its first five rows.
The original `gpfg` object is not changed.

::: {.callout-note title="Why use `slice_head()` instead of `head()`?"}
Both functions can show the first rows of a table, and `head(gpfg)` is valid R
code. This book uses `slice_head()` because it belongs to dplyr, works naturally
in a tidyverse pipeline, and makes the action explicit. In
`slice_head(n = 5)`, the named argument `n` clearly says how many rows to show.
:::

### What is a parsing problem?

A CSV stores values as text. During import, `read_csv()` tries to convert that
text into suitable R values, such as numbers or dates. This conversion is
called **parsing**.

Imagine that a column should contain numbers:

```text
amount
120
unknown
95
```

The value `unknown` cannot be converted into a number. This is a **parsing
problem**. Depending on the column type, the value may become `NA`, and readr
records where the problem occurred.

After importing a file, use `problems()` to check whether readr recorded any
such values:

```{r}
problems(gpfg)
```

If problems are found, the result identifies the row and column, what readr
expected, and the value it actually found. An empty result means readr did not
detect a parsing problem. It does not prove that every value in the source is
correct—for example, an incorrect but valid number would not be reported.

## What should you inspect?

Importing a file successfully does not guarantee that the data are ready for
analysis. Ask these questions before calculating anything:

| Check | Why it matters |
|---|---|
| Number of rows and columns | An incomplete download or wrong file may have an unexpected size |
| Column names | Names should identify the variables clearly |
| Data types | A numeric column imported as `<chr>` may contain symbols or unexpected text |
| Missing values | `NA` means a value is unavailable; it is not the same as zero |
| Unexpected values | Negative amounts, impossible percentages, or unusual spellings may need investigation |
| Repeated rows | A duplicate may be an error, or it may represent a legitimate repeated observation |
| Units and definitions | A number cannot be interpreted correctly without knowing what it measures |
| Parsing problems | A parsing problem means readr could not interpret a value as expected |

At this stage, use `glimpse()`, `slice_head()`, `problems()`, and the data
dictionary to form questions about the table. In the next chapter, we will use
tidyverse functions to investigate missing, unexpected, and repeated values
more systematically.

## Refer to the data dictionary

Before using a column, check what it means and what unit it uses. Open the
[GPFG data dictionary](https://github.com/binchen19/djr/blob/main/data/gpfg_data_dictionary.csv)
and keep it as a reference. You do **not** need to import the dictionary into
R for this lesson.

For example, the dictionary explains the difference between market value in
Norwegian kroner and US dollars, and it explains that ownership values are
percentage points.

One row in the imported file represents one reported equity holding at the end
of the year. It does not represent a purchase or sale.

## Practice

In an R Markdown document:

1. load the tidyverse;
2. import `data/gpfg.csv` as `gpfg`;
3. inspect it with `glimpse()`;
4. show its first five rows; and
5. check for parsing problems.

Then write five short answers:

1. How many rows and columns were imported?
2. What data types do `company` and `market_value_usd` have?
3. What does one row represent?
4. Name one important issue that `glimpse()` alone cannot fully check.
5. Give one reason for importing the saved local file instead of downloading
   it from the URL every time the document runs.

## Takeaways

| Function or idea | What it does |
|---|---|
| `library(tidyverse)` | Loads the core tidyverse packages |
| `read_csv("path")` | Imports a CSV file as a tibble |
| Relative path | Locates a file from the RStudio Project folder |
| Local source copy | Preserves the version used in the analysis |
| `glimpse()` | Shows columns, types, and example values |
| `slice_head(n = 5)` | Shows the first five rows |
| `problems()` | Reports values readr could not parse as expected |

::: {.callout-tip title="Want more control?"}
The file path is the only argument needed in this lesson. The official
[`read_csv()` reference](https://readr.tidyverse.org/reference/read_delim.html)
explains optional controls for missing values, column types, delimiters, and
unusual files when you need them later.
:::
