---
title: "Data Cleaning"
---

## Learning objectives

By the end of this chapter, you should be able to:

- turn observations from the import chapter into a short cleaning plan;
- rename awkward columns with `rename()`;
- change a date column to the correct type with `mutate()` and `dmy()`;
- inspect and handle missing values;
- identify and remove a confirmed duplicate row; and
- save a clean CSV without overwriting the imported source.

## Continue with the imported table

Chapter 3 imported `gpfg_messy.csv` as `gpfg_raw`. Most of the table arrived
correctly: the amounts and percentages are already numeric, and the category
values are consistent. We only need to solve four visible problems:

1. a few column names are awkward;
2. the reporting date is stored as text;
3. one incomplete row does not describe a holding; and
4. one complete row has been repeated.

This smaller set of problems lets us practise a cleaning workflow without
changing every part of the data at once.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `04-data-cleaning.Rmd`. Reuse
`data/gpfg_messy.csv`; do not download another copy.
:::

Start the new document by recreating the object. Each R Markdown document must
contain the code needed to run on its own:

```{r}
library(tidyverse)

gpfg_raw <- read_csv("data/gpfg_messy.csv")
```

One row is intended to represent one reported equity holding at the end of
2025.

## Turn observations into a cleaning plan

Run the inspection from Chapter 3 again:

```{r}
glimpse(gpfg_raw)
```

Translate what we see into actions:

| Observation | Cleaning action |
|---|---|
| A few column names contain spaces and capitals | Rename those columns |
| `Report Date` is `<chr>` text | Convert it to a date |
| The file has a row without a company or value | Inspect the missing row |
| The file has one more complete row than expected | Check for a duplicate |

Choose a function only after identifying the problem it needs to solve.

## Rename the awkward columns

Use `rename()` with `new_name = old_name`. An old name containing spaces needs
backticks:

```{r}
gpfg <- gpfg_raw |>
  rename(
    year = Year,
    report_date = `Report Date`,
    company = `Company Name`,
    market_value_nok = `Market Value NOK`
  )

glimpse(gpfg)
```

The new names use lowercase `snake_case`, the naming style introduced in
Chapter 2. The other column names were already suitable, so we did not rename
them. Renaming changes labels, not values.

`year` appears as `<dbl>`, which is still a numeric type and is suitable for
these whole-year values. We do not need to convert a column merely to make its
type label look different.

## Convert the date type

The values in `report_date` use day-month-year order, such as `31/12/2025`,
but R imported the column as `<chr>` text. The data source tells us what the
order means.

Use `mutate()` to change the existing column. `dmy()` means day, month, year:

```{r}
gpfg <- gpfg |>
  mutate(report_date = dmy(report_date))

glimpse(gpfg)
```

The type should now appear as `<date>`. `mutate()` can create a new column or
change an existing one; here it replaces the text version with a date version.

::: {.callout-important title="Confirm the date order"}
Do not guess whether `03/04/2025` means 3 April or March 4. Check the source
documentation. Use `dmy()` for day-month-year, `mdy()` for month-day-year, or
`ymd()` for year-month-day.
:::

## Check missing values

R represents unavailable information with `NA`. A missing value differs from
zero: zero is a known amount, while `NA` means the value is unavailable.

Use `is.na()` inside `filter()` to see rows with a missing company name:

```{r}
gpfg |>
  filter(is.na(company))
```

The result contains one deliberately incomplete row. It has a year and date
but no company, country, industry, or market value, so it cannot represent the
holding described by the data dictionary. After inspecting the whole row,
remove it:

```{r}
gpfg <- gpfg |>
  filter(!is.na(company))
```

The `!` means “not,” so the code keeps rows whose company is not missing.

Never remove every row containing `NA` automatically. In another dataset, a
missing value may be expected or may require more reporting. Here the decision
is justified because the entire supposed holding is incomplete.

## Check duplicate rows

The file should contain one row per reported holding. Count combinations of
company, country, and industry to find possible repetition:

```{r}
gpfg |>
  count(company, country, industry) |>
  filter(n > 1)
```

`count()` creates `n`, the number of matching rows. Inspect the result in the
original table:

```{r}
gpfg |>
  filter(company == "3P Learning Ltd")
```

The two complete rows are identical. After confirming that, use `distinct()`
without column names to keep each complete row once:

```{r}
gpfg <- gpfg |>
  distinct()
```

A repeated company name is not always a duplicate. The decision depends on
the meaning and values of the complete row.

## Verify and save the clean table

Finish by checking the structure, dimensions, and a few rows:

```{r}
glimpse(gpfg)

gpfg |>
  slice_head(n = 5)
```

The clean table has 7,201 rows and 11 columns. Its names are convenient, the
reporting date is a date, the incomplete row is gone, and the confirmed
duplicate has been removed.

Keep `gpfg_messy.csv` unchanged as the imported source. Save the clean object
under a new filename:

```{r}
write_csv(gpfg, "data/gpfg_2025.csv")
```

Chapter 5 will combine this clean file with four earlier annual files. If you
begin the book later, an identical ready-made
[`gpfg_2025.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2025.csv)
is available from the repository.

## Practice

Restart R and knit `04-data-cleaning.Rmd` from the beginning. Then answer:

1. Which observation justified each cleaning step?
2. What did `rename()` change, and what did it leave unchanged?
3. Why was `dmy()` the correct date function?
4. What evidence justified removing the incomplete row?
5. What evidence justified removing the duplicate row?
6. Why was the original CSV kept under a different name?

## Takeaways

| Function or idea | What it does |
|---|---|
| `rename(new = old)` | Gives selected columns clearer names |
| `mutate()` | Creates or changes columns |
| `dmy()` | Converts day-month-year text to a date |
| `is.na()` | Identifies missing values |
| `filter()` | Keeps rows that meet a condition |
| `!` | Means “not” inside a logical condition |
| `count()` | Counts repeated combinations |
| `distinct()` | Keeps each complete row once |
| `write_csv()` | Saves a tibble as a new CSV file |

::: {.callout-tip title="Want more control?"}
Real reporting data may contain many date formats, number symbols, or text
inconsistencies. Begin with the four problems in this lesson. The official
[dplyr reference](https://dplyr.tidyverse.org/reference/index.html) and
[lubridate reference](https://lubridate.tidyverse.org/reference/index.html)
provide additional tools when another dataset genuinely needs them.
:::
