---
title: "Data Cleaning"
---

## Learning objectives

By the end of this chapter, you should be able to:

- explain why data should be checked before analysis;
- decide whether columns have suitable data types;
- distinguish parsing text from converting an existing R value;
- convert text to integers, numbers, and dates;
- convert several common date formats into R dates;
- inspect missing, unexpected, and repeated values;
- make targeted corrections with `filter()`, `rename()`, and `mutate()`; and
- keep the downloaded source separate from a prepared table.

## From importing to cleaning

The previous chapter brought a CSV file into R. Before using it to answer a
reporting question, we should check whether its structure and values make
sense. This process is often called **data cleaning**.

Cleaning does not mean changing every dataset. It means looking for possible
problems, deciding whether they really are problems, making necessary changes,
and checking the result.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `03-data-cleaning.Rmd`. Reuse `data/gpfg.csv`;
there is no need to download or copy it again.
:::

Start by loading the tidyverse and importing the same file:

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg.csv")
```

One row represents one reported equity holding at the end of 2025. Keep that
row meaning in mind throughout the cleaning process.

## A simple cleaning workflow

Use the same basic approach whenever you receive a new table:

1. Understand what one row and each column represent.
2. Inspect the column names, data types, and example values.
3. Identify a specific possible problem.
4. Correct it in a new R object if necessary.
5. Inspect the result and record what you changed.

Do not remove or change a value merely because it looks unusual. It may be a
real and newsworthy observation. Check the source and column definition first.

## Check the column types

Begin with `glimpse()`:

```{r}
glimpse(gpfg)
```

The abbreviations beside the column names show how R has stored each column:

| Type | Meaning | Example in `gpfg` |
|---|---|---|
| `<chr>` | Character, or text | `company`, `country` |
| `<dbl>` | A number that can contain decimals | `market_value_usd`, `ownership_pct` |
| `<date>` | A calendar date | `report_date` |

A type is suitable when it matches how we need to use the variable. Company
names should be character values, market values should be numeric, and report
dates should be dates. The `year` column appears as `<dbl>`, which is still a
valid numeric type for a year. It is not automatically a problem.

## Correct a column type

Sometimes numbers and dates are imported as character values. The following
small table illustrates the problem. Every column starts as text because its
values are inside quotation marks:

```{r}
type_example <- tibble(
  year = c("2024", "2025"),
  report_date = c("2024-12-31", "2025-12-31"),
  value_usd = c("1200000", "1350000")
)

glimpse(type_example)
```

`mutate()` creates or changes columns. Inside `mutate()`, readr's parsing
functions convert the text to the types we want:

```{r}
type_example_clean <- type_example |>
  mutate(
    year = parse_integer(year),
    report_date = parse_date(report_date),
    value_usd = parse_double(value_usd)
  ) |>
  rename(market_value_usd = value_usd)

glimpse(type_example_clean)
```

The tidyverse loads readr, so we do not need a separate `library(readr)` line.

| Function | Use it when text should become... |
|---|---|
| `parse_integer()` | A whole number |
| `parse_double()` | A number that may contain decimals |
| `parse_date()` | A date written as `YYYY-MM-DD` |

`rename()` changes a column name using `new_name = old_name`. Here the more
specific name `market_value_usd` makes the unit clear.

::: {.callout-important title="Do not convert identifiers into numbers"}
A variable containing digits is not always a number. Telephone numbers, ID
numbers, and postal codes are labels, so they are usually better stored as
character values. Converting them to numbers can remove meaningful leading
zeros.
:::

## Parsing and converting are different

The function names can be confusing because readr and base R provide two
related families:

| Task | Tidyverse/readr | Base R |
|---|---|---|
| Turn text into whole numbers | `parse_integer()` | `as.integer()` |
| Turn text or values into decimal numbers | `parse_double()` | `as.double()` |
| Turn text into a date | `parse_date()` | `as.Date()` |

The `parse_*()` functions are designed for character values read from files.
They expect a particular kind of text and report values they cannot interpret.
For example, `parse_integer("2.5")` fails because `2.5` is not a whole number.

The `as.*()` functions **coerce**, or convert, an existing R value into another
type. They are useful when the value is already clean and its meaning is known.
For example, `as.integer(2.5)` returns `2`, which removes the decimal part. That
may be intended, but it could also silently change the meaning of the value.

::: {.callout-note title="Exact spelling matters"}
The base R names are `as.integer()`, `as.double()`, and `as.Date()`—with periods
and a capital `D` in `Date`. There is no regular `as_integer()` or `as_double()`
function loaded by the tidyverse.

Lubridate does provide `as_date()`. It is most useful for converting an
existing date-time value into a date; it is not our first choice for reading a
column containing several messy character formats.
:::

For imported character columns, this book begins with the tidyverse parsing
functions because their warnings help us find values that need investigation.

::: {.callout-tip title="Which family does this book use?"}
Both families are valid. For imported character data, the main lessons use
readr's `parse_*()` functions because they fit the tidyverse workflow and make
failed conversions easier to inspect. Base R functions such as `as.integer()`,
`as.double()`, and `as.Date()` are useful and may appear in code elsewhere, but
you do not need to learn both versions at the same time. Begin with the
tidyverse examples in this book and consult the alternatives when a project
requires them.
:::

## Work with real-world dates

A date column should be stored as an R `<date>`, not merely as character text.
R can then sort dates chronologically, calculate intervals, group observations
by time, and draw a correctly ordered time series. R normally displays a Date
in the standard `YYYY-MM-DD` form.

### Identify the order first

Before converting dates, determine the order of the components:

| Function | Expected order | Example |
|---|---|---|
| `ymd()` | year, month, day | `2025-03-31` |
| `dmy()` | day, month, year | `31/03/2025` |
| `mdy()` | month, day, year | `03/31/2025` |

These functions come from lubridate, which is loaded by
`library(tidyverse)`. The letters in the function name describe the order.

Dates can use different separators while keeping the same order. `ymd()` can
handle all three of these values:

```{r}
year_first_dates <- tibble(
  date_text = c("2025-01-31", "2025/02/01", "2025.03.02")
) |>
  mutate(report_date = ymd(date_text))

year_first_dates
```

Use `dmy()` when the day comes first:

```{r}
dmy(c("31/01/2025", "1-2-2025", "3 March 2025"))
```

Use `mdy()` only when the source follows month-day-year order.

### Handle several orders in one column

Some real-world files mix date orders. In that case,
`parse_date_time()` can try a short list of known orders. It first returns a
date-time, and `as_date()` then keeps its calendar date:

```{r}
mixed_dates <- tibble(
  date_text = c("2025-01-31", "13/02/2025", "March 3, 2025")
) |>
  mutate(
    report_date = date_text |>
      parse_date_time(orders = c("ymd", "dmy", "mdy")) |>
      as_date()
  )

mixed_dates
```

Only list orders that the source could genuinely use. Trying many possible
orders can hide errors rather than clean them.

::: {.callout-important title="R cannot resolve an ambiguous date for you"}
The value `03/04/2025` could mean 3 April or March 4. Examine the source's
documentation, location, and nearby unambiguous dates before choosing `dmy()`
or `mdy()`. Never guess when the distinction matters to the story.
:::

### Check the result

Always keep the original text beside the converted date while checking. Use
`glimpse()` to confirm that the new column is `<date>`, then find values that
failed to convert:

```{r}
glimpse(mixed_dates)

mixed_dates |>
  filter(is.na(report_date))
```

An empty result means every value in this example became a date. In a real
file, investigate every returned row. A failed conversion may reveal a typo,
an incomplete date, a different language, or another format that needs an
explicit rule.

::: {.callout-note title="Not every time-related value is a date"}
A year such as `2025` can remain an integer. A label such as academic year
`2024/25` represents a period rather than one calendar day, so it should not
automatically be converted to a Date.
:::

## Check missing values

R represents a missing value as `NA`. A missing value is different from zero:
zero is a known amount, while `NA` means the value is unavailable.

`is.na()` asks whether a value is missing. We can use it inside `filter()` to
keep rows where a particular column contains `NA`:

```{r}
gpfg |>
  filter(is.na(company))
```

```{r}
gpfg |>
  filter(is.na(market_value_usd))
```

Both checks return no rows, so these two columns do not contain missing values
in this file. If rows appeared, we would investigate why the values were
missing before deciding whether to keep, correct, or remove them.

## Check unexpected values

For a category such as region or industry, `distinct()` shows each different
value once:

```{r}
gpfg |>
  distinct(region)
```

```{r}
gpfg |>
  distinct(industry)
```

Look for spelling differences, unexpected spaces, or categories that do not
match the data dictionary. Two labels that differ by only one character will
be treated as different groups later.

We can also use `filter()` to inspect numeric values that may be impossible or
need explanation. For example:

```{r}
gpfg |>
  filter(market_value_usd < 0)
```

This check returns no rows. In another dataset, a negative value might be an
error, a special accounting value, or a legitimate observation. The code finds
the rows; reporting and source checking determine what they mean.

## Check repeated rows

A repeated row is not automatically a duplicate. First decide which columns
should identify one observation. In this file, we can begin by checking the
combination of company, country, and industry:

```{r}
gpfg |>
  count(company, country, industry) |>
  filter(n > 1)
```

`count()` creates a column called `n` containing the number of matching rows.
`filter(n > 1)` keeps combinations that appear more than once. This check
returns no rows in the current file.

If it found repeated combinations, we would compare them with the original
source before deleting anything. A company can sometimes legitimately have
more than one reported security or observation.

## Keep the source intact

The checks show that `gpfg.csv` already has sensible types and no missing or
repeated values in the columns examined. That is a valid cleaning conclusion:
not every imported file needs correction.

The downloaded CSV should remain unchanged. When corrections are necessary,
make them in a new R object and keep the code that produced it. Keeping the
source and cleaned data separate makes the work easier to audit and reproduce.

## Practice

Copy this deliberately messy table into your R Markdown document:

```{r}
practice_holdings <- tibble(
  company = c("Company A", "Company B", "Company B"),
  year = c("2024", "2025", "2025"),
  report_date = c("2024-12-31", "31/12/2025", "December 31, 2025"),
  value_usd = c("1200000", "not available", "not available")
)
```

Then:

1. Use `glimpse()` to identify the current column types.
2. Rename `value_usd` as `market_value_usd`.
3. Convert `year` and `market_value_usd` with suitable readr functions.
4. Convert the mixed values in `report_date` into one Date column.
5. Use `filter()` and `is.na()` to find the value that could not become a
   number.
6. Use `count()` and `filter(n > 1)` to identify the repeated observation.
7. Write two sentences explaining which values need investigation and why you
   should not automatically delete them.

## Takeaways

| Function or idea | What it does |
|---|---|
| `glimpse()` | Shows column names, types, and example values |
| `mutate()` | Creates or changes columns |
| `parse_integer()` | Converts suitable text to whole numbers |
| `parse_double()` | Converts suitable text to numeric values |
| `parse_date()` | Converts suitable text to dates |
| `as.integer()`, `as.double()`, `as.Date()` | Convert existing R values using base R |
| `ymd()`, `dmy()`, `mdy()` | Parse dates when the component order is known |
| `parse_date_time()` | Tries specified orders in a mixed date or date-time column |
| `as_date()` | Converts a date-time into a calendar date |
| `rename(new = old)` | Changes a column name |
| `filter()` | Keeps rows that meet a condition |
| `is.na()` | Tests whether a value is missing |
| `distinct()` | Shows unique rows or combinations |
| `count()` | Counts matching rows |

::: {.callout-tip title="Want more control?"}
The official [readr column-type guide](https://readr.tidyverse.org/articles/column-types.html)
explains additional parsing functions. The official
[lubridate date-parsing reference](https://lubridate.tidyverse.org/reference/ymd.html)
and [mixed-format reference](https://lubridate.tidyverse.org/reference/parse_date_time.html)
cover more date formats. The [dplyr reference](https://dplyr.tidyverse.org/reference/index.html)
covers more tools for selecting, filtering, and transforming data. Add them
only when a dataset presents a problem that requires them.
:::
