---
title: "Find, Audit, and Import Data"
---

## Learning objectives

By the end of this chapter, you should be able to:

- evaluate a source before opening the data;
- distinguish raw, processed, and analysis-ready files;
- import CSV and Excel files with explicit paths;
- inspect dimensions, names, types, and parsing problems; and
- write a one-sentence statement of the unit of analysis.

## Newsroom question

> What did Norway's sovereign wealth fund report holding at the latest year-end,
> and what can this file legitimately tell us?

The analysis begins before R. Open the
[NBIM all-investments page](https://www.nbim.no/en/investments/all-investments/)
and record:

- publisher and institutional responsibility;
- reporting date;
- asset type;
- available file formats;
- definitions or terms of use;
- update frequency; and
- the URL and date you accessed it.

Authority does not remove the need for scrutiny. Official data can contain
classification changes, revisions, omissions, and definitions that differ from
ordinary language.

## Raw, processed, and analysis-ready

This project uses three layers:

```{mermaid}
flowchart LR
  A["Official annual Excel files"] --> B["Reproducible preparation script"]
  B --> C["Clean UTF-8 CSV and RDS files"]
  C --> D["Chapter-specific analysis"]
```

- **Raw files** are unchanged evidence from the publisher.
- **Processed files** standardize names, types, dates, and provenance.
- **Analysis objects** are filtered or aggregated for a specific question.

Never overwrite the raw evidence with a cleaned file.

For this chapter, download or use
[the cleaned latest-year snapshot](https://raw.githubusercontent.com/binchen19/djr/main/data/processed/gpfg_equities_latest.csv).
The canonical teaching files are kept once in the repository rather than
duplicated inside the rendered website. The cleaning process is documented in
`scripts/prepare_gpfg.R` and the
[data audit](reports/gpfg-data-audit.md).

## Import a CSV

```{r}
library(tidyverse)

gpfg_latest <- read_csv(
  "data/processed/gpfg_equities_latest.csv",
  show_col_types = FALSE
)

latest_year <- unique(gpfg_latest$year)
stopifnot(length(latest_year) == 1)
```

Use a project-relative path. The code should work on another student's
computer after they open the same project.

Immediately inspect the result:

```{r}
dim(gpfg_latest)
names(gpfg_latest)
glimpse(gpfg_latest)
problems(gpfg_latest)
```

`dim()` reports the number of rows and columns. `problems()` should return an empty
table. An empty parsing-problem table does not prove the values are correct; it
only means `read_csv()` did not detect a conversion failure.

## State the grain

Before calculating anything, finish this sentence:

> One row represents one equity holding reported by NBIM at the latest year-end.

Most rows name companies, but some names may be funds or other listed issuers.
The source has no transaction identifier. A row does **not** represent a
purchase, a sale, a return, or an amount paid to the named company.

Check whether the apparent identifier is unique:

```{r}
gpfg_latest |>
  count(company) |>
  filter(n > 1)
```

The result should have no rows, showing that published company names are unique within
this snapshot. That does not make a company name a stable identifier across
years.

## Read the variables as evidence

```{r}
gpfg_dictionary <- read_csv(
  "data/processed/gpfg_data_dictionary.csv",
  show_col_types = FALSE
)

gpfg_dictionary |>
  select(variable, description, unit_or_example)
```

Pay particular attention to units:

- `market_value_nok` is the year-end holding value in Norwegian kroner;
- `market_value_usd` is the same snapshot converted to US dollars;
- `ownership_pct` and `voting_pct` are stored in **percentage points**, so
  `0.65` means `0.65%`; and
- `country` is the investment market, while `incorporation_country` is the
  issuer's legal incorporation country.

## Import the official Excel file

The raw official download is an Excel workbook. The following reads the latest
archived source directly:

```{r}
library(readxl)

latest_raw_file <- unique(gpfg_latest$source_file)
stopifnot(length(latest_raw_file) == 1)

gpfg_raw_latest <- read_excel(
  file.path("data/raw/gpfg", latest_raw_file),
  sheet = "Holdings Report"
)

names(gpfg_raw_latest)
```

The cleaned CSV uses easier `snake_case` names and adds the year, report date,
source file, and source URL. Those additions are provenance, not new reporting
from NBIM.

::: {.callout-note title="Why the preparation script uses Excel"}
NBIM's CSV and Excel exports have the same visible source columns, but the CSV
export leaves incorporation country blank. The Excel files retain it and also
store money and percentage fields as numeric values. The original CSV exports
are still archived for comparison.
:::

## A compact source audit

Create a data diary entry before analysis:

| Question | Answer for this file |
|---|---|
| Publisher | Norges Bank Investment Management |
| Asset type | Listed equities |
| Reporting date | Latest available year-end |
| Grain | One reported equity holding at year-end |
| Main measures | Market value, ownership, voting rights |
| Key limitation | Snapshot values do not reveal transactions or returns |
| Local raw file | The `source_file` recorded in the processed snapshot |
| Cleaning record | `scripts/prepare_gpfg.R` |

## Validation checkpoint

```{r}
stopifnot(nrow(gpfg_latest) > 0)
stopifnot(!anyNA(gpfg_latest$company))
stopifnot(!anyNA(gpfg_latest$market_value_nok))
stopifnot(all(gpfg_latest$market_value_nok >= 0))
stopifnot(all(gpfg_latest$year == latest_year))
```

These checks encode justified expectations for the latest-year teaching file. They
should not be copied blindly to the full archive, which contains legitimate
historical negative records.

## Independent task

Choose one row and trace it from the processed CSV back to the official Excel
file. Submit:

1. the company, country, and NOK market value;
2. the row's meaning in plain language;
3. one claim the row supports;
4. one tempting claim it does not support; and
5. a screenshot or code output showing the values agree across the two files.
