---
title: "Data Cleaning"
---

## Learning objectives

By the end of this chapter, you should be able to:

- explain what data cleaning is and what it is intended to achieve;
- turn observations from the import chapter into a short cleaning plan;
- rename awkward columns with `rename()`;
- change a date column to the correct type with `mutate()` and `dmy()`;
- inspect and handle missing values;
- identify and remove a confirmed duplicate row; and
- save a clean CSV without overwriting the imported source.

## What is data cleaning?

**Data cleaning** is the process of identifying, investigating, and correcting
problems that could make an analysis inaccurate, confusing, or difficult to
reproduce. Common problems include unclear column names, incorrect types,
missing information, inconsistent categories, and duplicated records.

The goal is not to make a dataset look perfect or to remove every unusual
value. The goal is to produce a table that is **fit for the reporting task**:

- each row and column has a clear meaning;
- values are stored in types appropriate for their meaning;
- missing and repeated records have been investigated;
- every change is supported by evidence and recorded in code; and
- the original source remains unchanged so the work can be checked and rerun.

What counts as “clean” depends on the question. A missing value that prevents
one analysis may be acceptable for another. Cleaning therefore requires
judgment and source knowledge, not just a list of functions.

::: {.callout-note title="Goal for this chapter"}
We will turn the deliberately imperfect 2025 file into a reproducible,
analysis-ready table while preserving the imported CSV. Every change will
respond to a problem observed in Chapter 3.
:::

## Continue with the imported table

Chapter 3 imported `gpfg_messy.csv` as `gpfg_raw`. Most of the table arrived
correctly: the amounts and percentages are already numeric, and the category
values are consistent. We only need to solve four visible problems:

1. a few column names are awkward;
2. the reporting date is stored as text;
3. one incomplete row does not describe a holding; and
4. one complete row has been repeated.

This smaller set of problems lets us practise a cleaning workflow without
changing every part of the data at once.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `04-data-cleaning.Rmd`. Reuse
`data/gpfg_messy.csv`; do not download another copy.
:::

Start the new document by recreating the object. Each R Markdown document must
contain the code needed to run on its own:

```{r}
library(tidyverse)

gpfg_raw <- read_csv("data/gpfg_messy.csv")
```

One row is intended to represent one reported equity holding at the end of
2025.

## Turn observations into a cleaning plan

Run the inspection from Chapter 3 again:

```{r}
glimpse(gpfg_raw)
```

Translate what we see into actions:

| Observation | Cleaning action |
|---|---|
| A few column names contain spaces and capitals | Rename those columns |
| `Report Date` is `<chr>` text | Convert it to a date |
| The file has a row without a company or value | Inspect the missing row |
| The file has one more complete row than expected | Check for a duplicate |

Choose a function only after identifying the problem it needs to solve.

## Rename the awkward columns

Use `rename()` with `new_name = old_name`. An old name containing spaces needs
backticks:

```{r}
gpfg <- gpfg_raw |>
  rename(
    year = Year,
    report_date = `Report Date`,
    company = `Company Name`,
    market_value_nok = `Market Value NOK`
  )

glimpse(gpfg)
```

The new names use lowercase `snake_case`, the naming style introduced in
Chapter 2. The other column names were already suitable, so we did not rename
them. Renaming changes labels, not values.

`year` appears as `<dbl>`, which is still a numeric type and is suitable for
these whole-year values. We do not need to convert a column merely to make its
type label look different.

## Convert the date type

The values in `report_date` use day-month-year order, such as `31/12/2025`,
but R imported the column as `<chr>` text. The data source tells us what the
order means.

Use `mutate()` to change the existing column. `dmy()` means day, month, year:

```{r}
gpfg <- gpfg |>
  mutate(report_date = dmy(report_date))

glimpse(gpfg)
```

The type should now appear as `<date>`. `mutate()` can create a new column or
change an existing one; here it replaces the text version with a date version.

::: {.callout-important title="Confirm the date order"}
Do not guess whether `03/04/2025` means 3 April or March 4. Check the source
documentation. Use `dmy()` for day-month-year, `mdy()` for month-day-year, or
`ymd()` for year-month-day.
:::

## Check missing values

R represents unavailable information with `NA`. A missing value differs from
zero: zero is a known amount, while `NA` means the value is unavailable.

### How `filter()` works

`filter()` is a dplyr function that keeps rows where a condition is `TRUE`:

```r
table |>
  filter(condition)
```

It selects rows; it does not change column names or values. If the result is
only printed, the original object remains unchanged. If the pipeline is
assigned back to an object with `<-`, that object is updated to contain only
the kept rows.

The conditions used in this chapter are:

| Condition | Meaning |
|---|---|
| `is.na(company)` | The company value is missing |
| `!is.na(company)` | The company value is not missing |
| `n > 1` | The count is greater than one |
| `company == "3P Learning Ltd"` | The company equals this exact text |

`!` means “not,” `>` means “greater than,” and `==` tests whether two values
are equal. A single `=` has a different role: it names arguments and options
inside functions.

First use `is.na()` inside `filter()` to display rows with a missing company
name. Because this result is not assigned, it does not change `gpfg`:

```{r}
gpfg |>
  filter(is.na(company))
```

The result contains one deliberately incomplete row. It has a year and date
but no company, country, industry, or market value, so it cannot represent the
holding described by the data dictionary. After inspecting the whole row,
remove it:

```{r}
gpfg <- gpfg |>
  filter(!is.na(company))
```

This time the result is assigned back to `gpfg`, so the working object now
keeps only rows whose company is not missing.

Never remove every row containing `NA` automatically. In another dataset, a
missing value may be expected or may require more reporting. Here the decision
is justified because the entire supposed holding is incomplete.

## Check duplicate rows

The file should contain one row per reported holding. Count combinations of
company, country, and industry to find possible repetition:

```{r}
gpfg |>
  count(company, country, industry) |>
  filter(n > 1)
```

`count()` creates `n`, the number of matching rows. Inspect the result in the
original table:

```{r}
gpfg |>
  filter(company == "3P Learning Ltd")
```

The two complete rows are identical. After confirming that, use `distinct()`
without column names to keep each complete row once:

```{r}
gpfg <- gpfg |>
  distinct()
```

A repeated company name is not always a duplicate. The decision depends on
the meaning and values of the complete row.

## Verify and save the clean table

Finish by checking the structure, dimensions, and a few rows:

```{r}
glimpse(gpfg)

gpfg |>
  slice_head(n = 5)
```

The clean table has 7,201 rows and 11 columns. Its names are convenient, the
reporting date is a date, the incomplete row is gone, and the confirmed
duplicate has been removed.

Keep `gpfg_messy.csv` unchanged as the imported source. Save the clean object
under a new filename:

```{r}
write_csv(gpfg, "data/gpfg_2025.csv")
```

Chapter 5 will analyse this clean annual file, and Chapter 6 will combine it
with four earlier annual files. If you
begin the book later, an identical ready-made
[`gpfg_2025.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_2025.csv)
is available from the repository.

## Practice

Restart R and knit `04-data-cleaning.Rmd` from the beginning. Then answer:

1. Which observation justified each cleaning step?
2. What did `rename()` change, and what did it leave unchanged?
3. Why was `dmy()` the correct date function?
4. What evidence justified removing the incomplete row?
5. What evidence justified removing the duplicate row?
6. Why was the original CSV kept under a different name?
7. What is the difference between printing a `filter()` result and assigning
   that result back to `gpfg`?

## Takeaways

| Function or idea | What it does |
|---|---|
| `rename(new = old)` | Gives selected columns clearer names |
| `mutate()` | Creates or changes columns |
| `dmy()` | Converts day-month-year text to a date |
| `is.na()` | Identifies missing values |
| `filter()` | Keeps rows where a condition is true; assignment saves the subset |
| `!` | Means “not” inside a logical condition |
| `count()` | Counts repeated combinations |
| `distinct()` | Keeps each complete row once |
| `write_csv()` | Saves a tibble as a new CSV file |

::: {.callout-tip title="Want more control?"}
Real reporting data may contain many date formats, number symbols, or text
inconsistencies. Begin with the four problems in this lesson. The official
[dplyr reference](https://dplyr.tidyverse.org/reference/index.html) and
[lubridate reference](https://lubridate.tidyverse.org/reference/index.html)
provide additional tools when another dataset genuinely needs them.
:::
