---
title: "Data Import"
format: html
---

::: {.callout-note}
## Part 2: Data Import & Wrangling

In this part, you'll learn how to get data into R and transform it into usable form. Data journalism starts here. While exciting stories hide in data, that data is often messy, scattered across multiple files, or structured in ways that aren't ideal for analysis.

**What you'll learn:**
- Import data from various sources (CSV, Excel, URLs)
- Clean and reshape data for analysis
- Join multiple datasets together
- Prepare publication-ready datasets

These are not exciting skills to discuss at parties, but they're essential. Studies suggest data professionals spend 70-80% of their time wrangling data before ever touching analysis. Master this foundation, and everything else becomes easier.
:::

## Importing Data in R

### Built-in Datasets from Packages

R comes with a variety of built-in datasets that can be loaded directly from packages, here are some popular ones:


| Package       | Key Datasets          | Load Command         |
|---------------|-----------------------|----------------------|
| `datasets`    | mtcars, iris          | Built-in             |
| `ggplot2`     | diamonds, mpg         | `library(tidyverse)` |
| `nycflights13`| flights, weather      | `install.packages()` |
| `gapminder`   | gapminder             | `library(gapminder)` |

**Usage Example:**
```{r}
#| eval: false
mtcars
```

### Downloading External Data

#### From TidyTuesday
```{r}
#| eval: false
#install.packages("tidytuesdayR")
library(tidytuesdayR)

# Load 2024 Olympics data
tuesdata <- tt_load('2024-08-06')  
olympics <- tuesdata$olympics
```

#### Direct from URL
```{r}
#| eval: true
#| message: false
#| warning: false

library(tidyverse)

# Hong Kong graduates salary data

data_url = "https://www.ugcs.gov.hk/datagovhk/Average_Annual_Salaries_FT_Employment(Eng).csv"
hksalary_download <- read_csv(data_url)

hksalary_download
```


### Local File Import

In data analysis projects, importing local files is more common than importing data from the web. Here are some common file types and their uses:

-  **CSV Files**: Simple, human-readable, and widely supported. Ideal for tabular data.
-  **Excel Files**: Used for spreadsheets with multiple sheets or formatting. Imported with `readxl` or `openxlsx`.
-  **SPSS, SAS, Stata Files**: Common in social science and survey research. Use specialized R packages to import.
-  **RDS Files**: Binary format for storing R objects, preserving their structure and class information.
-  **RData Files**: Binary format for saving multiple R objects in a single file, often used for workspaces.

::: {.callout-note}
For this course, we will focus on **CSV** files, as they are simple and widely used.
:::

### CSV Files with `readr` (`tidyverse`) package

First, we download the CSV file from the web and save it locally as `hksalary.csv`. Then, we import it using the `read_csv()` function from the `readr` package.

::: {.callout-important}
## `read_csv()` vs. `read.csv()`
Note that `read_csv()` from `readr` is preferred over `read.csv()` from base R for its speed and consistency. In this course, we recommend using `read_csv()` for CSV files.
:::

```{r}
#| eval: true
# Relative path (recommended)
hksalary <- read_csv("data/hksalary.csv")
hksalary
```

#### File Path Management

| Path Type     | Example                     | When to Use          |
|---------------|-----------------------------|----------------------|
| **Relative**  | `data/hksalary.csv`         | Default in projects  |
| **Absolute**  | `C:/Users/.../hksalary.csv` | Temporary analysis   |

::: {.callout-tip}
## Relative Path
Use relative paths for portability and to avoid hardcoding directory paths.
:::



## Data Inspection

After importing data, it's essential to inspect it to understand its structure and contents. Here are some common functions to help you get started:

### First Look Tools

- **`head()`**: Shows the first few rows of the dataset.

```{r}
# First 6 rows
head(hksalary)
```

-  **`glimpse()`**: Provides a concise summary of the dataset's structure.


```{r}
# # Tidyverse alternative to str()
glimpse(hksalary)  
```

-  **`summary()`**: Displays a statistical summary of the dataset.

```{r}
summary(hksalary)
```

| Function     | Output Focus             | Tidyverse Equivalent |
|--------------|--------------------------|----------------------|
| `head()`     | Top rows                 | `slice_head()`       |
| `str()`      | Data types & structure   | `glimpse()`          |


## Practice: Import `hksalary.csv` Data

### Step-by-Step Practice

1. **Setup Workspace**
   - Create `import-practice` project
   - Make `/data` subfolder

2. **Store Data**
   - Download [Hong Kong Graduates Annual Salary Data](https://www.ugcs.gov.hk/datagovhk/Average_Annual_Salaries_FT_Employment(Eng).csv)
   - Save as `hksalary.csv` in `/data`

3. **Import Data**
   ```{r}
   library(tidyverse)
   hksalary <- read_csv("data/hksalary.csv")
   ```

4. **Initial Inspection**
   ```{r}
   #| eval: false
   glimpse(hksalary)
   summary(hksalary)
   ```

## Key Functions Recap

| Task                | Function                 | Example                     |
|---------------------|--------------------------|-----------------------------|
| Load package        | `library()`              | `library(tidyverse)`        |
| Read CSV            | `read_csv()`             | `read_csv("data/file.csv")` |
| View structure      | `glimpse()`/`str()`      | `glimpse(df)`               |
| Show first rows     | `head()`                 | `head(df, n = 10)`          |
| Statistical summary | `summary()`              | `summary(df$salary)`        |
