---
title: "Wrangle Data Without Losing Its Meaning"
---

## Learning objectives

By the end of this chapter, you should be able to:

- use `select()`, `filter()`, `arrange()`, and `mutate()`;
- count rows and distinct entities correctly;
- group and summarise with an explicit new grain;
- inspect missing and implausible values; and
- validate that a transformation preserved the expected total.

## Newsroom question

> Which markets accounted for the largest share of NBIM's latest reported
> year-end equity holdings?

This sounds like a ranking question, but the denominator matters. We need both
country totals and the grand total before calculating a share.

## Set up and preserve a baseline

```{r}
library(tidyverse)

gpfg <- read_csv(
  "data/processed/gpfg_equities_latest.csv",
  show_col_types = FALSE
)

baseline_rows <- nrow(gpfg)
baseline_total_nok <- sum(gpfg$market_value_nok)
latest_year <- unique(gpfg$year)
```

Saving simple baseline values makes later checks easy.

## Select variables

`select()` changes columns, not rows:

```{r}
gpfg_core <- gpfg |>
  select(
    company,
    country,
    region,
    industry,
    market_value_nok,
    ownership_pct
  )

glimpse(gpfg_core)
```

Keep provenance columns in the master object even if a smaller analysis object
does not display them.

## Filter observations

`filter()` keeps rows that satisfy a logical condition:

```{r}
east_asia <- gpfg |>
  filter(country %in% c("China", "Hong Kong", "Japan", "South Korea"))

nrow(east_asia)
```

Use `%in%` when one variable can match several allowed values. Do not use
`country == "China" | "Japan"`; the second part is text, not a complete logical
test.

Multiple conditions can be written with commas, which means “and”:

```{r}
large_technology_holdings <- gpfg |>
  filter(
    industry == "Technology",
    market_value_usd >= 1e9
  )
```

## Arrange and inspect extremes

```{r}
gpfg |>
  arrange(desc(market_value_usd)) |>
  select(company, country, industry, market_value_usd) |>
  slice_head(n = 10)
```

A ranking is a lead, not a conclusion. Check whether the largest rows are
companies, funds, duplicate names, or units expressed differently from the
rest.

## Create useful variables

`mutate()` adds or changes columns:

```{r}
gpfg_scaled <- gpfg |>
  mutate(
    market_value_usd_billions = market_value_usd / 1e9,
    ownership_group = case_when(
      ownership_pct < 1 ~ "Below 1%",
      ownership_pct < 3 ~ "1% to below 3%",
      TRUE ~ "3% or more"
    )
  )

gpfg_scaled |>
  count(ownership_group)
```

The category boundaries are analytical choices. Record them in the methods
note and check whether a different threshold would change the story.

## Audit missingness before removing it

```{r}
gpfg |>
  summarise(
    across(
      everything(),
      ~ sum(is.na(.x)),
      .names = "missing_{.col}"
    )
  )
```

The processed latest-year file is complete. In other datasets, inspect missingness by
year, institution, or category before calling `drop_na()`. Systematic missing
data can be the most important finding.

## Count rows and entities separately

In this latest-year snapshot, company names are unique, but write the question
explicitly:

```{r}
gpfg |>
  summarise(
    holding_records = n(),
    distinct_company_names = n_distinct(company)
  )
```

`n()` counts records. `n_distinct(company)` counts distinct published names.
They happen to agree here; that agreement should be checked, not assumed.

## Group and summarise

The input grain is one holding. After summarising by country, the output grain
is one country.

```{r}
country_totals <- gpfg |>
  summarise(
    market_value_usd = sum(market_value_usd),
    holding_records = n(),
    companies = n_distinct(company),
    .by = country
  )

country_totals |>
  arrange(desc(market_value_usd)) |>
  slice_head(n = 10)
```

Write the grain change in your notebook:

| Object | One row represents |
|---|---|
| `gpfg` | one latest-year equity holding |
| `country_totals` | one investment market |

## Calculate a denominator explicitly

```{r}
grand_total_usd <- sum(country_totals$market_value_usd)

country_shares <- country_totals |>
  mutate(share_of_total = market_value_usd / grand_total_usd) |>
  arrange(desc(share_of_total))

country_shares |>
  select(country, market_value_usd, share_of_total) |>
  slice_head(n = 10)
```

The result supports a sentence such as: “The United States accounted for X%
of the reported market value in the latest equity-holdings file.” It does not
show the share of purchases made during that year.

## Validate the transformation

```{r}
stopifnot(nrow(gpfg) == baseline_rows)
stopifnot(sum(country_totals$holding_records) == baseline_rows)
stopifnot(
  isTRUE(all.equal(
    sum(country_totals$market_value_usd),
    sum(gpfg$market_value_usd)
  ))
)
stopifnot(isTRUE(all.equal(sum(country_shares$share_of_total), 1)))
```

These checks protect against accidental filtering, double-counting, and a
wrong denominator.

## Checkpoint exercise

Create an industry summary containing:

- total NOK market value;
- number of holding records;
- number of distinct company names;
- median holding value; and
- share of the overall NOK total.

Then answer:

1. What does one row in your result represent?
2. Why might the mean holding value be less useful than the median?
3. Which check proves the industry totals reconcile with the source?

<details>
<summary>Show one possible pipeline</summary>

```{r}
industry_summary <- gpfg |>
  summarise(
    market_value_nok = sum(market_value_nok),
    holding_records = n(),
    companies = n_distinct(company),
    median_holding_nok = median(market_value_nok),
    .by = industry
  ) |>
  mutate(
    share_of_total = market_value_nok / sum(market_value_nok)
  ) |>
  arrange(desc(market_value_nok))

stopifnot(isTRUE(all.equal(sum(industry_summary$share_of_total), 1)))
```

</details>

## Independent task

Investigate one region. Produce:

1. a clearly named filtered object;
2. a ranked table of its five largest holdings;
3. a country-level summary with shares within the region;
4. at least two validation checks; and
5. a 100-word methods note stating the grain, denominator, and one claim limit.
