---
title: "Data Import"
---

::: {.callout-note title="Part 2: Data Import & Cleaning"}
The previous chapters used values that we typed ourselves. We will now bring a
real table into R, inspect what arrived, and prepare it for analysis. Chapters
3 and 4 follow the same file from import to a clean dataset.
:::

## Learning objectives

By the end of this chapter, you should be able to:

- explain what a CSV file is;
- explain the relationship between the tidyverse, readr, and `read_csv()`;
- import a local CSV with a relative path;
- read the main parts of `glimpse()` output; and
- identify clues that a newly imported table needs cleaning.

## From a tibble to an external file

In Chapter 2, we used vectors and `tibble()` to create a small table ourselves.
Journalists usually receive a much larger table from a government agency,
company, researcher, or other source. **Importing** brings that external file
into R as an object so we can inspect and analyse it.

We will use a **CSV**, or comma-separated values file. A CSV stores a table as
plain text. It is widely used, easy to share, and can be opened by many
programs without depending on one spreadsheet application.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `03-data-import.Rmd`. Download
[`gpfg_messy.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_messy.csv)
and save it inside the project's `data` folder. Do not open and resave it in
spreadsheet software before importing it.
:::

The file is adapted from the 2025 equity-holdings data published by
[Norges Bank Investment Management](https://www.nbim.no/en/investments/all-investments/).
It contains the published values, with a few deliberately introduced problems
for teaching: awkward names, a date stored as text, one incomplete row, and
one duplicate. The amounts and percentages remain ordinary numeric values.

## Tidyverse, readr, and `read_csv()`

The **tidyverse** is a collection of packages designed to work together. One
of those packages is **readr**, which imports rectangular data files. The
readr function for CSV files is `read_csv()`.

Load the tidyverse at the beginning of the R Markdown document:

```{r}
library(tidyverse)
```

Loading the tidyverse also loads readr, so we do not need a second
`library(readr)` line. The simplest use of the function is:

```r
read_csv("file_path.csv")
```

The function needs the path to the file. A path is text, so it appears inside
quotation marks.

::: {.callout-note title="Why this book uses `read_csv()`"}
Base R also has a valid function named `read.csv()`. This book uses readr's
`read_csv()` because it returns the tibble used throughout the tidyverse and
reports the column types it detected. Learning one consistent import workflow
is enough for now.
:::

## Use a relative path

The learner project should now look like this:

```text
djr/
├── data/
│   └── gpfg_messy.csv
├── 01-setup.Rmd
├── 02-r-basics.Rmd
├── 03-data-import.Rmd
└── djr.Rproj
```

Starting from the project folder, the path to the file is:

```text
data/gpfg_messy.csv
```

This is a **relative path**: it describes where the file is located inside the
project. Everyone who opens `djr.Rproj` starts from the same project folder, so
the same path can work on different computers.

::: {.callout-tip title="Avoid personal paths"}
A path such as `/Users/name/Desktop/...` or `C:/Users/name/Desktop/...` points
to one person's computer. Keep course files inside the project and use short
relative paths.
:::

## Import the CSV

Import the table and assign it to an object named `gpfg_raw`:

```{r}
gpfg_raw <- read_csv("data/gpfg_messy.csv")
```

Read the line from right to left: `read_csv()` reads the file, and `<-` saves
the resulting tibble as `gpfg_raw`. The suffix `_raw` reminds us that this
object still reflects the imported file; we have not cleaned it.

readr may print a message describing the rows, columns, and column types. That
message is information about the import, not an error.

## Inspect what arrived

Importing without an error does not mean the table is ready for analysis.
Begin by looking at its structure and a few rows.

### Look at the structure

`glimpse()` shows the number of rows and columns, each column name and type,
and a few example values:

```{r}
glimpse(gpfg_raw)
```

Read the output from top to bottom:

- **`Rows: 7,203`** is the number of imported rows;
- **`Columns: 11`** is the number of variables;
- each line beginning with **`$`** describes one column;
- `<chr>` means character text and `<dbl>` means a number; and
- the values at the end of each line are examples, not the entire column.

Some observations should make us pause:

- names such as `Company Name` and `Market Value NOK` contain spaces and
  punctuation, making them awkward to type in code;
- `Report Date` is character text rather than an R date;
- the market-value and percentage columns correctly appear as numbers; and
- missing and duplicate rows will require checks that a quick preview cannot
  settle by itself.

These are not import failures. They are clues about work needed in Chapter 4.

### Preview a few rows

Use the pipe from Chapter 2 to pass the imported table to `slice_head()`:

```{r}
gpfg_raw |>
  slice_head(n = 5)
```

`slice_head(n = 5)` returns the first five rows without changing `gpfg_raw`.
The named argument `n = 5` tells the function how many rows to show.

::: {.callout-note title="Why not use `head()`?"}
`head(gpfg_raw)` is valid base R. This book uses `slice_head()` because it is a
dplyr function and fits the tidyverse pipelines used in later chapters.
:::

## Make an inspection checklist

After importing any dataset, ask:

| Check | Question to ask |
|---|---|
| Rows and columns | Does the size match what the source describes? |
| Column names | Are the names clear and convenient to use? |
| Column types | Are amounts numeric and dates stored as dates? |
| Missing values | Does `NA` mean unavailable, and how many are present? |
| Category values | Are spelling, capitalization, and spaces consistent? |
| Repeated records | Is each apparent duplicate an error or a valid observation? |
| Units and definitions | What exactly does every number measure? |

Chapter 4 will turn a small part of this checklist into code. We will rename a
few columns, convert the date, inspect a missing row, identify a repeated row,
and save the clean result for later chapters.

## Keep the local source copy

`read_csv()` can also read a public URL, and some packages provide example
datasets. Those approaches are useful to recognize, but the guided lessons use
a saved local file. A local copy:

- preserves the exact version used in the reporting;
- continues to work if a URL changes or the internet is unavailable; and
- avoids downloading the same file every time the document knits.

Record the publisher, original URL, and access date in your notes. Keeping a
local file does not replace source documentation.

## Refer to the data dictionary

Open the
[GPFG data dictionary](https://github.com/binchen19/djr/blob/main/data/gpfg_data_dictionary.csv)
and keep it beside the lesson. It explains what each column means and which
unit it uses. You do not need to import the dictionary.

One row should represent one reported equity holding at the end of the year.
It is not a purchase, sale, investment return, or payment received by a
company. We will check that intended row meaning during cleaning.

## Practice

In `03-data-import.Rmd`:

1. load the tidyverse;
2. import `data/gpfg_messy.csv` as `gpfg_raw`;
3. inspect the table with `glimpse()`;
4. show its first five rows; and
5. compare the imported types with the types you think the columns need.

Then answer:

1. How many rows and columns were imported?
2. Which columns did R correctly import as numbers?
3. Which column still needs to become a date?
4. Name one awkward column name.
5. Why does a successful import not prove that the table is clean?
6. What should one row represent?

## Takeaways

| Function or idea | What it does |
|---|---|
| `library(tidyverse)` | Loads the core tidyverse packages |
| `read_csv("path")` | Imports a CSV as a tibble |
| Relative path | Locates a file from the RStudio Project folder |
| `<-` | Saves the imported tibble as an object |
| `glimpse()` | Shows column names, types, and example values |
| `slice_head(n = 5)` | Shows the first five rows |
| Local source copy | Preserves the exact file used in the reporting |

::: {.callout-tip title="Want more control?"}
The file path is the only `read_csv()` argument needed in this lesson. The
official [`read_csv()` reference](https://readr.tidyverse.org/reference/read_delim.html)
explains optional controls for missing values, column types, delimiters, and
unusual files when a project requires them.
:::
