---
title: "Text Analysis"
---

::: {.callout-important title="The new challenge: count language without losing context"}
Words begin inside complete speaking turns, not in a ready-made numeric table.
To compare speakers, we must turn text into analysable rows and standardize the
counts. We must then return to the transcript because frequency alone cannot
explain what a speaker meant.
:::

## Learning objectives

By the end of this case, you should be able to:

- import a prepared transcript table;
- turn speaking turns into word rows;
- remove common stop words with a join;
- compare word use with a rate; and
- return to the transcript before interpreting a word count.

## Set up the case

The original transcript is a text file rather than a rectangular table. A
preparation script separates it into speaker turns and saves a teaching CSV.
This lets the lesson focus on tidy text analysis instead of loops and custom
parsing functions.

The reproducible preparation is available in `scripts/prepare_debate.R` for
readers who want to study it later.

If needed, install `tidytext` once in the Console with
`install.packages("tidytext")`. `tidytext` is a companion to the tidyverse: it
turns language into tidy rows that can be handled with familiar verbs such as
`filter()`, `count()`, and joins. Then load the packages:

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `12-text-analysis.Rmd`. Download
[`debate_turns.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/debate_turns.csv)
and save it in `data/`. The preparation script is optional; the lesson starts
from this ready-to-import CSV.
:::

```{r}
library(tidyverse)
library(tidytext)

turns <- read_csv("data/debate_turns.csv")

glimpse(turns)
```

One row represents one speaking turn.

## The strategy

We will solve the text challenge in six steps:

1. choose the speakers being compared;
2. turn speaking turns into word rows;
3. make a documented decision about common words;
4. keep the total number of words as a denominator;
5. compare word-use rates; and
6. return counted words to their original context.

## Strategy 1: choose the comparison

`filter()` keeps the candidate speakers, and `recode()` replaces their
all-capital labels with names that are easier to display:

```{r}
candidate_turns <- turns |>
  filter(speaker %in% c("HARRIS", "TRUMP")) |>
  mutate(
    speaker = recode(
      speaker,
      "HARRIS" = "Harris",
      "TRUMP" = "Trump"
    )
  )

candidate_turns |>
  count(speaker)
```

The row counts describe speaking turns, not speaking time or number of words.

## Strategy 2: turn text into word rows

`unnest_tokens()` from tidytext turns each speaking turn into separate word
rows. It needs a name for the new word column and the name of the original text
column:

```{r}
tokens <- candidate_turns |>
  unnest_tokens(word, text)

tokens |>
  slice_head(n = 10)
```

One row now represents one word token associated with a candidate.

## Strategy 3: decide how to handle common words

The tidytext package provides a table called `stop_words`, containing common
English words such as “the” and “of.” `anti_join()` keeps tokens that do not
match that reference table:

```{r}
content_words <- tokens |>
  anti_join(stop_words, by = "word")
```

Stop-word removal is an analytical choice. Words such as “not” may be important
in political speech, so inspect the list and describe the rule in the methods.

## Strategy 4: count words and keep a denominator

First count all tokens spoken by each candidate:

```{r}
token_totals <- tokens |>
  count(speaker) |>
  rename(all_tokens = n)
```

Then count each content word and join the denominator:

```{r}
word_rates <- content_words |>
  count(speaker, word) |>
  rename(uses = n) |>
  left_join(token_totals, by = "speaker") |>
  mutate(uses_per_1000_tokens = uses / all_tokens * 1000)
```

A rate is more comparable than a raw count when the candidates spoke different
numbers of words.

## Strategy 5: compare word-use rates

`slice_max()` keeps rows with the largest value. Group first so it selects the
largest rates separately for each candidate:

```{r}
top_words <- word_rates |>
  group_by(speaker) |>
  slice_max(uses_per_1000_tokens, n = 10)

top_words
```

Create a simple comparison chart:

```{r}
top_words |>
  ggplot(
    aes(
      x = uses_per_1000_tokens,
      y = fct_reorder(word, uses_per_1000_tokens),
      fill = speaker
    )
  ) +
  geom_col() +
  facet_wrap(~ speaker) +
  labs(
    title = "Frequently used content words differed between the candidates",
    x = "Uses per 1,000 candidate tokens",
    y = NULL,
    caption = "Source: supplied debate transcript"
  ) +
  theme_minimal()
```

Frequency does not measure importance, sincerity, truthfulness, or audience
effect.

## Optional extension: look at two-word phrases

Single words lose context. `unnest_tokens()` can also create adjacent two-word
phrases, called bigrams. Here `token = "ngrams"` selects word groups and `n = 2`
sets their length:

```{r}
bigrams <- candidate_turns |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  count(speaker, bigram) |>
  arrange(desc(n))

bigrams |>
  slice_head(n = 20)
```

Bigrams preserve a little more context but still do not capture full sentence
meaning, sarcasm, quotation, or who a pronoun refers to.

## Strategy 6: return to the transcript

Before using a word in a story, find it in the original speaking turns:

```{r}
candidate_turns |>
  filter(str_detect(str_to_lower(text), "immigration")) |>
  select(speaker, text)
```

Read several examples. Check whether the candidate is making a claim,
answering a moderator, quoting an opponent, or denying something.

## Practice

Choose one topic and make a short list of related words. Then:

1. count their uses by candidate;
2. calculate uses per 1,000 tokens;
3. read at least five matching speaking turns for each candidate;
4. identify one false match or missing expression; and
5. write a short note distinguishing the automated count from your
   interpretation.

## Takeaways

| Function | What it does in this case |
|---|---|
| `unnest_tokens()` | Turns speaking turns into word or phrase rows |
| `anti_join()` | Removes words found in the stop-word table |
| `count()` | Counts tokens by candidate and word |
| `left_join()` | Adds the token denominator to each word count |
| `slice_max()` | Keeps the most frequent words within each candidate |
| `str_detect()` | Returns counted words to their original context |

::: {.callout-tip title="Want more control?"}
This case uses only words and bigrams. The official
[`unnest_tokens()` reference](https://juliasilge.github.io/tidytext/reference/unnest_tokens.html)
explains other token types and controls for later text projects.
:::

## What the case studies taught us

The four cases began with the familiar tidyverse workflow but concentrated on
different problems that journalists encounter in real datasets:

| Dataset | Unique challenge | Main strategy |
|---|---|---|
| Billboard Hot 100 | Titles alone do not identify songs, and “success” has several meanings | Use artist-title combinations and define the metric before ranking |
| Olympics | Athlete-medal rows and unique medallists are not official medals | Identify the unit, use stable athlete identifiers, and create event-level records |
| Graduate employment | The table mixes detailed and aggregated rows | Split, aggregate, recombine, reshape, and construct the denominator |
| Debate transcript | Word counts remove language from its context | Tokenize, standardize rates, and return to the original transcript |

Across all four cases, the most important question comes before the code:

> What does one row represent, and does that unit match the claim we want to
> make?

Functions make the calculation reproducible. Defining the observation, metric,
and denominator makes the result meaningful.
