---
title: "Text Analysis"
subtitle: "2024 US Presidential Debate"
format: html
---

## Text Analysis with R: 2024 US Presidential Debate

## Load Packages

```{r}
library(tidyverse)
library(tidytext)
library(skimr)
```

## Cleaning Text Data to Data Frame

### Import

```{r}
# Read the transcript
transcript <- readLines("data/debate.txt")
transcript[1:10]
```

```{r}
# Remove empty lines
transcript <- transcript[transcript != ""]
```

### Parse transcripts to a data frame

```{r}
# Initialize an empty data frame
data <- data.frame(speaker = character(),
                   text = character(),
                   stringsAsFactors = FALSE)

# Define a regular expression pattern to detect speaker lines
speaker_pattern <- "^[A-Z ]+:"  # Lines that start with uppercase letters followed by a colon
# Initialize variables
current_speaker <- NA
current_text <- ""

for (line in transcript) {
  # Check if the line matches the speaker pattern
  if (str_detect(line, speaker_pattern)) {
    # If there's accumulated text, save it before moving to the next speaker
    if (!is.na(current_speaker) && current_text != "") {
      data <- rbind(data, data.frame(speaker = current_speaker, text = current_text, stringsAsFactors = FALSE))
      current_text <- ""
    }
    # Extract the speaker
    split_line <- str_split_fixed(line, ":", 2)
    current_speaker <- str_trim(split_line[1])
    # Start accumulating text
    current_text <- str_trim(split_line[2])
  } else {
    # Accumulate text
    current_text <- paste(current_text, str_trim(line))
  }
}

# Add the last piece of text
if (!is.na(current_speaker) && current_text != "") {
  data <- rbind(data, data.frame(speaker = current_speaker, text = current_text, stringsAsFactors = FALSE))
}
```

```{r}
# write_csv(data, "out/debate_df.csv")
```

## Import Cleaned Data

```{r}
df <- data
```

```{r}
skim(df)
```

```{r}
unique(df$Speaker)
```


## Data Cleaning

### Filter Out Non-Candidate Speakers

```{r}
df1 <- df |>
  filter(speaker == "HARRIS" | speaker == "TRUMP")

unique(df1$speaker)
```

### Standardize Speaker Names

```{r}
# Standardize speaker names if necessary
df1$speaker <- ifelse(df1$speaker == "HARRIS", "Harris", "Trump")
```

## Text Analysis

### Tokenize and Remove Stop Words

```{r}
# Tokenize the text column
tidy_data <- df1 %>%
  unnest_tokens(word, text)

head(tidy_data)
```

```{r}
# Load stop words
data("stop_words")

# Remove stop words
tidy_data_clean <- tidy_data %>%
  anti_join(stop_words, by = "word") %>%
  filter(!str_detect(word, "\\d+"))
```

```{r}
# Calculate word frequencies
word_counts <- tidy_data_clean %>%
  count(speaker, word, sort = TRUE)
```

### Analyze Top Words

```{r}
# Get top 10 words for each candidate
top_words <- word_counts %>%
  group_by(speaker) %>%
  top_n(10, n) %>%
  ungroup() %>%
  arrange(speaker, -n)

# View the result
print(top_words)
```

### Analyze Top Word Pairs (Bigrams)

```{r}
# Create bigrams
bigrams <- df1 %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2)
```

```{r}
# Separate bigrams into two words
bigrams_separated <- bigrams %>%
  separate(bigram, into = c("word1", "word2"), sep = " ")

# Remove stop words
bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word,
         !word2 %in% stop_words$word) %>%
  # remove numbers 
  filter(!str_detect(word1, "\\d+"),
         !str_detect(word2, "\\d+"))
```

```{r}
# Unite the words back into bigrams
bigram_counts <- bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ") %>%
  count(speaker, bigram, sort = TRUE)
```

```{r}
# Get top 10 bigrams for each candidate
top_bigrams <- bigram_counts %>%
  group_by(speaker) %>%
  top_n(10, n) %>%
  ungroup() %>%
  arrange(speaker, -n)

# View the result
print(top_bigrams)
```

## Plot

### Top words

```{r}
# Plot
top_words %>%
  mutate(word = reorder_within(word, n, speaker)) %>%
  ggplot(aes(word, n, fill = speaker)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~speaker, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  labs(x = "Words", y = "Frequency", title = "Top Words by Candidate")
```

### Top Bigrams

```{r}
# Plot
top_bigrams %>%
  mutate(bigram = reorder_within(bigram, n, speaker)) %>%
  ggplot(aes(bigram, n, fill = speaker)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~speaker, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  labs(x = "Bigrams", y = "Frequency", title = "Top Bigrams by Candidate")
```

## Word Cloud

```{r}
library(wordcloud)
library(RColorBrewer)
```

```{r}
# Filter word counts for each candidate
harris_words <- word_counts %>%
  filter(speaker == "Harris")

trump_words <- word_counts %>%
  filter(speaker == "Trump")
```

### Single Plot

```{r}
wordcloud(words = harris_words$word, 
          freq = harris_words$n, 
          min.freq = 2,
          max.words = 100,
          random.order = FALSE, 
          rot.per = 0.35, 
          colors = brewer.pal(8, "Blues"),
          scale = c(4, 0.5))
title("Harris")
```

### Side-by-Side

```{r}
# Set up the plotting area for two plots
par(mfrow = c(1, 2))

# Word cloud for Harris
wordcloud(words = harris_words$word, 
          freq = harris_words$n, 
          min.freq = 2,
          max.words = 100,
          random.order = FALSE, 
          rot.per = 0.35, 
          colors = brewer.pal(8, "Blues"),
          scale = c(4, 0.5))
title("Harris")

# Word cloud for Trump
wordcloud(words = trump_words$word, 
          freq = trump_words$n, 
          min.freq = 2,
          max.words = 100,
          random.order = FALSE, 
          rot.per = 0.35, 
          colors = brewer.pal(8, "Reds"),
          scale = c(4, 0.5))
title("Trump")
```

## Compare on fixed categories

```{r}
# Define the list of agenda keywords
agenda_keywords <- c("border", "abortion", "economy", "immigration", "health", "security", "israel", "russia", "china", "ukraine")
```

```{r}
# Convert words to lowercase
tidy_data_clean <- tidy_data_clean %>%
  mutate(word = tolower(word))

# Filter words that are in the agenda_keywords list
agenda_data <- tidy_data_clean %>%
  filter(word %in% agenda_keywords)

# Count the frequency of each keyword per candidate
agenda_counts <- agenda_data %>%
  count(speaker, word) %>%
  arrange(speaker, desc(n))
```

```{r}
print(agenda_counts)
```

## Heatmap

```{r}
# Create a heatmap
ggplot(agenda_counts, aes(x = word, y = speaker, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Agenda Keyword Frequencies Heatmap",
       x = " ",
       y = " ") +
  theme_minimal() +
  # remove grid
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
        )
```

## Key Functions Recap

| **Function**                | **Package**     | **Purpose**                                       | **Example Use**                                                                                              |
|-----------------------------|-----------------|---------------------------------------------------|--------------------------------------------------------------------------------------------------------------|
| `readLines()`               | Base R          | Read lines from a text file                       | `readLines("data/debate.txt")`                                                                               |
| `str_detect()`              | `stringr`       | Detect pattern matching in strings                | `str_detect(line, speaker_pattern)`                                                                          |
| `str_split_fixed()`         | `stringr`       | Split string into fixed number of pieces         | `str_split_fixed(line, ":", 2)`                                                                              |
| `str_trim()`                | `stringr`       | Remove leading/trailing whitespace               | `str_trim(split_line[1])`                                                                                    |
| `rbind()`                   | Base R          | Bind rows of data frames together                | `rbind(data, data.frame(speaker = current_speaker, text = current_text))`                                     |
| `skim()`                    | `skimr`         | Quick data summary and exploration                | `skim(df)`                                                                                                   |
| `unique()`                  | Base R          | Extract unique values                            | `unique(df$speaker)`                                                                                         |
| `filter()`                  | `dplyr`         | Filter rows based on conditions                  | `filter(speaker == "HARRIS" \| speaker == "TRUMP")`                                                          |
| `ifelse()`                  | Base R          | Vector conditional function                      | `ifelse(df1$speaker == "HARRIS", "Harris", "Trump")`                                                          |
| `unnest_tokens()`           | `tidytext`      | Tokenize text into individual words              | `unnest_tokens(word, text)`                                                                                  |
| `anti_join()`               | `dplyr`         | Keep rows not matching another table             | `anti_join(stop_words, by = "word")`                                                                          |
| `count()`                   | `dplyr`         | Count occurrences of unique values                | `count(speaker, word, sort = TRUE)`                                                                          |
| `group_by()`                | `dplyr`         | Group data by one or more variables              | `group_by(speaker)`                                                                                          |
| `top_n()`                   | `dplyr`         | Select top n rows by value                       | `top_n(10, n)`                                                                                               |
| `ungroup()`                 | `dplyr`         | Remove grouping from data frame                  | `ungroup()`                                                                                                  |
| `arrange()`                 | `dplyr`         | Sort rows by one or more columns                 | `arrange(speaker, -n)`                                                                                       |
| `separate()`                | `tidyr`         | Separate a column into multiple columns          | `separate(bigram, into = c("word1", "word2"), sep = " ")`                                                     |
| `unite()`                   | `tidyr`         | Unite multiple columns into one                  | `unite(bigram, word1, word2, sep = " ")`                                                                     |
| `mutate()`                  | `dplyr`         | Create or modify columns                         | `mutate(word = tolower(word))`                                                                                |
| `tolower()`                 | Base R          | Convert strings to lowercase                     | `tolower(word)`                                                                                              |
| `reorder_within()`          | `tidytext`      | Reorder factor within groups for visualization   | `reorder_within(word, n, speaker)`                                                                            |
| `ggplot()`                  | `ggplot2`       | Create a new ggplot object                        | `ggplot(aes(word, n, fill = speaker))`                                                                       |
| `aes()`                     | `ggplot2`       | Specify aesthetic mappings (x, y, color, fill)   | `aes(x = word, y = speaker, fill = n)`                                                                       |
| `geom_col()`                | `ggplot2`       | Create a bar chart                               | `geom_col(show.legend = FALSE)`                                                                               |
| `geom_tile()`               | `ggplot2`       | Create a heatmap (tile plot)                     | `geom_tile()`                                                                                                |
| `facet_wrap()`              | `ggplot2`       | Create multiple plots based on categorical var   | `facet_wrap(~speaker, scales = "free_y")`                                                                     |
| `coord_flip()`              | `ggplot2`       | Flip x and y axes                                | `coord_flip()`                                                                                               |
| `scale_x_reordered()`       | `tidytext`      | Apply reordered scale to x-axis                  | `scale_x_reordered()`                                                                                        |
| `scale_fill_gradient()`     | `ggplot2`       | Create gradient color scale for fill              | `scale_fill_gradient(low = "white", high = "red")`                                                            |
| `labs()`                    | `ggplot2`       | Add titles, labels, and captions                 | `labs(title = "...", x = "Words", y = "Frequency")`                                                           |
| `theme_minimal()`           | `ggplot2`       | Apply minimal theme                              | `theme_minimal()`                                                                                            |
| `element_blank()`           | `ggplot2`       | Create blank (invisible) element                 | `element_blank()`                                                                                            |
| `theme()`                   | `ggplot2`       | Customize plot appearance                        | `theme(panel.grid.major = element_blank())`                                                                   |
| `wordcloud()`               | `wordcloud`     | Create a word cloud visualization                | `wordcloud(words = harris_words$word, freq = harris_words$n, max.words = 100)`                                |
| `brewer.pal()`              | `RColorBrewer`  | Generate color palettes                          | `brewer.pal(8, "Blues")`                                                                                     |
| `par()`                     | Base R          | Set or query graphical parameters                | `par(mfrow = c(1, 2))`                                                                                       |
| `data()`                    | Base R          | Load built-in datasets                           | `data("stop_words")`                                                                                         |
