---
title: "Text as Data: The 2024 US Presidential Debate"
---

## Learning objectives

By the end of this case, you should be able to:

- parse a transcript into speaker turns;
- tokenize text while preserving the speaker;
- compare word use with a rate rather than raw counts alone;
- inspect words in their original context; and
- identify interpretive limits of frequency analysis.

## Import and parse the transcript

```{r}
library(tidyverse)
library(tidytext)

transcript_lines <- read_lines("data/debate.txt") |>
  discard(~ str_trim(.x) == "")
```

The transcript marks a new speaker with capital letters followed by a colon.
The parser below combines wrapped lines into one record per speaker turn.

```{r}
parse_transcript <- function(lines) {
  speaker_pattern <- "^[A-Z][A-Z ]+:"
  turns <- list()
  current_speaker <- NA_character_
  current_text <- character()

  save_turn <- function(speaker, text_parts) {
    tibble(
      speaker = speaker,
      text = str_squish(str_c(text_parts, collapse = " "))
    )
  }

  for (line in lines) {
    if (str_detect(line, speaker_pattern)) {
      if (!is.na(current_speaker)) {
        turns[[length(turns) + 1]] <- save_turn(
          current_speaker,
          current_text
        )
      }

      pieces <- str_split_fixed(line, ":", 2)
      current_speaker <- str_trim(pieces[, 1])
      current_text <- str_trim(pieces[, 2])
    } else {
      current_text <- c(current_text, str_trim(line))
    }
  }

  if (!is.na(current_speaker)) {
    turns[[length(turns) + 1]] <- save_turn(
      current_speaker,
      current_text
    )
  }

  bind_rows(turns)
}

turns <- parse_transcript(transcript_lines)

turns |>
  count(speaker, sort = TRUE)
```

Inspect this table before filtering. Unexpected speaker labels can reveal a
formatting problem or a false match in the regular expression.

```{r}
candidate_turns <- turns |>
  filter(speaker %in% c("HARRIS", "TRUMP")) |>
  mutate(speaker = recode(speaker, HARRIS = "Harris", TRUMP = "Trump"))

stopifnot(setequal(unique(candidate_turns$speaker), c("Harris", "Trump")))
```

One row now represents one candidate speaking turn.

## Tokenize and keep denominators

```{r}
all_tokens <- candidate_turns |>
  unnest_tokens(word, text) |>
  filter(!str_detect(word, "^[0-9]+$"))

token_totals <- all_tokens |>
  count(speaker, name = "all_tokens")

token_totals
```

The candidates produced different numbers of tokens. A raw frequency can be
higher simply because one person spoke more. Keep the denominator before
removing stop words.

```{r}
data("stop_words")

content_tokens <- all_tokens |>
  anti_join(stop_words, by = "word")

word_rates <- content_tokens |>
  count(speaker, word, name = "uses") |>
  left_join(
    token_totals,
    by = "speaker",
    relationship = "many-to-one"
  ) |>
  mutate(uses_per_1000_tokens = uses / all_tokens * 1000)
```

Stop-word removal is an editorial and linguistic choice. Negations such as
“not” can matter greatly in political speech, and a generic English stop-word
list may remove them. Inspect the list and document any changes.

## Compare the most frequent content words

```{r}
top_words <- word_rates |>
  group_by(speaker) |>
  slice_max(
    uses_per_1000_tokens,
    n = 12,
    with_ties = FALSE
  ) |>
  ungroup()

top_words
```

```{r}
top_words |>
  mutate(word = reorder_within(word, uses_per_1000_tokens, speaker)) |>
  ggplot(aes(x = uses_per_1000_tokens, y = word, fill = speaker)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ speaker, scales = "free_y") +
  scale_y_reordered() +
  labs(
    title = "Frequently used content words differed between the candidates",
    subtitle = "Uses per 1,000 candidate tokens after standard stop-word removal",
    x = "Uses per 1,000 tokens",
    y = NULL,
    caption = "Source: supplied debate transcript; analysis choices affect results"
  ) +
  theme_minimal()
```

Rate normalization improves comparability but does not measure importance,
sincerity, policy attention, or audience effect.

## Bigrams preserve a little context

```{r}
bigrams <- candidate_turns |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate_wider_delim(
    bigram,
    delim = " ",
    names = c("word_1", "word_2"),
    too_few = "align_start"
  ) |>
  filter(
    !word_1 %in% stop_words$word,
    !word_2 %in% stop_words$word,
    !is.na(word_2)
  ) |>
  unite(bigram, word_1, word_2, sep = " ") |>
  count(speaker, bigram, sort = TRUE)

bigrams |>
  group_by(speaker) |>
  slice_max(n, n = 10, with_ties = FALSE) |>
  ungroup()
```

Bigrams retain adjacent words but still lose sentence meaning, sarcasm,
questions, quotation, and who or what a pronoun refers to.

## Return to the transcript

Any word that becomes part of a story must be checked in context:

```{r}
candidate_turns |>
  filter(str_detect(str_to_lower(text), "immigration")) |>
  select(speaker, text)
```

Read several examples, not just the most convenient one. Check whether the
candidate is stating a position, responding to a moderator, quoting an
opponent, or denying a claim.

## Claim limits

Word frequency can support a sentence such as “the word appeared at a higher
rate in Candidate A's transcript.” It does not by itself support:

- “Candidate A cared more about the issue”;
- “the issue dominated the debate” without a coding rule;
- “Candidate A was more truthful”;
- sentiment or intent inferred from an isolated token; or
- audience impact.

Transcript accuracy, moderation, speaking time, interruptions, tokenization,
and stop-word choices all affect the result.

## Independent task

Choose one topic and create a small, transparent dictionary of related terms.
Then:

1. calculate topic-term uses per 1,000 tokens by candidate;
2. inspect at least five turns in context for each candidate;
3. record one false positive and one missed expression;
4. revise the dictionary once; and
5. write a 200-word methods note that distinguishes automated counts from your
   qualitative interpretation.
