---
title: "Debugging"
---

## When code does not work

Errors are a normal part of learning R. The goal is not to memorize every
message. It is to follow the same small set of checks each time.

## Errors, warnings, and messages

- An **error** stops the code.
- A **warning** allows the code to finish but reports a possible problem.
- A **message** gives information about what a function did.
- An **unexpected result** may appear without any warning, so always inspect
  the output.

::: {.callout-tip title="Read before changing"}
Read the final line of an error and identify the object, column, or function it
mentions. Do not add several new functions at once in the hope that the error
will disappear.
:::

## A simple debugging routine

1. Run the code again and read the complete message.
2. Check the spelling and capitalization of object and column names.
3. Confirm that the required package was loaded.
4. Inspect the table with `glimpse()`.
5. Run one pipeline step at a time.
6. Change one thing, then run the code again.

## Example: a column name is wrong

```{r}
library(tidyverse)

holdings <- tibble(
  company = c("A Ltd", "B Ltd", "C Ltd"),
  market_value_nok = c(120, 80, 100)
)
```

This code asks for a column that does not exist:

```{r}
#| error: true
holdings |>
  arrange(desc(market_value))
```

`glimpse()` shows the exact column names:

```{r}
glimpse(holdings)
```

Make the smallest correction:

```{r}
holdings |>
  arrange(desc(market_value_nok))
```

## Example: find the first bad step

Break a longer pipeline into small objects:

```{r}
step_1 <- holdings |>
  filter(market_value_nok >= 100)

step_2 <- step_1 |>
  mutate(value_millions = market_value_nok / 1000000)

step_1
step_2
```

The code runs, but the second result is unexpectedly small. The toy values were
not measured in full kroner. This is a unit problem, not a syntax problem.
Code that runs can still be wrong.

## Check missing values with familiar functions

Use `summarise()` to display simple checks. `is.na()` identifies missing
values, so `sum(is.na(company))` counts missing company names:

```{r}
holdings |>
  summarise(
    rows = n(),
    missing_company_names = sum(is.na(company)),
    smallest_value = min(market_value_nok)
  )
```

The expected results are three rows, no missing company names, and a positive
smallest value. A useful check begins with a reasonable expectation about the
source.

## Ask for help effectively

Include:

1. what you expected;
2. what happened;
3. the exact error or warning;
4. the smallest example that reproduces it; and
5. the packages being used.

Do not share confidential data or passwords when asking for help.

## Practice

Find and correct the two problems in this code:

```{r}
#| error: true
story <- tibble(
  outlet = c("A", "B", "C"),
  articles = c(12, NA, 20)
)

story |>
  filter(Article >= 15) |>
  summarise(average = mean(articles))
```

Explain the column-name problem and decide how the missing value should be
handled. Do not add `na.rm = TRUE` until you can explain what the `NA` means.

## Takeaways

| Practice | Why it helps |
|---|---|
| Read the message | Identifies the immediate problem |
| Check spelling and case | Finds many object and column errors |
| `glimpse()` | Shows names, types, and example values |
| Run one step at a time | Locates the first step that changes unexpectedly |
| Change one thing | Makes cause and effect easier to understand |
| Display simple checks | Catches results that are possible in R but implausible in the story |
