---
title: "Debugging Without Guessing"
---

## Learning objectives

By the end of this chapter, you should be able to:

- distinguish errors, warnings, messages, and unexpected results;
- locate the first failing step in a pipeline;
- inspect object names, classes, dimensions, and missing values;
- create a small reproducible example; and
- explain what you changed and why.

## Four kinds of trouble

An **error** stops execution. A **warning** allows execution to continue but
signals a possible problem. A **message** provides information. An
**unexpected result** may produce no warning at all, which is why validation is
part of debugging.

### Error

```{r}
#| error: true
mean(c("10", "20"))
```

The function received text when it expected numbers. The error is evidence
about the input, not an instruction to add random code.

### Warning

```{r}
#| warning: true
as.numeric(c("10", "not reported", "20"))
```

R returns numbers and an `NA`, but warns that one conversion failed. A reporter
should locate that source value before deciding how to handle it.

### Expected `NA`, with no warning

```{r}
mean(c(10, 20, NA))
```

This result is not an error and normally produces no warning. R is following
the rule that a mean is unknown when an input is unknown. Use
`na.rm = TRUE` only after checking what the missing value means.

## A repeatable debugging routine

When code fails, use the same sequence each time.

1. **Read the final error line.** Copy it into your notes.
2. **Identify the object and function named in the error.**
3. **Check that the object exists.** Run `ls()` or type its name.
4. **Inspect its structure.** Use `class()`, `names()`, `dim()`, and `glimpse()`.
5. **Run the pipeline one step at a time.** Save intermediate results.
6. **Reduce the problem.** Test a few rows and only the relevant columns.
7. **Change one thing.** Re-run from a fresh session after the fix.

## Example: a missing column

```{r}
library(tidyverse)

holdings <- tibble(
  company = c("A Ltd", "B Ltd", "C Ltd"),
  market_value_nok = c(120, 80, 100)
)
```

This code fails because the column name is wrong:

```{r}
#| error: true
holdings |>
  arrange(desc(market_value))
```

Inspect names before editing:

```{r}
names(holdings)
```

Now make the smallest correction:

```{r}
holdings |>
  arrange(desc(market_value_nok))
```

Avoid “fixing” the problem by renaming every column or rewriting the whole
pipeline. Small changes make cause and effect easier to see.

## Example: find the first bad step

Suppose a long pipeline gives an unexpected result. Break it apart:

```{r}
step_1 <- holdings |>
  filter(market_value_nok >= 100)

step_2 <- step_1 |>
  mutate(value_millions = market_value_nok / 1e6)

step_1
step_2
```

The code runs, but the new values are tiny because the toy input was not
actually measured in full NOK. That is a **unit error**, not a syntax error.
Checking units and plausible ranges is therefore part of debugging.

## Assertions turn expectations into checks

Use `stopifnot()` for assumptions that must be true:

```{r}
stopifnot(nrow(holdings) == 3)
stopifnot(!anyNA(holdings$company))
stopifnot(all(holdings$market_value_nok >= 0))
```

For a real analysis, useful assertions include:

- expected years are present;
- a join does not change the row count unexpectedly;
- key identifiers are not missing;
- shares fall in a plausible range; and
- subtotals reconcile with a published total.

An assertion should represent a justified expectation. Do not assert that all
financial values must be positive if the source can legitimately contain short
positions.

## Build a reproducible example

When asking for help, include the smallest data and code that still reproduce
the problem:

```{r}
problem_data <- tibble(
  amount = c("1,200", "not reported", "900")
)

problem_data |>
  mutate(amount_number = readr::parse_number(amount))
```

A useful help request states:

1. what you expected;
2. what happened instead;
3. the exact error or warning;
4. a small example; and
5. the output of `sessionInfo()` when package versions may matter.

## Independent task

Debug this code without rewriting it from scratch:

```{r}
#| error: true
story <- tibble(
  outlet = c("A", "B", "C"),
  articles = c(12, NA, 20)
)

story |>
  filter(Article >= 15) |>
  summarise(average = mean(articles))
```

Your answer should:

- identify both problems;
- show the corrected code;
- explain the decision about the missing value; and
- add one assertion that would catch a future data problem.
