---
title: "HK Graduates Employment"
---

::: {.callout-important title="The new challenge: split, reshape, and recombine mixed data"}
The table mixes two levels of detail. Full-time employment is divided into
occupation rows, while every other employment status is already a single
total. We must separate these structures, aggregate them to the same level,
recombine them, and only then construct a denominator.
:::

## Learning objectives

By the end of this case, you should be able to:

- recognize a table that mixes different levels of detail;
- split rows with different structures;
- recombine detailed occupation rows into comparable status totals;
- reshape the status totals into columns;
- build a denominator from mutually exclusive categories; and
- calculate a weighted rate.

## Why the familiar workflow is not enough

The Norway holdings have a relatively simple row structure. The
[Hong Kong graduate employment data](https://data.gov.hk/en-data/dataset/hk-ugc-ugc-student-ges-statistics)
are more complicated. A single `group_by()` can produce a result, but first
separating the two row structures makes the aggregation rule visible and
easier to check.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `11-employment.Rmd`. Download
[`employment.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/employment.csv)
and save it in `data/`. The familiar folder structure lets you concentrate on
the more complicated row structure.
:::

## Set up the case

```{r}
library(tidyverse)

employment <- read_csv("data/employment.csv") |>
  rename(
    academic_year = `Academic Year`,
    university = University,
    level = `Level of study`,
    status = `Employment Situation`,
    occupation = Occupation,
    headcount = `Number of Graduates (Headcount)`
  )

glimpse(employment)
```

## Understand the row structure

Count the combinations of status and occupation:

```{r}
employment |>
  count(status, occupation) |>
  arrange(status, desc(n))
```

For full-time employment, the occupation column contains many detailed
categories. For other statuses, it contains “Not in Full-time Employment.”

One raw row therefore represents one headcount cell for an academic year,
university, study level, employment status, and—only for full-time
employment—occupation.

```{mermaid}
flowchart TD
  A["Year × university × level"] --> B["Full-time employment"]
  A --> C["Other employment status"]
  B --> D["Several occupation rows"]
  C --> E["One row"]
```

Counting raw rows would count categories, not graduates.

## The strategy

We will solve the mixed structure in six steps:

1. split full-time and other-status rows;
2. aggregate the detailed occupation rows;
3. recombine comparable status totals;
4. reshape statuses into columns;
5. construct a denominator; and
6. add headcounts before calculating a combined rate.

Each step changes what one row represents. State that unit after every major
transformation.

## Strategy 1: split the two row structures

Use `filter()` to create one table for detailed full-time records and another
for the already aggregated statuses:

```{r}
full_time <- employment |>
  filter(status == "FT employment")

other_statuses <- employment |>
  filter(status != "FT employment")
```

The split is based on data structure, not on which outcome is more important.
It prevents the occupation rows from being treated as if they were already
comparable with unemployment or further-study totals.

## Strategy 2: aggregate the detailed rows

Within the full-time table, add occupation headcounts for each cohort:

```{r}
full_time_totals <- full_time |>
  group_by(academic_year, university, level, status) |>
  summarise(headcount = sum(headcount))

full_time_totals
```

One row now represents the full-time total for one
year-university-level cohort.

The other-status table already has one row for each status in each cohort. Keep
only the columns needed for the next step:

```{r}
other_status_totals <- other_statuses |>
  select(academic_year, university, level, status, headcount)
```

## Strategy 3: recombine comparable totals

The two tables now have the same columns and the same level of detail.
`bind_rows()` stacks them:

```{r}
status_totals <- bind_rows(
  full_time_totals,
  other_status_totals
)

status_totals |>
  count(status)
```

One row now represents one employment status within one cohort. Recombining
before the full-time rows were aggregated would recreate the original problem.

## Strategy 4: reshape statuses into columns

`pivot_wider()` gives each status its own column:

```{r}
cohorts <- status_totals |>
  pivot_wider(
    names_from = status,
    values_from = headcount
  ) |>
  rename(
    full_time = `FT employment`,
    further_studies = `Further studies`,
    others = Others,
    underemployed = Underemployed,
    unemployed = Unemployed
  )

cohorts
```

One row now represents one year-university-level group.

## Strategy 5: build the denominator

The employment statuses are mutually exclusive. Add them to obtain the total
reported graduate outcomes, then calculate rates:

```{r}
cohorts <- cohorts |>
  mutate(
    total = full_time + further_studies + others + underemployed + unemployed,
    full_time_rate = full_time / total,
    unemployment_rate = unemployed / total
  )

cohorts
```

The numerator and denominator now refer to the same cohort. If any status is
missing, investigate it before treating the value as zero.

## Strategy 6: calculate a weighted university rate

To combine study levels, add the headcounts first and calculate the rate from
those totals:

```{r}
university_rates <- cohorts |>
  group_by(academic_year, university) |>
  summarise(
    full_time = sum(full_time),
    unemployed = sum(unemployed),
    total = sum(total)
  ) |>
  mutate(
    full_time_rate = full_time / total,
    unemployment_rate = unemployed / total
  )

university_rates
```

Do not simply average the study-level rates. That would give a small cohort the
same influence as a large cohort. Adding the headcounts first weights the
result by the number of graduates.

::: {.callout-important title="A descriptive comparison is not a league table"}
Universities differ in subjects, study levels, student populations, and career
paths. The rates describe reported outcomes. They do not prove that attending
one university caused a better or worse outcome.
:::

## Optional extension: examine one occupation

Suppose the question concerns authors, journalists, and related writers. First
stay within full-time employment, then create an indicator column before
grouping. `if_else()` assigns one value when a condition is true and another
when it is false:

```{r}
journalism_share <- employment |>
  filter(status == "FT employment") |>
  mutate(
    journalism_headcount = if_else(
      occupation == "Authors, Journalists and Related Writers",
      headcount,
      0
    )
  ) |>
  group_by(academic_year, university, level) |>
  summarise(
    journalists = sum(journalism_headcount),
    all_full_time = sum(headcount)
  ) |>
  mutate(share_of_full_time = journalists / all_full_time)

journalism_share
```

The denominator is all full-time employed graduates in the same cohort,
because that is the comparison being made.

## Practice

Choose unemployment, underemployment, or further study. Create a university-
year rate by:

1. describing the raw row structure;
2. splitting the detailed and already aggregated rows;
3. aggregating full-time occupation rows and recombining the tables;
4. reshaping the statuses;
5. adding headcounts before calculating the rate; and
6. writing one sentence explaining why the result is not a causal university
   ranking.

## Takeaways

| Function or idea | What it does in this case |
|---|---|
| `filter()` | Splits rows with different structures |
| `group_by()` + `summarise()` | Combine detailed occupation rows into status totals |
| `bind_rows()` | Recombines tables after their row structures match |
| `pivot_wider()` | Put mutually exclusive statuses into columns |
| `mutate()` | Build totals and rates with explicit denominators |
| Weighted rate | Adds headcounts before calculating the combined rate |

::: {.callout-tip title="Want more control?"}
The official [dplyr reference](https://dplyr.tidyverse.org/reference/index.html)
and [tidyr reference](https://tidyr.tidyverse.org/reference/) cover further
grouping and reshaping options for more complicated official statistics.
:::
