---
title: "Aggregated Official Data: Hong Kong Graduate Employment"
---

## Why this case comes later

The NBIM chapters use a relatively simple holding-level grain. This official
Hong Kong dataset is harder: full-time employment is split into many occupation
rows, while every other employment status is stored in one row labelled “Not
in Full-time Employment.” The row structure is nested and uneven.

That complexity is pedagogically valuable once students can already filter,
group, join, and validate. It teaches a central data-journalism lesson: a tidy-
looking table can still contain incompatible levels of aggregation.

The source is the
[Hong Kong Government open-data portal](https://data.gov.hk/en-data/dataset/hk-ugc-ugc-student-ges-statistics),
which describes graduate outcomes for full-time UGC-funded programmes.

## Learning objectives

By the end of this chapter, you should be able to:

- diagnose a nested unit of analysis;
- collapse detailed rows to comparable status cells;
- calculate a weighted rate with the correct denominator;
- distinguish row frequency from graduate headcount; and
- explain why a university ranking may reflect programme composition.

## Import and rename

```{r}
library(tidyverse)
library(scales)

employment <- read_csv(
  "data/employment.csv",
  show_col_types = FALSE
) |>
  rename(
    academic_year = `Academic Year`,
    university = University,
    level = `Level of study`,
    status = `Employment Situation`,
    occupation = Occupation,
    headcount = `Number of Graduates (Headcount)`
  ) |>
  mutate(
    year_start = as.integer(str_sub(academic_year, 1, 4))
  )

glimpse(employment)
```

## Audit the grain before calculating a rate

```{r}
employment |>
  count(status, occupation) |>
  arrange(status, desc(n))
```

For full-time employment, `occupation` contains many detailed categories. For
Further studies, Others, Underemployed, and Unemployed, the occupation is always
“Not in Full-time Employment.”

Therefore, one raw row is best described as:

> one headcount cell for an academic year, university, study level, employment
> status, and—only for full-time employment—occupation.

The data mix two grains:

```{mermaid}
flowchart TD
  A["Year × university × level"] --> B["Full-time employment"]
  A --> C["Other status"]
  B --> D["Many occupation rows"]
  C --> E["One 'Not in Full-time Employment' row"]
```

Counting rows would count occupation categories, not graduates.

## Collapse to comparable status cells

First sum the detailed full-time occupation rows so every status has the same
grain:

```{r}
status_cells <- employment |>
  summarise(
    headcount = sum(headcount),
    .by = c(
      academic_year,
      year_start,
      university,
      level,
      status
    )
  )

status_cells |>
  count(status)
```

One row now represents one employment status within a year–university–level
cohort.

Validate that collapsing did not change the number of graduates represented:

```{r}
stopifnot(
  sum(status_cells$headcount) == sum(employment$headcount)
)
```

## Pivot statuses and construct denominators

```{r}
cohorts <- status_cells |>
  pivot_wider(
    names_from = status,
    values_from = headcount,
    values_fill = 0
  ) |>
  rename(
    ft = `FT employment`,
    further_studies = `Further studies`,
    others = Others,
    underemployed = Underemployed,
    unemployed = Unemployed
  ) |>
  mutate(
    total = ft + further_studies + others + underemployed + unemployed,
    ft_rate = ft / total,
    unemployment_rate = unemployed / total
  )

cohorts
```

One row now represents one year–university–level cohort. The denominator is the
sum of mutually exclusive employment statuses, after full-time occupations
have been collapsed.

```{r}
stopifnot(sum(cohorts$total) == sum(employment$headcount))
stopifnot(all(cohorts$ft_rate >= 0 & cohorts$ft_rate <= 1))
stopifnot(all(cohorts$unemployment_rate >= 0 & cohorts$unemployment_rate <= 1))
```

## Calculate a weighted university rate

```{r}
latest_year <- max(cohorts$year_start)

latest_university <- cohorts |>
  filter(year_start == latest_year) |>
  summarise(
    ft = sum(ft),
    unemployed = sum(unemployed),
    total = sum(total),
    ft_rate = ft / total,
    unemployment_rate = unemployed / total,
    .by = university
  ) |>
  arrange(desc(ft_rate))

latest_university
```

This calculation weights each study-level cohort by its graduate headcount.
The following tempting alternative is not equivalent:

```{r}
cohorts |>
  filter(year_start == latest_year) |>
  summarise(
    unweighted_mean_ft_rate = mean(ft_rate),
    .by = university
  )
```

An unweighted mean gives a small cohort the same influence as a large cohort.
Use the rate-of-sums for an overall graduate rate unless the research question
explicitly concerns the typical programme-level rate.

## Visualize without overclaiming

```{r}
latest_university |>
  ggplot(aes(x = ft_rate, y = fct_reorder(university, ft_rate))) +
  geom_col(fill = "#1769AA") +
  geom_text(
    aes(label = percent(ft_rate, accuracy = 0.1)),
    hjust = -0.1,
    size = 3.2
  ) +
  scale_x_continuous(
    labels = label_percent(),
    limits = c(0, max(latest_university$ft_rate) * 1.12)
  ) +
  labs(
    title = "Full-time employment rates varied across Hong Kong universities",
    subtitle = str_wrap(
      paste0(
        "Full-time employment as a share of reported graduate outcomes, ",
        latest_year,
        "/",
        str_sub(as.character(latest_year + 1), 3, 4)
      ),
      width = 85
    ),
    x = "Full-time employment rate",
    y = NULL,
    caption = paste(
      "Rates combine study levels and are weighted by graduate headcount.",
      "Source: University Grants Committee via DATA.GOV.HK."
    )
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0, size = 8.5)
  )
```

The chart is descriptive. Universities differ in study levels, disciplines,
student populations, response or reporting conditions, and labour-market
pathways. The data do not establish institutional quality or a causal effect of
attending one university.

## A denominator within full-time employment

Suppose the story concerns graduates recorded as authors, journalists, and
related writers. The relevant denominator is full-time employed graduates, not
all graduates:

```{r}
journalism_share <- employment |>
  summarise(
    journalists = sum(
      headcount[
        status == "FT employment" &
          occupation == "Authors, Journalists and Related Writers"
      ]
    ),
    all_ft = sum(headcount[status == "FT employment"]),
    .by = c(academic_year, university, level)
  ) |>
  mutate(share_of_ft = journalists / all_ft)

journalism_share
```

The numerator and denominator now refer to the same status and cohort. A
different denominator would answer a different question.

## Independent task

Choose either unemployment, underemployment, or further study. Produce:

1. a diagram or sentence defining the raw grain;
2. a correctly collapsed cohort table;
3. a weighted university-year rate;
4. a comparison with the unweighted mean of cohort rates;
5. three validation checks; and
6. a 150-word editor's note explaining why the result should not be presented
   as a causal university league table.
