---
title: "Final Project"
---

## Plan the project

A final project should use the same workflow as the main chapters:

1. choose a focused reporting question;
2. record the source and access date;
3. state what one row represents;
4. clean and analyse with a reproducible R Markdown document;
5. check the main calculations;
6. create one or more clear charts;
7. report beyond the dataset when an explanation is needed; and
8. explain what the evidence cannot show.

For the final story, create a **new RStudio Project** so it can be shared as one
complete piece of work. Use the same simple structure practised throughout the
book: a `data` folder, an `outputs` folder, and one main R Markdown document.

## Begin with a focused question and story hypothesis

A **story hypothesis** is a statement to test, not a conclusion already
assumed to be true. It gives the analysis direction while allowing the
evidence to confirm, revise, or reject the initial idea.

For example, a project might begin with:

> A small number of categories account for a large share of the reported
> total.

Turn the broad idea into a question that names the population, measure, place,
and period. Avoid wording that assumes a cause or outcome the dataset does not
record.

## Keep the downloaded file unchanged

Preserve the downloaded source file. If you need to save a cleaned table, give
it a new name using the tidyverse `write_csv()` function:

```{r}
#| eval: false
write_csv(analysis_data, "data/analysis_data.csv")
```

Load that prepared table in the document that uses it:

```{r}
#| eval: false
analysis_data <- read_csv("data/analysis_data.csv")
```

This keeps the project inside the same CSV and tidyverse workflow used
throughout the book.

## Choose a chart that matches the question

| Question | Common starting chart |
|---|---|
| How often does each category appear? | Bar chart |
| How is one numeric variable distributed? | Histogram or box plot |
| Which categories have larger values? | Ordered bar chart or dot plot |
| How do distributions differ across groups? | Box plot |
| Are two numeric variables related? | Scatterplot |
| How did a value change over time? | Line chart |
| Does geographic pattern matter? | Map supported by a table |

The chart title should name the measure actually calculated. Avoid a line
between observations that do not have a meaningful order.

## Save a chart

```{r}
#| eval: false
ggsave("outputs/main-chart.png", main_chart)
```

`ggsave()` has optional arguments for width, height, and resolution when a
publishing platform has specific requirements. Begin with the filename and
chart object, then add those options only if they are needed.

## Make an evidence table before drafting

An evidence table connects possible claims to calculations and reporting still
needed. Fill it with actual values and source checks rather than drafting from
memory or from the appearance of a chart:

| Possible claim | Evidence in the data | What still needs checking? |
|---|---|---|
| One category has the largest total | Group summary and source row | Confirm the value and unit |
| One group has the largest share | Numerator and full denominator | Check excluded or missing categories |
| A measure increased over time | Comparable annual observations | Investigate changes in definitions |
| Two variables appear related | Table and scatterplot | Find evidence that could explain the pattern |

The table can confirm, narrow, or reject the original story hypothesis.

## Separate description from explanation

A dataset can identify and measure a pattern without explaining why it
occurred. Explanations may require documents, interviews, subject expertise,
or another dataset.

::: {.callout-important title="A pattern is not an explanation"}
Use the data to describe what changed, differed, or appeared related. Treat a
possible explanation as a reporting lead until additional evidence supports
it. A plausible explanation does not become a fact because it fits the chart.
:::

## Verify the important numbers

Before drafting:

1. check important rows against the original source;
2. make sure every percentage has a clearly named denominator;
3. compare chart values with the table used to create the chart;
4. inspect unmatched categories from joins or maps;
5. record the reporting period and download date; and
6. write down at least one important limitation.

## Write a short methods note

A reader-friendly methods note should state:

- who published the data;
- what period the analysis covers;
- what one row represents;
- the important filters, groups, and calculations;
- how missing or unmatched values were handled; and
- what the data cannot establish.

Write the note for a reader, not as a list of R commands. Name the analytical
decisions and their consequences.

## Organize the story

A concise data story can follow this order:

1. **Lead:** the strongest verified finding.
2. **Context:** what the data cover and why the finding matters.
3. **Evidence:** a clear table or chart with exact values.
4. **Reporting:** documents or interviews that test explanations.
5. **Limitations:** what the evidence cannot show.
6. **Methods:** source, period, unit of analysis, and main calculation.

The goal is not to include every calculation. Select the evidence that answers
the reporting question and helps the reader understand why it matters.

## Prepare the final project proposal

Before completing the full story, submit:

- one focused reporting question;
- one testable story hypothesis;
- a four-row evidence table;
- one draft chart;
- two outside sources to consult;
- a draft methods paragraph; and
- one claim rejected because the available evidence cannot support it.

## Final project checklist

- The question is focused enough to answer with the available data.
- The original data are preserved unchanged.
- The source, access date, and reporting period are recorded.
- The meaning of one row is stated.
- Important filters and exclusions are explained.
- Every percentage has a clear denominator.
- Join keys and unmatched rows are checked.
- The main numbers agree with the source or a second calculation.
- Every major possible claim appears in the evidence table.
- Chart titles, labels, units, and sources are complete.
- Explanations and causal language have evidence beyond a visible pattern.
- Relevant documents, experts, or interviews have been used to test explanations.
- The limitations are written for the reader.
- The R Markdown document knits after restarting R.

## Takeaways

| Function or practice | Why it matters |
|---|---|
| `write_csv()` | Saves a prepared table without overwriting the source |
| `read_csv()` | Reloads the prepared table with the familiar import workflow |
| `ggsave()` | Saves a chart for publication |
| Story hypothesis | Gives the analysis an idea that can be tested and revised |
| Evidence table | Connects possible claims to calculations and checks |
| Outside reporting | Tests explanations the dataset cannot provide |
| Source and methods note | Makes the reporting process transparent |
| Rejected claim | Shows that the evidence was evaluated rather than merely collected |
| Fresh knit | Checks that the document contains every step needed to reproduce it |

## Previous student projects

The projects below were created by previous students in **JMSC1003:
Fundamentals of Data Journalism**. New students can browse them to understand
the expected scope of the final project and see the different forms a
data-driven story can take.

As you look through them, notice how each student:

- chooses a focused reporting question;
- uses data as evidence;
- selects and explains the most important findings; and
- combines writing and visualization to guide the reader.

The topics and approaches vary. They are examples of what is possible, not
templates that every project must follow.

### Fall 2024

- [HK Elderly Abuse](https://yongershen.quarto.pub/yonger-final-project-dj/) — *Shen, Yonger*
- [US Mass Shootings in 2024](https://cindy-ningxy.quarto.pub/xiuyuan-final-project/) — *Ning, Xiuyuan Cindy*
- [China's Movie Revenue](https://angelicawan927.quarto.pub/angelicazhixin_final/) — *Wan, Zhixin Angelica*
- [Child Abuse Cases in Hong Kong](https://tesstang.quarto.pub/xinyue_final_project/) — *Tang, Xinyue Tess*
- [Esports Revenue Trends](https://eveqiao.quarto.pub/yushan_finalproject/) — *Qiao, Yushan Eve*
- [Tracking Carbon Emissions](https://lanqingm.quarto.pub/carbon-emissions/) — *Mao, Lanqing*

### Spring 2025

#### Undergraduate students

- [War on Drugs in the Philippines (2016–2022)](https://sarahjody.quarto.pub/sarahjody-finalproject-website/) — *Siquig Castañeto, Sarah Jody*
- [The Journey of Women in the Olympics (1896–2024)](https://snow-brandon.quarto.pub/brandon-finalproject/) — *Feng, Brandon*
- [Global Migration Deaths](https://ethanlim.quarto.pub/ethan-website/) — *Lim, Haan Hoong Ethan*
- [Trends in SEN Student Numbers (2019–2024)](https://chloechan.quarto.pub/demo-website/) — *Chan, Lok Yiu Chloe*
- [Olympic Performance Over Time](https://huangshixian.quarto.pub/olympic-performance/) — *Huang, Shixian*
- [Global Video Game Preference 2024](https://chonwun-shen.quarto.pub/shen-chon-wuns-final-project/) — *Shen, Chon Wun*
- [Government Funding Allocation to National Sports Associations](https://joshuakwok.quarto.pub/joshuakwok-quarto/analysis.html) — *Kwok, Joshua*

#### Master's students

- [Global EV Revolution](https://lexitian.quarto.pub/website/) — *Tian, Xinlu Lexi*
- [Global HIV/AIDS Condition](https://cindyxie.quarto.pub/cindy-website/) — *Xie, Xindi Cindy*
- [Global Animal Health Outbreak Report](https://rubypeng.quarto.pub/httpquartopubcomrubypengglobal-animal-health-outbreak-report/) — *Peng, Ruixin Ruby*


### Fall 2025

#### Undergraduate students

- [Global Energy Statistics](https://harrisonnazer.quarto.pub/harrison-website/) - *Jonah, Nazer Harrison*
- [The Fight Against Infant Mortality](https://haisonpham.quarto.pub/sons-final-project/) - *Pham, Son Haisong*
- [HK Rental Pressure](https://gao-xiangze-zoe.quarto.pub/zoes-final/) - *Gao, Xiangze Ze*
- [Global Health about Respiratory Disease](https://claire77.quarto.pub/claires-final-project/) - *Gu, Raomeng Claire*
- [Plastic Pollution and GDP Patterns Across East & Southeast Asia](https://yunseonkim.quarto.pub/yunseon-finalproject-website/) - *Kim, Yunseon*

#### Master's students

- [CO2 Emissions](https://vickyzhang0328.quarto.pub/vickyzhang/) - *Zhang, Linzhuo*
- [AI Investment: Future or Bubble](https://caoruopeng.quarto.pub/ai-investment-future-or-bubble/) - *Cao, Ruopeng*
- [Femicide — A Persistent Crisis Marked by Regional Disparities and Gendered Violence](https://chenyinuo.quarto.pub/global-femicides/) - *Chen, Yinuo*
- [Afghan Refugee Movements in a Changing World (2019–2024)](https://raewang852.quarto.pub/afghan-asylum-trends/) - *Wang, Yixin*
- [World Military Spending](https://eunicexu.quarto.pub/world-military-spending/) - *Xu, Ran*
- [Hong Kong’s Refugee Asylum Pattern in a Global Context](https://li-jiachen02.quarto.pub/li-jiachen-data-journalism-final-project/) - *Li, Jiachen*








