---
title: "Metric Design: Billboard Hot 100"
---

## Learning objectives

By the end of this case, you should be able to:

- state the difference between a row count and a distinct-song count;
- design a metric before ranking artists;
- label charts with the measure actually calculated;
- define a time period with exact dates; and
- identify credit and identity problems in cultural data.

## Import and clean

```{r}
library(tidyverse)

hot100 <- read_csv(
  "data/hot100.csv",
  show_col_types = FALSE
) |>
  rename(
    date = Date,
    song = Song,
    artist = Artist,
    rank = Rank,
    last_week = `Last Week`,
    peak_position = `Peak Position`,
    weeks_in_chart = `Weeks in Charts`,
    image_url = `Image URL`
  ) |>
  mutate(
    weeks_in_chart = parse_number(weeks_in_chart)
  )

glimpse(hot100)
```

One row represents one song's position on one weekly chart date. A song that
appears for 20 weeks contributes 20 rows. This grain makes several plausible
“most successful” metrics possible—and they are not interchangeable.

## Define the period exactly

```{r}
period_start <- as.Date("2000-01-01")

number_ones <- hot100 |>
  filter(rank == 1, date >= period_start)

range(number_ones$date)
```

“Since 2000” is clearer than “in the 21st century” when the code uses a date
boundary. If the source ends in 2024, say so in the subtitle or methods.

## Metric 1: weeks appearing at No. 1

```{r}
artist_weeks_at_number_one <- number_ones |>
  count(artist, name = "weeks_at_number_one", sort = TRUE)

artist_weeks_at_number_one |>
  slice_head(n = 10)
```

This metric counts weekly No. 1 chart appearances. If one song remains No. 1
for ten weeks, it contributes ten to the artist's total.

## Metric 2: distinct No. 1 songs

```{r}
artist_number_one_songs <- number_ones |>
  distinct(artist, song) |>
  count(artist, name = "number_one_songs", sort = TRUE)

artist_number_one_songs |>
  slice_head(n = 10)
```

This metric counts unique artist–song pairs that reached No. 1 at least once.
The leading artist can differ from the weekly-appearance ranking because one
long-running song and several shorter songs represent different kinds of chart
success.

## Metric 3: longest-running No. 1 song

```{r}
song_weeks_at_number_one <- number_ones |>
  count(
    artist,
    song,
    name = "weeks_at_number_one",
    sort = TRUE
  )

song_weeks_at_number_one |>
  slice_head(n = 15)
```

The song and artist must remain together. Grouping by song title alone could
combine unrelated recordings with the same name.

## Put the metric in the chart title and axis

```{r}
top_artists <- artist_number_one_songs |>
  slice_head(n = 10)

top_artists |>
  ggplot(
    aes(
      x = number_one_songs,
      y = fct_reorder(artist, number_one_songs)
    )
  ) +
  geom_col(fill = "#1769AA") +
  geom_text(
    aes(label = number_one_songs),
    hjust = -0.2,
    size = 3.2
  ) +
  scale_x_continuous(
    breaks = scales::breaks_width(1),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Artists with the most distinct No. 1 songs since 2000",
    subtitle = paste0(
      "Each artist–song pair is counted once; source through ",
      max(year(hot100$date))
    ),
    x = "Distinct songs that reached No. 1",
    y = NULL,
    caption = "Source: Billboard Hot 100 dataset supplied with the course"
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

Do not label the y-axis “number of songs” if the code counts weekly rows. That
small mismatch changes the finding.

## Credit and identity problems

The `artist` field uses the credit published for each song. Collaborations can
appear as one combined string, while an artist may also appear alone. Names,
punctuation, featured-artist conventions, and group membership can change over
time.

Therefore:

- this analysis ranks published artist-credit strings;
- splitting collaborations requires an explicit, documented rule;
- a person's total cannot be inferred safely from text matching alone; and
- title capitalization and remixes should be checked before deduplication.

Inspect possible variants before merging them:

```{r}
hot100 |>
  distinct(artist) |>
  filter(str_detect(str_to_lower(artist), "taylor swift"))
```

## Validate the measures

```{r}
stopifnot(
  sum(artist_weeks_at_number_one$weeks_at_number_one) == nrow(number_ones)
)

stopifnot(
  sum(artist_number_one_songs$number_one_songs) ==
    nrow(distinct(number_ones, artist, song))
)
```

The first check reconciles weekly rows. The second reconciles distinct songs.
Neither proves the source is complete, but each proves the aggregation matches
its stated metric.

## Independent task

Design three different measures of chart success for a decade of your choice.
At least one must use weekly appearances and one must use a distinct count.
Submit:

1. a one-sentence definition of each metric;
2. the code and top-five table for each;
3. one example showing how the rankings differ;
4. a chart whose title and axis match one metric exactly; and
5. a note about collaborations, name variants, or duplicate song titles.
