---
title: "Set Up R, RStudio, and Quarto"
---

## Learning objectives

By the end of this chapter, you should be able to:

- explain the different jobs of R, RStudio, and Quarto;
- create and open an RStudio Project;
- run code in the Console and in a Quarto document;
- install a package once and load it in a working session; and
- render a reproducible HTML page.

## Three tools, three jobs

**R** is the language and calculation engine. **RStudio** is the working
environment where you write code, inspect data, and manage files. **Quarto**
combines prose, code, results, and citations in one reproducible document.

```{mermaid}
flowchart LR
  A["R code"] --> B["R computes"]
  C["Prose and citations"] --> D["Quarto document"]
  B --> D
  D --> E["HTML story, report, or website"]
```

Install R from [CRAN](https://cran.r-project.org/), then install
[RStudio Desktop](https://posit.co/download/rstudio-desktop/). Current RStudio
releases include Quarto. If the **Render** button is visible when you open a
`.qmd` file, Quarto is available.

## Work inside a project

An RStudio Project gives your analysis a stable home. It prevents the most
common beginner problem: code that works only because R is looking in an
unknown folder.

1. Choose **File → New Project → Existing Directory**.
2. Select the folder containing this textbook.
3. Confirm that the Files pane shows `_quarto.yml`, `index.qmd`, and `data/`.

The project folder is now the starting point for relative paths. For example:

```{r}
file.exists("data/processed/gpfg_equities_latest.csv")
```

The result should be `TRUE`. Do not solve path problems with a long personal
path such as `/Users/name/Desktop/course/...`; that path will fail on another
computer.

## Console, script, and Quarto document

Use the **Console** for quick experiments. Use an `.R` script for reusable data
preparation. Use a `.qmd` document when the explanation, code, output, and
sources belong together.

Try this in the Console:

```r
2 + 2
```

Then create a Quarto document with **File → New File → Quarto Document** and
place this code in a code chunk:

````markdown
```{r}
headline <- "A reproducible result"
headline
```
````

Click **Render**. Quarto runs the document from the beginning in a fresh
session. A document that renders is stronger evidence of reproducibility than
code that only works interactively.

## Packages

Packages extend R. Installation and loading are different actions.

Run installation once in the Console:

```r
install.packages("tidyverse")
```

Load the package in every new R session or at the top of a chapter:

```{r}
library(tidyverse)
```

Do not put `install.packages()` in a rendered chapter. Rendering should not
change a reader's computer or pause to ask which download server to use.

For this book, install the main packages with:

```r
install.packages(c(
  "tidyverse", "readxl", "scales", "ggrepel", "sf",
  "rnaturalearth", "rnaturalearthdata", "tidytext"
))
```

## A five-minute system check

Create `setup-check.qmd` and render the following:

````markdown
---
title: "Setup check"
format: html
---

```{r}
library(tidyverse)

check <- tibble(
  tool = c("R", "tidyverse", "course data"),
  ready = c(
    TRUE,
    packageVersion("dplyr") > "1.0.0",
    file.exists("data/processed/gpfg_equities_latest.csv")
  )
)

check
```
````

All three values in `ready` should be `TRUE`.

## Common setup problems

| Symptom | Likely cause | First check |
|---|---|---|
| `there is no package called ...` | The package is not installed | Run `install.packages()` in the Console |
| `could not find function ...` | The package is installed but not loaded | Run `library(...)` near the top |
| `cannot open file ...` | The project or relative path is wrong | Run `getwd()` and `file.exists()` |
| Code works line by line but Render fails | An object was created only in the Console | Restart R and render from the beginning |
| A package cannot be installed | R or system software is too old | Record the exact error before changing anything |

## Independent task

Create a small Quarto page that:

1. loads `tidyverse`;
2. creates a three-row tibble about three possible story ideas;
3. filters the table to one idea; and
4. renders successfully after **Session → Restart R**.

At the bottom, add a short note explaining why a successful fresh render is a
useful newsroom check.
