---
title: "Map Geographic Patterns Carefully"
---

## Learning objectives

By the end of this chapter, you should be able to:

- aggregate holding records to a geographic grain;
- preserve map geometry during a join;
- audit unmatched place names;
- choose and disclose a color transformation; and
- explain what an investment map does and does not show.

## Newsroom question

> Where was the reported market value of NBIM's latest year-end equity holdings
> concentrated?

The source contains holdings, but a country map needs one value per country.
Aggregate before joining to boundaries.

```{r}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)

gpfg <- read_csv(
  "data/processed/gpfg_equities_latest.csv",
  show_col_types = FALSE
)

latest_year <- unique(gpfg$year)

country_summary <- gpfg |>
  summarise(
    market_value_usd = sum(market_value_usd),
    holding_records = n(),
    .by = country
  )
```

One row in `country_summary` represents one NBIM investment market.

## Get boundary data

```{r}
world <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)

class(world)
```

An `sf` object contains ordinary variables plus a geometry column. Loading
`sf` explicitly ensures that spatial methods are available during a fresh
Quarto render.

## Harmonise names explicitly

The current source and Natural Earth disagree on three labels. Make the decisions
visible instead of hiding them inside a spreadsheet:

```{r}
country_summary <- country_summary |>
  mutate(
    map_name = recode(
      country,
      "Russia" = "Russian Federation",
      "South Korea" = "Republic of Korea",
      "Türkiye" = "Turkey"
    )
  )
```

Audit the proposed join:

```{r}
unmatched_holdings <- country_summary |>
  anti_join(
    world |>
      st_drop_geometry() |>
      distinct(name_long),
    by = c("map_name" = "name_long")
  )

unmatched_holdings
```

The result should contain no rows. If a later source year produces an unmatched
country, stop and investigate it before drawing the map.

Also check for boundary names that have data only under a different definition:

```{r}
world |>
  st_drop_geometry() |>
  distinct(name_long) |>
  semi_join(country_summary, by = c("name_long" = "map_name")) |>
  nrow()
```

## Preserve all polygons

Use the map as the left table so countries without a reported holding remain
visible:

```{r}
world_holdings <- world |>
  filter(name_long != "Antarctica") |>
  left_join(
    country_summary,
    by = c("name_long" = "map_name"),
    relationship = "one-to-one"
  )

stopifnot(!any(st_is_empty(world_holdings)))
stopifnot(
  sum(!is.na(world_holdings$market_value_usd)) == nrow(country_summary)
)
```

An inner join would silently remove all countries without a match and could
make the world appear smaller.

## Draw the map

```{r}
ggplot(world_holdings) +
  geom_sf(
    aes(fill = market_value_usd),
    color = "white",
    linewidth = 0.08
  ) +
  scale_fill_viridis_c(
    option = "C",
    trans = "sqrt",
    breaks = c(10e9, 100e9, 500e9),
    labels = label_dollar(scale = 1e-9, suffix = "B"),
    na.value = "#E5E7EB"
  ) +
  coord_sf(expand = FALSE) +
  labs(
    title = "NBIM's reported equity value was concentrated in a limited set of markets",
    subtitle = paste0("Year-end market value by investment market, ", latest_year, "; square-root color scale"),
    fill = "Market value\n(USD)",
    caption = str_wrap(
      paste(
        "Gray indicates no matched holding in the source, not necessarily zero economic exposure.",
        "Sources: Norges Bank Investment Management and Natural Earth."
      ),
      width = 105
    )
  ) +
  guides(
    fill = guide_colorbar(
      title.position = "top",
      barwidth = grid::unit(11, "cm"),
      barheight = grid::unit(0.45, "cm")
    )
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0, size = 8.5),
    plot.caption.position = "plot"
  )
```

The square-root transformation prevents the largest market from visually
flattening all other differences. Because color distance is transformed, the
subtitle discloses it. A ranked bar chart remains better for exact comparison.

## Interpret the geography cautiously

This map shows where NBIM classifies the investment market of a holding. It does
not necessarily show:

- where a company earns revenue;
- where its workers or customers are located;
- its legal place of incorporation;
- the destination of a purchase during the reporting year; or
- Norway's political relationship with a country.

Compare `country` with `incorporation_country` to see why a single corporate
geography is often inadequate.

Boundary datasets also embed naming and political choices. Record the boundary
source and version, preserve the source's original country label, and explain
any harmonisation that affects sensitive places.

## Independent task

Create a map for another year in the teaching panel. Your submission must
include:

1. the country-level aggregation;
2. an `anti_join()` table of unmatched source places;
3. a row-count or match-count assertion;
4. a ranked table that verifies the five strongest map values; and
5. a caption explaining the color scale and the meaning of gray areas.
