---
title: "Mapping"
---

## Learning objectives

By the end of this chapter, you should be able to:

- decide whether geography is important to a reporting question;
- prepare one value per place;
- match place names to a boundary table;
- check unmatched names before joining;
- build a choropleth map step by step; and
- explain what the map's colors and missing areas mean.

## When is a map useful?

A ranked bar chart is usually better when the main task is comparing exact
country values. A map is useful when location and geographic pattern matter.

Our question is:

> Where was the reported market value of the fund's equity holdings
> concentrated in 2025?

The question needs `country` and `market_value_usd`. To draw country shapes,
we also need longitude, latitude, and a polygon group from a boundary table.
The analysis and boundary data therefore have to be joined.

To keep this lesson focused on the tidyverse, the course provides a prepared
CSV of world boundary coordinates. The coordinates come from Natural Earth;
their reproducible preparation is recorded in
`scripts/prepare_map_boundaries.R`.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `10-mapping.Rmd`. Reuse
`data/gpfg_5_years.csv`, then download
[`world_boundaries.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/world_boundaries.csv)
and save it in `data/`.
:::

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_5_years.csv")
world <- read_csv("data/world_boundaries.csv")

gpfg_latest <- gpfg |>
  filter(year == 2025)
```

## Step 1: prepare one value per place

The holdings file contains many companies per country. A choropleth map needs
one value per geographic area, so reuse the summary pattern from Chapter 5:

```{r}
country_summary <- gpfg_latest |>
  group_by(country) |>
  summarise(market_value_usd = sum(market_value_usd)) |>
  mutate(market_value_usd_billions = market_value_usd / 1000000000)

country_summary |>
  arrange(desc(market_value_usd_billions)) |>
  slice_head(n = 5)
```

One row now represents one investment market. Inspecting the five largest
values also gives us a table against which to check the finished map.

## Step 2: make place names match

The holdings and boundary files do not always spell a place in the same way.
Keep the original `country` column and create a separate matching name:

```{r}
country_summary <- country_summary |>
  mutate(
    map_name = recode(
      country,
      "Russia" = "Russian Federation",
      "South Korea" = "Republic of Korea",
      "Türkiye" = "Turkey"
    )
  )
```

`recode()` replaces only the listed values. A separate `map_name` column makes
the changes visible and preserves the publisher's original labels.

## Step 3: check matches before joining

Chapter 9 introduced `anti_join()` as a diagnostic. Use it again to find
investment markets without a matching boundary name:

```{r}
map_names <- world |>
  distinct(name_long)

country_summary |>
  anti_join(map_names, by = c("map_name" = "name_long"))
```

An empty result means every source market found a boundary. If names appear,
investigate them and add only justified corrections before continuing.

## Step 4: join values to coordinates

Keep the world boundaries on the left so that countries without a reported
value still appear on the map:

```{r}
world_holdings <- world |>
  left_join(
    country_summary,
    by = c("name_long" = "map_name")
  )
```

Each country has many coordinate rows describing its outline. The join repeats
the country's market value across those coordinate rows so `ggplot2` can fill
the complete shape. This is expected here; it would not be safe to sum those
repeated values after the join.

## Step 5: make the simplest map

First decide what each aesthetic represents:

| Aesthetic | Column | Meaning |
|---|---|---|
| x | `longitude` | Horizontal coordinate |
| y | `latitude` | Vertical coordinate |
| group | `polygon_group` | Points belonging to the same shape |
| fill | `market_value_usd_billions` | Reported value represented by color |

Now draw the polygons with only the essential mappings:

```{r}
ggplot(
  world_holdings,
  aes(
    x = longitude,
    y = latitude,
    group = polygon_group,
    fill = market_value_usd_billions
  )
) +
  geom_polygon()
```

This map works, but the coordinate ratio, missing-value color, and legend need
attention before publication.

## Step 6: correct the map shape and borders

Add a map-friendly coordinate system and quiet borders:

```{r}
ggplot(
  world_holdings,
  aes(
    x = longitude,
    y = latitude,
    group = polygon_group,
    fill = market_value_usd_billions
  )
) +
  geom_polygon(color = "white", linewidth = 0.1) +
  coord_quickmap(expand = FALSE)
```

`coord_quickmap()` keeps the world from being stretched incorrectly. The thin
white borders separate neighboring shapes without dominating the data.

## Step 7: choose a meaningful color scale

Market value is a continuous quantity, so use one light-to-dark sequential
color scale. Darker blue will mean a larger value:

```{r}
ggplot(
  world_holdings,
  aes(
    x = longitude,
    y = latitude,
    group = polygon_group,
    fill = market_value_usd_billions
  )
) +
  geom_polygon(color = "white", linewidth = 0.1) +
  coord_quickmap(expand = FALSE) +
  scale_fill_gradient(
    low = "#DCEAF4",
    high = "#006D77",
    trans = "sqrt",
    na.value = "#E6E8EB"
  )
```

The square-root transformation makes differences among smaller values easier
to see without changing the values printed on the legend. Light gray means no
matched value in this table; it does not automatically mean zero.

## Step 8: add the story and simplify the design

Finally, add the title, legend label, source, and a minimal map theme:

```{r}
ggplot(
  world_holdings,
  aes(
    x = longitude,
    y = latitude,
    group = polygon_group,
    fill = market_value_usd_billions
  )
) +
  geom_polygon(color = "white", linewidth = 0.1) +
  coord_quickmap(expand = FALSE) +
  scale_fill_gradient(
    low = "#DCEAF4",
    high = "#006D77",
    trans = "sqrt",
    na.value = "#E6E8EB"
  ) +
  labs(
    title = "Reported equity value was concentrated in a limited set of markets",
    subtitle = "Year-end 2025 market value by NBIM investment market",
    fill = "USD billions",
    caption = "Sources: Norges Bank Investment Management and Natural Earth\nGray indicates no matched value"
  ) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", color = "#16323F"),
    plot.subtitle = element_text(color = "#52606D"),
    plot.caption = element_text(color = "#6B7280", hjust = 0),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.margin = margin(12, 18, 12, 18)
  )
```

The final version is the same map as the basic one. Each layer solves one
communication problem: shape, separation, color meaning, labels, or visual
clutter.

## Interpret the map carefully

The map uses NBIM's investment-market column. It does not necessarily show a
company's place of incorporation, where its workers are located, or where its
revenue is earned. Boundary files also contain naming and political choices,
so record the boundary source and explain consequential name changes.

The map is good for showing a broad geographic pattern. Use the ranked table
when readers need exact values or close comparisons between countries.

## Practice

Create the same map for another year in `gpfg_5_years.csv`. Submit:

1. the filtered annual data;
2. the country summary;
3. the unmatched-name result;
4. the basic and final maps;
5. a ranked table of the five largest values; and
6. a caption explaining what the colors and gray areas mean.

## Takeaways

| Function | What it does |
|---|---|
| `recode()` | Replaces selected place names explicitly |
| `anti_join()` | Finds source places without a boundary match |
| `left_join()` | Adds the data values to boundary coordinates |
| `geom_polygon()` | Draws filled shapes from coordinate rows |
| `coord_quickmap()` | Uses a map-friendly aspect ratio |
| `scale_fill_gradient()` | Adds a continuous light-to-dark color scale |
| `theme_void()` | Removes axes and background elements from a map |

::: {.callout-tip title="Want more control?"}
The official [ggplot2 reference](https://ggplot2.tidyverse.org/reference/)
documents more options for polygons, coordinates, themes, and color scales.
Start with a clear geographic question and a checked join before exploring
additional styling.
:::
