# Mapping Global Life Expectancy {#sec-lifexmap}

## Set up
```{r}
library(tidyverse)
library(rnaturalearth) # For world map data

load("out/expectancy_merged.RData")
```

## Overview of Life Expectancy

Here let's start by looking at the summary statistics of worldwide life expectancy in the most recent year available (2022).

```{r}
# Summary statistics of worldwide life expectancy in 2022
df_merge |>
  filter(year == 2022) |>
  summarise(mean_life_expectancy = mean(life_expectancy),
            median_life_expectancy = median(life_expectancy),
            min_life_expectancy = min(life_expectancy),
            max_life_expectancy = max(life_expectancy))
```

### Top 10 Countries/Regions with Highest Life Expectancy

```{r}
# Top 10 countries with the highest life expectancy in 2022
df_merge |>
  filter(year == 2022) |>
  arrange(desc(life_expectancy)) |>
  head(10)
```

## Data Visualization

### Bar Plot: Life Expectancy in 2022 by 10 Major Countries

```{r}
# List of 10 major countries (with highest GDP)
country_list <- c("United States", "China", "Japan", "Germany", "India", 
                  "United Kingdom", "France", "Brazil", "Italy", "Canada")
```


```{r}
df_merge |>
  filter(year == 2022) |>
  filter(country %in% country_list) |>
  ggplot(aes(x = reorder(country, life_expectancy), y = life_expectancy)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Life Expectancy in 2022 by 10 Major Countries",
       x = "Country",
       y = "Life Expectancy at Birth",
       fill = "Country",
       caption = "Source: World Bank | Author: Bin Chen") +
  geom_text(aes(label = round(life_expectancy, 1)), hjust = -0.1, size = 3) +
  theme_minimal()
```

### Line Plot: Life expectancy trend in China

```{r}
# Life expectancy trend in China
df_merge |>
  filter(country == "China") |>
  ggplot(aes(x = year, y = life_expectancy)) +
  geom_line(color = "steelblue") +
  labs(title = "Life Expectancy Trend in China",
       x = "Year",
       y = "Life Expectancy at Birth",
       caption = "Source: World Bank | Author: Bin Chen") +
  theme_minimal()
```

### Multiple Lines

```{r}
# Life expectancy trend in the 10 major countries
df_merge |>
  filter(country %in% country_list) |>
  ggplot(aes(x = year, y = life_expectancy, color = country)) +
  geom_line() +
  labs(title = "Life Expectancy Trend in 10 Major Countries",
       x = " ",
       y = " ",
       caption = "Source: World Bank | Author: Bin Chen") +
  theme_minimal() +
  facet_wrap(~country, scales = "free_y") + 
  theme(legend.position = "none")
```

### Map Plot: Life Expectancy by Country in 2022

```{r}
# Map of life expectancy by country in 2022
df_merge |>
  filter(year == 2022) |>
  ggplot(aes()) + # <1> 
  geom_sf(aes(geometry = geometry, fill = life_expectancy)) + # <2>
  scale_fill_gradient(low = "white", 
                      high = "#3182bd",
                      na.value = "grey90", 
                      name = "Life Expectancy", 
                      guide = guide_colorbar(
                        barwidth = 10,
                        barheight = 0.5,
                        title.position = "top",
                        title.hjust = 0.5
                      )) + # <3>
  labs(title = "Life Expectancy at Birth by Country in 2022",
       fill = "Life Expectancy at Birth",
       caption = "Source: World Bank | By: Bin Chen") + # <4>
  theme_void() + # <5>
  theme(legend.position = "bottom",
        legend.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 11)) # <6>
```
1. Set up the plot.
2. Create a spatial feature plot using the world map geometry and fill color based on life expectancy.
3. Set the color gradient for life expectancy.
4. Add titles and captions.
5. Remove the background.
6. Adjust the legend position and title size.
7. Adjust the plot title position and size.




