---
title: "From Analysis to a Data Story"
---

## Learning objectives

By the end of this chapter, you should be able to:

- turn exploratory findings into a focused story hypothesis;
- build an evidence table before drafting;
- separate description, interpretation, and explanation;
- verify a financial claim against the source; and
- write a transparent methods and limitations note.

## The reporting frame

We now combine the skills from the core workflow. The possible story is not
“where Norway sent money in the latest year.” The data do not contain transactions. A
defensible frame is:

> How concentrated were NBIM's reported equity holdings at the latest year-end,
> and how did their nominal market value change over the previous decade?

```{r}
library(tidyverse)
library(scales)

gpfg_latest <- read_csv(
  "data/processed/gpfg_equities_latest.csv",
  show_col_types = FALSE
)

gpfg_panel <- readRDS(
  "data/processed/gpfg_equities_last_10_years.rds"
)

latest_year <- unique(gpfg_latest$year)
panel_years <- sort(unique(gpfg_panel$year))
start_year <- min(panel_years)
end_year <- max(panel_years)

stopifnot(length(latest_year) == 1, latest_year == end_year)

total_latest_nok <- sum(gpfg_latest$market_value_nok)

country_evidence <- gpfg_latest |>
  summarise(market_value_nok = sum(market_value_nok), .by = country) |>
  mutate(share = market_value_nok / sum(market_value_nok)) |>
  arrange(desc(market_value_nok))

industry_evidence <- gpfg_latest |>
  summarise(market_value_nok = sum(market_value_nok), .by = industry) |>
  mutate(share = market_value_nok / sum(market_value_nok)) |>
  arrange(desc(market_value_nok))

largest_holding <- gpfg_latest |>
  arrange(desc(market_value_nok)) |>
  slice_head(n = 1)

annual_evidence <- gpfg_panel |>
  summarise(market_value_nok = sum(market_value_nok), .by = year) |>
  arrange(year)

decade_change <- annual_evidence$market_value_nok[annual_evidence$year == end_year] /
  annual_evidence$market_value_nok[annual_evidence$year == start_year] - 1
```

## Build an evidence table

Draft from checked evidence, not from memory or a chart image.

```{r}
evidence_table <- tibble(
  proposed_claim = c(
    paste0("Total reported equity market value in ", latest_year),
    "Largest investment market and its share",
    "Largest industry and its share",
    "Largest individual holding",
    paste0("Change in nominal NOK value from ", start_year, " to ", end_year)
  ),
  evidence = c(
    paste0(round(total_latest_nok / 1e12, 2), " trillion NOK"),
    paste0(
      country_evidence$country[1],
      ": ",
      percent(country_evidence$share[1], accuracy = 0.1)
    ),
    paste0(
      industry_evidence$industry[1],
      ": ",
      percent(industry_evidence$share[1], accuracy = 0.1)
    ),
    largest_holding$company,
    percent(decade_change, accuracy = 1)
  ),
  verification = c(
    "Reconcile company rows with country and industry totals",
    "Check denominator equals the full latest-year equity total",
    "Check 2021 classification definitions",
    "Verify the company row in the official Excel file",
    "Confirm both endpoints and describe values as nominal snapshots"
  )
)

evidence_table
```

An evidence table helps an editor see which sentence rests on which
calculation and what still needs reporting.

## Three defensible findings

The computed evidence supports statements such as:

1. The `r latest_year` file reports approximately
   `r round(total_latest_nok / 1e12, 2)` trillion NOK in listed-equity holdings.
2. `r country_evidence$country[1]` was the largest investment market,
   accounting for `r percent(country_evidence$share[1], accuracy = 0.1)` of the
   reported equity market value.
3. The nominal NOK value at the end of `r end_year` was
   `r percent(decade_change, accuracy = 1)` higher than at the end of `r start_year`.

Each sentence identifies the reporting period and measure. None says that the
fund transferred that amount during the reporting year or earned that percentage as a return.

## Separate evidence from explanation

The holdings files can describe:

- the year-end value of a named holding;
- the number of reported holdings;
- country, region, and industry totals;
- ownership and voting stakes; and
- changes between snapshots.

They cannot by themselves explain why a value changed. Possible explanations
require other evidence, such as:

- NBIM annual reports and mandate documents;
- transaction or flow data;
- security-price and exchange-rate changes;
- company mergers, spin-offs, and name changes;
- interviews with NBIM or independent specialists; and
- documentation of classification changes.

A story can report a pattern first and then test competing explanations. It
should not promote a plausible explanation to a fact without evidence.

## Verify representative records

Before drafting:

1. check at least three top-ranked company rows in the official Excel file;
2. recalculate the main total through a second grouping route;
3. confirm every percentage denominator;
4. inspect unmatched categories used in joins or maps;
5. compare the latest values with NBIM's own published summary where the
   definitions align; and
6. record the source date, download date, and script version.

For example, country and industry totals should reconcile with the company-row
total:

```{r}
stopifnot(isTRUE(all.equal(
  sum(country_evidence$market_value_nok),
  total_latest_nok
)))

stopifnot(isTRUE(all.equal(
  sum(industry_evidence$market_value_nok),
  total_latest_nok
)))
```

## A methods note readers can understand

> We downloaded NBIM's official year-end listed-equity holdings for `r start_year`–`r end_year`
> and analyzed the standardized files in R. Each row represents one reported
> holding at one year-end. We summed nominal market values in Norwegian kroner
> by investment market and industry and calculated shares using the full annual
> equity total. Snapshot changes combine prices, currencies, purchases, sales,
> and changes in portfolio composition; they are not investment flows or
> returns. We did not match companies longitudinally because published names
> are not stable identifiers.

## Story structure

A concise data story could use this order:

1. **Lead:** the strongest verified concentration finding.
2. **Nut paragraph:** what the data cover and why the pattern matters.
3. **Evidence:** a ranked chart and exact values.
4. **Context:** the ten-year nominal trend.
5. **Reporting:** expert or institutional explanations tested against evidence.
6. **Limitations:** snapshot, classification, currency, and identity caveats.
7. **Methods/source box:** links, dates, code, and definitions.

## Independent newsroom assignment

Develop a 600–800 word story pitch based on the holdings data. Submit:

- a one-sentence hypothesis;
- a five-row evidence table;
- one publication chart and its verification table;
- two external sources you would interview or consult;
- a methods paragraph;
- three claims you rejected because the data could not support them; and
- a headline that avoids the words *received*, *spent*, *profit*, and *return*
  unless you add separate evidence for those concepts.
