---
title: "Olympics Data"
---

::: {.callout-important title="The new challenge: medallists are not medals"}
One row associates one athlete with one medal result. A team medal therefore
appears once for every team member, and one athlete can appear in several medal
rows. We must distinguish athlete-medal rows, unique medallists, and official
event-level medals before making a ranking.
:::

## Learning objectives

By the end of this case, you should be able to:

- distinguish athlete-medal rows from unique medallists;
- use an athlete identifier rather than a displayed name;
- distinguish team-member rows from event-level medals;
- create a simple bar chart of the result; and
- recognize that different sources use different country codes.

## Set up the case

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `10-olympics.Rmd`. Download
[`medallists.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/medallists.csv)
and save it in `data/`.
:::

```{r}
library(tidyverse)

medallists <- read_csv("data/medallists.csv")

glimpse(medallists)
```

One row represents one athlete associated with one medal result. An athlete
with several medals appears several times. In a team event, one official medal
creates a row for every team member.

These are therefore different measures:

| Measure | Meaning |
|---|---|
| Athlete-medal rows | Athlete appearances in medal results |
| Unique medallists | People with at least one medal row |
| Official medals | Event-level awards, which are not counted directly by athlete rows |

## Strategy 1: separate rows from people

Begin by comparing the number of rows with the number of unique athlete
identifiers:

```{r}
medallists |>
  summarise(
    athlete_medal_rows = n(),
    unique_medallists = n_distinct(code_athlete)
  )
```

The two totals answer different questions. `n()` counts records, while
`n_distinct(code_athlete)` counts people. The identifier is safer than the
displayed name because names can be repeated or formatted inconsistently.

To find athletes associated with the most medal results, keep the identifier
and readable name together:

```{r}
athlete_medals <- medallists |>
  count(code_athlete, name, country) |>
  arrange(desc(n))

athlete_medals |>
  slice_head(n = 10)
```

The careful label is “athlete-medal rows,” not simply “medals.”

## Strategy 2: expose team-event duplication

Count rows within each event, country, and medal type. The largest counts make
team events visible:

```{r}
medallists |>
  count(discipline, event, country, medal_type) |>
  arrange(desc(n)) |>
  slice_head(n = 10)
```

For a football or hockey result, `n` is close to the number of listed team
members. Those rows describe medallists, but the team received one official
medal for that event.

## Strategy 3: create an event-level medal table

To approximate official medal awards, keep each event, country, and medal type
once. `event_type` is retained because similarly named events can have
different forms:

```{r}
official_medals <- medallists |>
  distinct(
    discipline,
    event,
    event_type,
    country_code,
    country,
    medal_type
  )

official_medals |>
  summarise(number_of_official_medals = n())
```

This changes the unit from an athlete associated with an award to one
country-medal result in one event. For publication, compare the resulting
total with the official medal source and investigate any ties, mixed teams, or
missing event identifiers.

## Strategy 4: compare country measures

First count athlete records and unique people within each country:

```{r}
country_medallists <- medallists |>
  group_by(country_code, country) |>
  summarise(
    unique_medallists = n_distinct(code_athlete),
    athlete_medal_rows = n()
  )

country_official_medals <- official_medals |>
  count(country_code, country, name = "official_medals")

country_measures <- country_medallists |>
  left_join(
    country_official_medals,
    by = c("country_code", "country")
  ) |>
  arrange(desc(official_medals))

country_measures |>
  slice_head(n = 10)
```

The three columns should not be used interchangeably. A country with many
large successful teams can have far more athlete-medal rows than official
medals.

## Visualize the chosen measure

```{r}
country_measures |>
  slice_head(n = 10) |>
  ggplot(
    aes(
      x = official_medals,
      y = fct_reorder(country, official_medals)
    )
  ) +
  geom_col() +
  labs(
    title = "Countries winning the most official medals at Paris 2024",
    x = "Official event-level medals",
    y = NULL,
    caption = "Source: supplied Paris 2024 medallists data"
  ) +
  theme_minimal()
```

If we charted `unique_medallists` instead, both the title and axis would need to
say medallists. The wording follows the unit actually counted.

## Country codes are not universal

The Olympics file uses International Olympic Committee codes. Many map sources
use ISO-style codes instead. For example, Germany can appear as `GER` in one
system and `DEU` in another.

Before joining to a map:

1. identify the coding system used by each source;
2. inspect unmatched countries with `anti_join()`;
3. create a documented lookup table for differences; and
4. do not force non-country delegations onto a country polygon.

The Refugee Olympic Team and Individual Neutral Athletes are important
examples: they should not automatically be assigned to one national map area.

## Practice

Choose one discipline and calculate:

1. athlete-medal rows;
2. unique medallists; and
3. official event-level medals.

Find one team event and show how many athlete rows represent one medal result.
Then create a top-five table and chart for one measure, using the exact unit in
the title.

## Takeaways

| Function or idea | What it does in this case |
|---|---|
| `count()` | Counts athlete-medal rows at the selected grouping |
| `n_distinct()` | Counts different athlete identifiers |
| `distinct()` | Keeps one event-country-medal combination |
| `left_join()` | Places country measures in one comparison table |
| `fct_reorder()` | Orders the countries by the chosen measure |
| Join-key check | Prevents different country-code systems from silently dropping places |

::: {.callout-tip title="Want more control?"}
See the official [dplyr reference](https://dplyr.tidyverse.org/reference/index.html)
for other ways to count and rank, and the
[ggplot2 reference](https://ggplot2.tidyverse.org/reference/) for additional
chart controls.
:::
