---
title: "Unit of Analysis: Paris 2024 Medallists"
---

## Learning objectives

By the end of this case, you should be able to:

- distinguish athlete-medal records, unique medallists, and official medals;
- deduplicate with a defensible identifier;
- avoid a misleading regression on aggregated counts; and
- audit the difference between IOC and ISO country codes.

## Import and inspect

```{r}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

medallists <- read_csv(
  "data/medallists.csv",
  show_col_types = FALSE
)

dim(medallists)
n_distinct(medallists$code_athlete)
```

The file has 2,315 rows but 2,054 unique athlete codes. One row represents one
athlete associated with one medal result. An athlete with several medals has
several rows. In a team event, one official medal produces a row for every team
member.

Therefore, the following are different measures:

| Measure | Calculation | Meaning |
|---|---|---|
| Athlete-medal records | `n()` | Participation in medal results |
| Unique medallists | `n_distinct(code_athlete)` | People with at least one medal record |
| Official medals | Event-level medal awards | Not equal to athlete rows in team events |

## Athletes with the most medal records

```{r}
athlete_medal_records <- medallists |>
  count(
    code_athlete,
    name,
    country,
    gender,
    name = "medal_records",
    sort = TRUE
  )

athlete_medal_records |>
  slice_head(n = 15)
```

The accurate label is “athlete-medal records,” not simply “medals,” because the
data represent athletes within medal results.

## Unique medallists by country

```{r}
country_medallists <- medallists |>
  filter(!is.na(country), !is.na(code_athlete)) |>
  summarise(
    unique_medallists = n_distinct(code_athlete),
    athlete_medal_records = n(),
    .by = country
  ) |>
  arrange(desc(unique_medallists))

country_medallists |>
  slice_head(n = 15)
```

This table is useful, but it is not the official medal table. Countries with
large team-event rosters can accumulate many athlete-medal records from one
official medal.

## Analyze age at the athlete grain

Counting medal rows by age overweights multi-medallists. Deduplicate athletes
first:

```{r}
athletes <- medallists |>
  filter(!is.na(code_athlete)) |>
  distinct(code_athlete, .keep_all = TRUE) |>
  mutate(
    age_at_opening = as.numeric(
      as.Date("2024-07-26") - birth_date
    ) / 365.25
  )

athletes |>
  summarise(
    athletes = n(),
    missing_birth_date = sum(is.na(birth_date)),
    median_age = median(age_at_opening, na.rm = TRUE),
    youngest = min(age_at_opening, na.rm = TRUE),
    oldest = max(age_at_opening, na.rm = TRUE)
  )
```

```{r}
athletes |>
  filter(!is.na(age_at_opening)) |>
  ggplot(aes(x = age_at_opening)) +
  geom_histogram(binwidth = 2, fill = "#1769AA", color = "white") +
  labs(
    title = "The age distribution of unique Paris 2024 medallists",
    subtitle = "Each athlete is counted once, regardless of medal records",
    x = "Age at the opening ceremony",
    y = "Unique medallists"
  ) +
  theme_minimal()
```

A regression of “number of medal rows by age” on age would mostly describe the
shape of this distribution. It would not estimate whether getting older causes
an athlete to win more or fewer medals.

## Country codes are not universal

The dataset uses International Olympic Committee codes, while Natural Earth
provides ISO-style geographic identifiers. Many codes differ: `GER` versus
`DEU`, `SUI` versus `CHE`, and `NED` versus `NLD`. A direct code join would
drop many countries.

Use published country names with an explicit concordance for the unmatched
labels:

```{r}
country_medallists <- country_medallists |>
  mutate(
    map_name = recode(
      country,
      "Great Britain" = "United Kingdom",
      "Korea" = "Republic of Korea",
      "Hong Kong, China" = "Hong Kong",
      "Republic of Moldova" = "Moldova",
      "Chinese Taipei" = "Taiwan",
      "Cabo Verde" = "Republic of Cabo Verde",
      "DPR Korea" = "Dem. Rep. Korea",
      "IR Iran" = "Iran",
      "Türkiye" = "Turkey",
      "Czechia" = "Czech Republic"
    )
  ) |>
  filter(!country %in% c("AIN", "EOR"))
```

`AIN` represents Individual Neutral Athletes and `EOR` represents the Refugee
Olympic Team. Assigning either delegation to one country polygon would be a
substantive error, so they remain in the table but are excluded from the map.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

unmatched <- country_medallists |>
  anti_join(
    world |>
      st_drop_geometry() |>
      distinct(name_long),
    by = c("map_name" = "name_long")
  )

unmatched
```

The result should have no rows before the spatial join.

```{r}
world_medallists <- world |>
  filter(name_long != "Antarctica") |>
  left_join(
    country_medallists,
    by = c("name_long" = "map_name"),
    relationship = "one-to-one"
  )

ggplot(world_medallists) +
  geom_sf(aes(fill = unique_medallists), color = "white", linewidth = 0.08) +
  scale_fill_viridis_c(na.value = "#E5E7EB") +
  labs(
    title = "Unique Paris 2024 medallists by represented country",
    fill = "Unique\nmedallists",
    caption = paste(
      "AIN and EOR are not assigned to country polygons.",
      "This is not an official medal-count map."
    )
  ) +
  theme_void() +
  theme(legend.position = "bottom")
```

## Independent task

Choose one discipline and report three different counts:

1. athlete-medal records;
2. unique medallists; and
3. distinct event–medal combinations.

Explain why none should automatically be labelled “the number of official
medals.” Add one chart at the unique-athlete grain and one validation check
using `code_athlete`.
