---
title: "R Basics"
---

## Learning objectives

By the end of this chapter, you should be able to:

- use R for simple calculations;
- create and name objects with `<-` and recognize other assignment symbols;
- combine values with `c()`;
- use a small set of mathematical functions;
- recognize the main types of data used in this book;
- create a small tibble; and
- explain what a pipe does.

## From setup to coding

The previous chapter introduced RStudio, Projects, and R Markdown. We will now
learn the small pieces of R code that appear in every later chapter. Try each
example in a code chunk and look at the result before moving on. The examples
in this chapter are deliberately small and familiar. We will introduce the
course's main dataset in the Data Import chapter.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create a new R Markdown document named
`01-r-basics.Rmd`. Save it beside `00-setup.Rmd`. Keep using the same `data`
and `outputs` folders for the rest of the book.
:::

## Use R as a calculator

R understands ordinary arithmetic:

```{r}
25 + 10
50 / 4
(25 + 35) / 2
```

Parentheses work as they do in a calculator: R evaluates what is inside them
first.

## Create an object

An **object** is a name that stores a value. Use `<-` to assign a value:

```{r}
x <- 10
x
```

Read the first line as “`x` gets 10.” R stores the value but does not print it
until you type the object's name.

Begin with short names that describe the value:

```{r}
score <- 85
test_score <- 90
```

When a name needs more than one word, use `snake_case`, with lowercase words
separated by underscores. For example, `test_score` is easier to read than
`testscore`. R is case-sensitive, so `score` and `Score` would be different
objects.

You may see three assignment styles in R code:

| Code | Meaning | Use in this book |
|---|---|---|
| `x <- 10` | Assign 10 to `x` from the left | Preferred |
| `10 -> x` | Assign 10 to `x` from the right | Valid, but uncommon |
| `x = 10` | Assign 10 to `x` using an equals sign | Valid in many situations, but not used for object assignment here |

We use `<-` because the object name appears first and because it clearly marks
an assignment. The equals sign has another job you will see later: it names an
option inside a function, written as `option = value`. Keeping `<-` for objects
and `=` for named options makes code easier to read. Later, we will use `==` to
ask whether two values are equal; `=` and `==` do not mean the same thing.

::: {.callout-tip title="Assignment shortcut"}
Use **Alt + -** on Windows or **Option + -** on Mac to insert `<-`.
:::

## Add comments

R ignores text after `#`. Comments are notes for yourself or another reader.
They can explain what a value means or why a step is needed.

```{r}
# Age at the beginning of the course
age <- 20
```

Good comments explain a decision. There is little value in writing
`# calculate the mean` immediately above `mean(numbers)` because the code
already says that.

## Combine values into a vector

`c()` combines several values of the same kind into a **vector**:

```{r}
numbers <- c(2, 4, 6, 8)

numbers
```

R can perform the same calculation on every value in a numeric vector:

```{r}
numbers / 2
```

## Use mathematical functions

A **function** is a named instruction that performs a task. Its name is followed
by parentheses containing the information it needs. R provides simple
functions for common calculations:

```{r}
sum(numbers)
mean(numbers)
min(numbers)
max(numbers)
```

| Function | Calculation |
|---|---|
| `sum()` | Adds the values |
| `mean()` | Calculates the average |
| `min()` | Finds the smallest value |
| `max()` | Finds the largest value |

The object `numbers` is the input, or **argument**, given to each function.

In this book, we begin with the arguments needed for the task. Many functions
have additional optional arguments. You can open a function's help page later
by typing its name after `?`, such as `?mean`.

## Recognize common data types

Four types appear frequently in journalism data:

| Type shown by R | Meaning | Example |
|---|---|---|
| `<dbl>` or `<int>` | Numbers | `10`, `3.5` |
| `<chr>` | Character, or text | `"Alex"` |
| `<lgl>` | Logical values | `TRUE`, `FALSE` |
| `<date>` | Calendar dates | `2025-01-01` |

Text needs quotation marks. Logical values `TRUE` and `FALSE` do not.
Missing information is normally represented by `NA`, not by zero.

## Create a tibble

A data table has variables in columns and observations in rows. The tidyverse
uses a modern data frame called a **tibble**.

First load the tidyverse, then create a small table:

```{r}
library(tidyverse)

students <- tibble(
  name = c("Ana", "Ben", "Chen"),
  age = c(20, 21, 22),
  present = c(TRUE, FALSE, TRUE)
)

students
```

One row represents one student. The three columns contain each student's name,
age, and attendance.

### What does “tidy” mean?

**Tidy data** are organized in a consistent way:

- each variable has its own column;
- each observation has its own row; and
- each value has its own cell.

In `students`, `name`, `age`, and `present` are variables, so each has a
column. Each student is one observation, so each student has one row.

“Tidy” describes the arrangement of a table; it is not another file type. A
tibble can contain tidy or untidy data. We will learn how to make untidy tables
tidy in the Combine and Reshape chapter.

::: {.callout-tip title="Learn more about tidy data"}
The official [tidyr introduction to tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html)
provides more examples. The three rules above are enough for now.
:::

## Use pipes

A **pipe** passes the result on its left to the function on its right. Modern R
has a built-in pipe written as `|>`:

```{r}
numbers |>
  sum()
```

Read this as “start with `numbers`, then calculate the sum.” This produces the
same result as `sum(numbers)`.

You may also see an older pipe written as `%>%` in R tutorials and existing
projects:

```{r}
numbers %>%
  sum()
```

For a simple example like this, both pipes do the same job. The main difference
for beginners is where they come from:

| Pipe | Where it comes from | Use in this book |
|---|---|---|
| `|>` | Built into R | Used throughout the book |
| `%>%` | Available after loading the tidyverse | Shown here so you can recognize older code |

Pipes become especially useful when several steps are connected. We will use
`|>` consistently in the later chapters.

## Useful habits when learning R

- **Copy, paste, and tweak:** start from a working example, change one part,
  and run it again.
- **Save frequently:** keep the code that produced an important number or
  chart.
- **Read errors from the bottom:** the final line often names the immediate
  problem.
- **Check spelling and capitalization:** small differences create different
  object and column names.
- **Ask for help with evidence:** include the code, the error, and what you
  expected to happen.

More detailed debugging guidance is available in the Debugging appendix.

## Practice

Create a tibble called `books` with three rows and these columns:

- `title`, containing three book titles;
- `pages`, containing the number of pages in each book; and
- `finished`, containing `TRUE` or `FALSE`.

Print the tibble. Then write one sentence describing what one row represents.

## Takeaways

| Function or symbol | What it does |
|---|---|
| `<-` | Assigns a value to an object name |
| `->` | Assigns in the opposite direction; valid but uncommon |
| `=` | Can assign a value, but is reserved here for named function arguments |
| `#` | Adds a comment that R does not run |
| `c()` | Combines values into a vector |
| `sum()` | Calculates a total |
| `mean()` | Calculates an average |
| `min()` | Finds the smallest value |
| `max()` | Finds the largest value |
| `tibble()` | Creates a tidyverse data table |
| `|>` | Passes an object to the next function |
| `%>%` | An older tidyverse pipe that appears in existing R code |

::: {.callout-tip title="Want more control?"}
The official [`tibble()` reference](https://tibble.tidyverse.org/reference/tibble.html)
and [tidyverse pipe guide](https://style.tidyverse.org/pipes.html) provide more
examples. The simple forms in this chapter are enough for the next lessons.
:::
