---
title: "R Basics for Journalists"
---

## Learning objectives

By the end of this chapter, you should be able to:

- create and inspect R objects;
- distinguish numbers, text, logical values, dates, and missing values;
- use comparison and logical operators;
- call functions and read an error message;
- work with vectors and data frames; and
- translate a short pipeline into plain language.

## Start with a question

Code is easier to understand when it answers a concrete question. Imagine a
reporter has the number of records in four annual data releases:

```{r}
records <- c(8985, 9106, 9157, 9202)
years <- c(2016, 2017, 2018, 2019)
```

`<-` assigns a value to a name. `c()` combines values into a vector. The
objects remain available until the R session ends or they are removed.

```{r}
length(records)
mean(records)
records[1]
```

Read each expression aloud: “the length of `records`,” “the mean of
`records`,” and “the first value in `records`.”

## Names and assignment

Use short, descriptive names written in `snake_case`:

```{r}
latest_year <- 2025
source_name <- "Norges Bank Investment Management"
data_checked <- TRUE
```

R is case-sensitive. `latest_year` and `Latest_year` would be different
objects. Avoid names such as `data`, `mean`, or `filter` when a more specific
name is available, because those names already describe common objects or
functions.

## Data types

The type of a value determines which operations make sense.

```{r}
typeof(1500)                 # double: a number
typeof("1500")               # character: text
typeof(TRUE)                 # logical
typeof(as.Date("2025-12-31")) # double internally, with a Date class
```

Quotation marks turn a value into text. R cannot safely add `10` to `"1500"`
until the text is converted and checked.

```{r}
as.numeric("1500") + 10
```

Do not convert blindly. Values such as `"not reported"` carry information
about the source and should usually become an explicit missing value only after
you document the rule.

## Comparisons and logical conditions

Comparisons return `TRUE` or `FALSE`.

```{r}
latest_year == 2025
latest_year != 2024
records > 9100
```

Use `==` to compare values. A single `=` is normally used to name a function
argument.

Logical operators combine conditions:

| Operator | Meaning | Example |
|---|---|---|
| `&` | and | `year >= 2016 & year <= 2025` |
| `|` | or | `country == "China" | country == "Japan"` |
| `!` | not | `!is.na(value)` |
| `%in%` | belongs to a set | `country %in% c("China", "Japan")` |

```{r}
years[years >= 2017 & years <= 2019]
```

## Missing values are not zero

R represents a missing value as `NA`. Missing is not the same as zero, an empty
string, or “no.”

```{r}
ownership <- c(0.8, 1.2, NA, 0.5)

mean(ownership)
mean(ownership, na.rm = TRUE)
sum(is.na(ownership))
```

The first mean is `NA` because R refuses to pretend the missing observation is
known. `na.rm = TRUE` tells R to calculate from available values. That may be
appropriate, but the reporter still needs to ask why the value is missing and
whether the remaining cases are representative.

::: {.callout-warning title="A newsroom rule"}
Never replace `NA` with zero just to make code run. First determine whether the
source means “none,” “not applicable,” “not collected,” or “not reported.”
:::

## Functions and arguments

A function takes inputs and returns a result.

```{r}
round(mean(records), digits = 1)
```

Here `records` is the main input and `digits = 1` is a named argument. Use
`?round` in the Console to open the help page.

When a function does not behave as expected, inspect the input before changing
the code:

```{r}
class(records)
length(records)
summary(records)
```

## Data frames

A data frame is a rectangular collection of variables. A tibble is the modern
tidyverse version.

```{r}
library(tidyverse)

annual <- tibble(
  year = years,
  holding_records = records,
  audited = c(TRUE, TRUE, TRUE, TRUE)
)

annual
```

Each column should contain one variable, each row should contain one
observation, and each cell should contain one value. This “tidy” structure is a
goal, not an assumption: real journalism datasets often need to be reshaped.

Useful inspection functions include:

```{r}
dim(annual)
names(annual)
glimpse(annual)
summary(annual)
```

## Pipes make a sequence readable

The base R pipe `|>` passes the result on the left into the function on the
right.

```{r}
annual |>
  filter(year >= 2017) |>
  arrange(desc(holding_records))
```

Translate the pipeline before running it:

> Start with `annual`, keep years from 2017 onward, then order the remaining
> rows from the largest record count to the smallest.

One line should usually perform one meaningful step. This makes the analysis
easier to check and explain to an editor.

## Checkpoint exercise

Create this vector:

```{r}
values <- c(120, 150, NA, 90, 200)
```

Answer the following with R:

1. How many values are recorded?
2. How many are missing?
3. What is the median of the available values?
4. Which available values are at least 150?
5. Why would replacing the missing value with zero change the story?

<details>
<summary>Show one possible solution</summary>

```{r}
length(values)
sum(is.na(values))
median(values, na.rm = TRUE)
values[!is.na(values) & values >= 150]
```

Replacing `NA` with zero would assert that the true value is zero. It would
lower the median and turn an unknown observation into a factual claim.

</details>

## Independent task

Build a tibble with five possible data sources for a local story. Include:

- `source` as text;
- `year` as a number;
- `downloaded` as a logical value; and
- `notes` with at least one `NA`.

Then write a pipeline that keeps downloaded sources from 2024 onward. Under the
code, explain the pipeline in one sentence and state what `NA` means in your
`notes` column.
