---
title: "Combine Years, Join Tables, and Reshape Data"
---

## Learning objectives

By the end of this chapter, you should be able to:

- distinguish stacking rows from joining columns;
- combine annual files while preserving the report year and source;
- use join keys and relationship checks deliberately;
- audit matched and unmatched records; and
- pivot between long and wide forms without losing the unit of analysis.

## Newsroom question

> How did the geographic distribution of reported equity holdings change from
> the first to the latest year in the rolling teaching panel?

This question requires more than one snapshot. The prepared panel contains ten
year-end files:

```{r}
library(tidyverse)

gpfg_panel <- readRDS(
  "data/processed/gpfg_equities_last_10_years.rds"
)

panel_years <- sort(unique(gpfg_panel$year))
start_year <- min(panel_years)
end_year <- max(panel_years)

stopifnot(length(panel_years) == 10)
stopifnot(all(diff(panel_years) == 1))

gpfg_panel |>
  count(year)
```

The input grain is one equity holding at one year-end. The same published name
may appear in several years, so the pair `year + company`—not `company`
alone—is the apparent key for this teaching period.

```{r}
duplicate_keys <- gpfg_panel |>
  count(year, company) |>
  filter(n > 1)

duplicate_keys
```

## Stack observations with `bind_rows()`

When tables have the same variables and represent different observations,
stack their rows:

```{r}
gpfg_start <- gpfg_panel |>
  filter(year == start_year)

gpfg_latest <- gpfg_panel |>
  filter(year == end_year)

two_years <- bind_rows(gpfg_start, gpfg_latest)

two_years |>
  count(year)
```

The report year must exist before stacking. Otherwise, identical-looking rows
from different snapshots cannot be distinguished.

The complete preparation script reads all available official annual Excel files,
extracts the year from each filename, standardizes names and types, and then
uses `map_dfr()` to combine them. The key pattern is:

```{r}
#| eval: false
raw_files <- list.files(
  "data/raw/gpfg",
  pattern = "xlsx$",
  full.names = TRUE
)

combined <- raw_files |>
  map_dfr(read_one_annual_file)
```

`read_one_annual_file()` is defined and validated in
`scripts/prepare_gpfg.R`. Keeping that logic in a script prevents every chapter
from cleaning the same files differently.

## Join variables with `left_join()`

A join adds columns by matching keys. First create a one-row-per-year table:

```{r}
annual_totals <- gpfg_panel |>
  summarise(
    annual_market_value_nok = sum(market_value_nok),
    annual_records = n(),
    .by = year
  )
```

Join each year's total back to every holding in that year:

```{r}
rows_before <- nrow(gpfg_panel)

gpfg_with_share <- gpfg_panel |>
  left_join(
    annual_totals,
    by = "year",
    relationship = "many-to-one"
  ) |>
  mutate(
    share_of_annual_value = market_value_nok / annual_market_value_nok
  )

stopifnot(nrow(gpfg_with_share) == rows_before)
```

`relationship = "many-to-one"` documents the expectation that many holdings
match exactly one annual total. If the right-hand table accidentally contains
two rows for a year, the join fails instead of silently duplicating holdings.

## Audit unmatched keys

Never assume that two tables use the same labels. Before mapping, for example,
country names may differ:

```{r}
country_lookup_example <- tibble(
  country = c("United States", "China", "Japan"),
  short_label = c("US", "China", "Japan")
)

gpfg_latest |>
  distinct(country) |>
  anti_join(country_lookup_example, by = "country") |>
  slice_head(n = 10)
```

`anti_join()` answers “which keys on the left have no match on the right?” A
missing match is not automatically safe to drop. It may reveal a spelling
difference, a changed country name, a different definition, or genuinely
missing reference data.

Use these checks around a consequential join:

```{r}
#| eval: false
left_only  <- anti_join(left_table, right_table, by = "key")
right_only <- anti_join(right_table, left_table, by = "key")

joined <- left_table |>
  left_join(right_table, by = "key", relationship = "many-to-one")

stopifnot(nrow(joined) == nrow(left_table))
```

## Summarise before comparing years

Company names are not stable identifiers across a decade. For the main trend
analysis, aggregate to a defensible grain such as country-year:

```{r}
country_year <- gpfg_panel |>
  summarise(
    market_value_nok = sum(market_value_nok),
    holding_records = n(),
    .by = c(year, country)
  )

country_year
```

One row now represents one country in one year.

## Pivot to compare columns

Long data are usually easiest for analysis and plotting. Wide data can make a
two-year comparison easier to read:

```{r}
country_change <- country_year |>
  filter(year %in% c(start_year, end_year)) |>
  mutate(period = if_else(year == start_year, "start", "end")) |>
  select(period, country, market_value_nok) |>
  pivot_wider(
    names_from = period,
    values_from = market_value_nok,
    names_prefix = "value_"
  ) |>
  mutate(
    absolute_change_nok = value_end - value_start,
    percent_change = absolute_change_nok / value_start
  )

country_change |>
  arrange(desc(absolute_change_nok)) |>
  slice_head(n = 10)
```

Inspect missing values after the pivot. A country present in the latest year but absent in
the starting year has no valid percent change because the denominator is missing or
zero.

Return to long form with:

```{r}
country_change |>
  select(country, value_start, value_end) |>
  pivot_longer(
    cols = starts_with("value_"),
    names_to = "year",
    names_prefix = "value_",
    values_to = "market_value_nok"
  ) |>
  mutate(year = if_else(year == "start", start_year, end_year))
```

## What a join cannot solve

A join matches labels; it does not prove that the matched records represent the
same real-world entity. Over ten years, company names can change because of
mergers, demergers, punctuation, rebranding, or data-entry conventions. Only
3,285 published names appear in all ten teaching snapshots.

For beginner analysis, use country-, region-, or carefully defined
industry-level trends. Treat company-level longitudinal matching as an advanced
record-linkage project requiring external identifiers and manual verification.

## Independent task

Choose three countries and create a country-year table for the rolling ten-year period. Then:

1. confirm there is at most one row per country-year;
2. pivot it wide and calculate start-to-end change;
3. identify any missing starting or ending values;
4. pivot it back to long form; and
5. write a methods note explaining why the result is about year-end market
   value, not net investment flows.
