---
title: "Combine and Reshape"
---

## Learning objectives

By the end of this chapter, you should be able to:

- combine similar tables with `bind_rows()`;
- add columns from another table with `left_join()`;
- choose a suitable join key;
- reshape data with `pivot_wider()` and `pivot_longer()`;
- calculate annual totals and year-to-year change; and
- describe a trend without confusing it with investment flows or returns.

## Where this chapter fits

The previous chapter worked with one table. Reporting projects often contain
several annual files or related tables. The tidyverse gives us tools for
preparing those tables and answering questions across time:

| Task | Function |
|---|---|
| Put similar observations underneath one another | `bind_rows()` |
| Add variables by matching a key | `left_join()` |
| Change the arrangement of rows and columns | `pivot_wider()` or `pivot_longer()` |

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `05-combine-reshape.Rmd`. Download
[`gpfg_equities_last_10_years.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_equities_last_10_years.csv)
and save it in the same `data` folder. Keep the latest-year file there too.
:::

## Load the multi-year data

```{r}
library(tidyverse)

gpfg_panel <- read_csv("data/gpfg_equities_last_10_years.csv")

gpfg_panel |>
  count(year)
```

One row represents one reported holding in one year. The `year` column keeps
observations from different files distinguishable.

The result should show ten consecutive years. Check this before calculating a
trend. If a year is missing, determine whether the source file is absent or
whether the publisher changed the data.

## Stack rows with `bind_rows()`

Suppose two annual tables have the same columns. `bind_rows()` places the rows
of the second table below the first:

```{r}
holdings_2024 <- gpfg_panel |>
  filter(year == 2024)

holdings_2025 <- gpfg_panel |>
  filter(year == 2025)

two_years <- bind_rows(holdings_2024, holdings_2025)

two_years |>
  count(year)
```

Use `bind_rows()` when the tables describe the same variables but different
observations. Make sure a year or date column exists before combining annual
files.

::: {.callout-tip title="Check names before stacking"}
Columns with the same meaning should have the same name. If one file uses
`company` and another uses `company_name`, rename them consistently before
using `bind_rows()`.
:::

## Add columns with `left_join()`

A join connects two tables by a shared column called a **key**. First create a
country summary and a country-to-region reference table. `distinct()` keeps
each country-region combination once:

```{r}
country_totals <- holdings_2025 |>
  group_by(country) |>
  summarise(market_value_nok = sum(market_value_nok))

country_regions <- holdings_2025 |>
  distinct(country, region)
```

`left_join()` keeps every row from the table on the left and adds matching
columns from the table on the right:

```{r}
country_totals <- country_totals |>
  left_join(country_regions, by = "country")

country_totals |>
  slice_head(n = 5)
```

The essential join argument is `by`, which names the matching key. Here the
value in `country` must match in both tables.

Check keys that did not match with `anti_join()`:

```{r}
country_totals |>
  anti_join(country_regions, by = "country")
```

An empty result means every country in the left table found a match. It does
not prove that the labels themselves are correct, so important matches still
need source checking.

## Long and wide data

**Long data** usually place the year in one column and the value in another:

| country | year | market_value_nok |
|---|---:|---:|
| China | 2024 | ... |
| China | 2025 | ... |

**Wide data** place different years in different columns:

| country | 2024 | 2025 |
|---|---:|---:|
| China | ... | ... |

Long data are convenient for grouping and charts. Wide data can make a direct
two-year comparison easier.

Create a long country-year table:

```{r}
country_year <- two_years |>
  group_by(country, year) |>
  summarise(market_value_nok = sum(market_value_nok))

country_year
```

## Make data wider

`pivot_wider()` needs to know which column supplies the new column names and
which column supplies their values:

```{r}
country_wide <- country_year |>
  pivot_wider(
    names_from = year,
    values_from = market_value_nok
  )

country_wide
```

Each country now has separate columns for 2024 and 2025.

## Make data longer

`pivot_longer()` gathers those year columns back into two columns:

```{r}
country_long <- country_wide |>
  pivot_longer(
    cols = c(`2024`, `2025`),
    names_to = "year",
    values_to = "market_value_nok"
  )

country_long
```

The essential arguments identify the columns to gather, the name of the new
category column, and the name of the new value column.

## Use the combined data to answer trend questions

Combining annual observations is not an end in itself. It allows us to ask:

> How did reported year-end equity market value change over the ten-year
> period?

A trend requires several comparable time points. Before calculating it, make
sure the row meaning, units, and categories are reasonably consistent across
the years.

### Calculate annual totals

Group the combined data by year and add the market values:

```{r}
annual <- gpfg_panel |>
  group_by(year) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(year)

annual
```

One row now represents one annual snapshot. Ordering by year matters because
the next calculation compares each row with the row before it.

### Calculate year-to-year change

`lag()` returns the previous value in a column. Add it first so the comparison
is visible:

```{r}
annual_change <- annual |>
  mutate(previous_year_value = lag(market_value_nok))

annual_change
```

The first row contains `NA` because this file has no earlier year with which to
compare it. Now calculate absolute and percentage change:

```{r}
annual_change <- annual_change |>
  mutate(
    change_nok = market_value_nok - previous_year_value,
    change_percent = change_nok / previous_year_value * 100
  )

annual_change
```

::: {.callout-tip title="Sort before using lag()"}
`lag()` uses the current row order. Arrange a time series by year before
calculating change, even when the source appears to be correctly ordered.
:::

### Follow the same markets over time

Choose a small, fixed set of markets so every year refers to the same groups:

```{r}
selected_countries <- c("United States", "China", "Japan")

country_trends <- gpfg_panel |>
  filter(country %in% selected_countries) |>
  group_by(year, country) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(country, year)

country_trends
```

Selecting a different set of leading markets in every year would change the
membership of the comparison and make the pattern harder to interpret.

### Compare amounts and shares

A market's value can rise while its share of the full portfolio falls. First
create the annual denominator:

```{r}
annual_totals <- gpfg_panel |>
  group_by(year) |>
  summarise(annual_market_value_nok = sum(market_value_nok))
```

Join the denominator to the selected-market table and calculate the share:

```{r}
country_trends <- country_trends |>
  left_join(annual_totals, by = "year") |>
  mutate(
    share_percent = market_value_nok / annual_market_value_nok * 100
  )

country_trends
```

The value answers “how many Norwegian kroner?” The share answers “what
percentage of the reported annual equity total?” A finding should name which
measure it uses.

### Interpret changes carefully

A change in year-end market value can reflect:

- changes in security prices;
- exchange-rate movements;
- purchases and sales;
- companies entering or leaving the portfolio; or
- changes in classifications and reporting.

The holdings data do not separate these causes. Do not call the percentage
change an investment return, and do not assume it equals the amount purchased
during the year.

::: {.callout-note title="Categories can change over time"}
NBIM's industry labels change within the historical series. Compare the labels
used in each period before drawing an industry trend. Company names can also
change after mergers or rebranding, so market- or industry-level trends are
usually safer for a beginner project.
:::

## Practice

### Activity 1: Reshape two years

Use the 2024 and 2025 holdings to:

1. calculate one row per industry and year;
2. pivot the result so the two years appear in separate columns;
3. identify industries missing from either year; and
4. pivot the table back to long form.

Explain what one row represents before and after each pivot.

### Activity 2: Investigate one market over time

Choose one market and create a year-by-year table containing:

1. market value in Norwegian kroner;
2. previous-year market value;
3. absolute annual change;
4. percentage annual change; and
5. share of the full annual total.

Identify the largest rise and fall by sorting the table. Write a headline that
describes the change without using the words *return*, *profit*, or
*purchased*.

## Takeaways

| Function | What it does |
|---|---|
| `bind_rows()` | Stacks similar observations underneath one another |
| `distinct()` | Keeps unique combinations of selected columns |
| `left_join()` | Adds columns by matching a key while keeping the left table |
| `anti_join()` | Shows rows on the left that have no match on the right |
| `pivot_wider()` | Spreads one category across several columns |
| `pivot_longer()` | Gathers several columns into a names column and a values column |
| `group_by(year)` + `summarise()` | Creates one summary row per year |
| `arrange(year)` | Places observations in chronological order |
| `lag()` | Returns the value from the previous row |
| `mutate()` | Calculates changes, percentages, and shares |
| Fixed comparison set | Keeps the same groups in every period |
| Amount and share | Answer different questions about change |

::: {.callout-tip title="Want more control?"}
Begin with the essential arguments above and inspect every result. The official
[tidyr pivoting guide](https://tidyr.tidyverse.org/articles/pivot.html) and
[dplyr reference](https://dplyr.tidyverse.org/reference/index.html) cover
additional join, reshape, and time-comparison options for unusual cases.
:::
