---
title: "Join and Reshape"
---

::: {.callout-note title="Part 5: Advanced Workflows"}
The earlier chapters used one long table: each row was a holding in one year.
Real reporting projects often arrive as several tables, or in a layout made
for reading rather than analysis. This chapter introduces two ways to prepare
those data: reshape columns into rows and join related tables.
:::

## Learning objectives

By the end of this chapter, you should be able to:

- distinguish long data from wide data;
- use `pivot_longer()` to turn year columns into observations;
- identify a suitable key shared by two tables;
- use `anti_join()` to find unmatched keys;
- use `left_join()` to add variables from a lookup table; and
- use `pivot_wider()` when a reporting output needs separate columns.

## Begin with a reporting need

Suppose a colleague sends two files:

- `gpfg_country_wide.csv` contains one row per country and a separate column
  for each year;
- `gpfg_country_lookup.csv` connects each country to a region.

We want to answer:

> How did reported equity value change by region from 2021 through 2025?

Neither file can answer the question alone. The values must first be reshaped
so that `year` is a variable, and then the region labels must be added from the
second table.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `09-join-reshape.Rmd`. Download
[`gpfg_country_wide.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_country_wide.csv)
and
[`gpfg_country_lookup.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_country_lookup.csv),
then save both in `data/`.
:::

```{r}
library(tidyverse)

country_wide <- read_csv("data/gpfg_country_wide.csv")
country_lookup <- read_csv("data/gpfg_country_lookup.csv")
```

Inspect the two files before changing them:

```{r}
glimpse(country_wide)
```

```{r}
glimpse(country_lookup)
```

The first table contains the values, but its years are column names. The
second contains `region`, but no annual values. Both contain `country`, which
can connect them.

## Wide and long data

In the wide table, one row represents one country and each year has its own
column:

| country | 2021 | 2022 | 2023 | 2024 | 2025 |
|---|---:|---:|---:|---:|---:|
| Australia | ... | ... | ... | ... | ... |

For analysis and visualization, we usually want one observation per row. A
long table stores the year in one column and its value in another:

| country | year | market_value_nok |
|---|---:|---:|
| Australia | 2021 | ... |
| Australia | 2022 | ... |

The long layout makes `year` available for `filter()`, `group_by()`, and the
x-axis of a chart. It also avoids writing separate code for every annual
column.

## Convert wide data to long data

`pivot_longer()` needs three pieces of information:

- `cols`: the columns to move into rows;
- `names_to`: the name of the new column that will store the old column
  names; and
- `values_to`: the name of the new column that will store their values.

Here, every column except `country` is a year column:

```{r}
country_year <- country_wide |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "market_value_nok"
  )

country_year
```

`cols = -country` means “reshape every column except `country`.” The old
column names become values of `year`.

### Correct the new year type

Column names are text, so `year` is now a character column even though its
values look like numbers. Convert it after reshaping:

```{r}
country_year <- country_year |>
  mutate(year = parse_integer(year))

glimpse(country_year)
```

`parse_integer()` reads text such as `"2021"` as an integer. We use it here
because the year values came from column names. This is a useful example of a
cleaning step being prompted by a real change in the data.

Check the new structure:

```{r}
country_year |>
  count(year)
```

One row now represents one country in one year. Count empty values as well:

```{r}
country_year |>
  group_by(year) |>
  summarise(missing_values = sum(is.na(market_value_nok)))
```

A missing value means the wide source had an empty cell for that country-year
combination. It does not mean the fund reported a value of zero.

## Join variables from another table

The long table still has no region. A **join** matches rows across tables by a
shared column called a **key**. Here, `country` is the proposed key.

Before joining, check that every country appears only once in the lookup
table:

```{r}
country_lookup |>
  count(country) |>
  filter(n > 1)
```

An empty result means there are no duplicate country keys. If a country
appeared twice, joining it could unexpectedly create extra rows.

Next, find source countries that have no match in the lookup table:

```{r}
country_year |>
  distinct(country) |>
  anti_join(country_lookup, by = "country")
```

`anti_join()` keeps keys from the left table that are absent from the right.
An empty result means every country found a match. This is a useful check, not
the final combined table.

Now add `region`:

```{r}
country_year <- country_year |>
  left_join(country_lookup, by = "country")

glimpse(country_year)
```

`left_join()` keeps every row from `country_year`, the table on the left, and
adds the matching `region` from `country_lookup`. The essential `by` argument
states which column should match.

::: {.callout-important title="Always check the row meaning after a join"}
Before the join, one row represented one country-year. That should still be
true afterward. Duplicate keys in the lookup table can multiply rows and
inflate later totals.
:::

## Answer the question

Now the combined table contains all three columns needed for the question:
`year`, `region`, and `market_value_nok`.

```{r}
region_year <- country_year |>
  group_by(year, region) |>
  summarise(market_value_nok = sum(market_value_nok, na.rm = TRUE)) |>
  arrange(year, desc(market_value_nok))

region_year
```

The result has one row per region and year. It can now be used for a comparison
table or a multi-line chart. Notice the workflow: the reporting question told
us which variables were missing and therefore which reshape and join were
necessary. Here `na.rm = TRUE` tells `sum()` to ignore unavailable
country-year cells; it does not replace those cells with known zeros.

## When would we make data wider?

Long data are usually easier to analyse in R. Wide data can still be helpful
for a compact reporting table—for example, one row per country with 2024 and
2025 values side by side.

```{r}
country_change <- country_year |>
  filter(year %in% c(2024, 2025)) |>
  select(country, year, market_value_nok) |>
  pivot_wider(
    names_from = year,
    values_from = market_value_nok
  )

country_change
```

`pivot_wider()` uses values from `year` as new column names and values from
`market_value_nok` to fill those columns. Use this layout when the output calls
for it; keep the long version for most analysis and charts.

## Choose the operation by the problem

| Situation | Function |
|---|---|
| Same columns, different observations | `bind_rows()` |
| Variables spread across repeated columns | `pivot_longer()` |
| Add variables by matching a key | `left_join()` |
| Find keys that did not match | `anti_join()` |
| Put category values into separate columns | `pivot_wider()` |

Chapter 6 used `bind_rows()` because the annual files had the same columns and
needed to be stacked. This chapter uses `left_join()` because two tables hold
different variables about the same countries.

## Practice

1. Import the wide and lookup files and explain what one row represents in
   each.
2. Reshape the annual columns into `year` and `market_value_nok`.
3. Confirm that `year` has the correct type.
4. Check duplicate and unmatched country keys before joining.
5. Join the region labels and calculate one annual total per region.
6. Write one sentence explaining why the reshape and join were necessary.

## Takeaways

| Function | What it does |
|---|---|
| `pivot_longer()` | Moves repeated columns into name and value columns |
| `parse_integer()` | Reads whole numbers stored as text |
| `count()` | Checks how often each key or category appears |
| `anti_join()` | Finds keys from the left table without a match |
| `left_join()` | Keeps the left table and adds matching columns |
| `pivot_wider()` | Moves category values into separate columns |

In Chapter 10, we will use the same join logic to connect investment-market
summaries to geographic boundary coordinates.

::: {.callout-tip title="Want more control?"}
The [tidyr pivoting guide](https://tidyr.tidyverse.org/articles/pivot.html)
shows more reshaping patterns. The [dplyr joins guide](https://dplyr.tidyverse.org/articles/two-table.html)
explains additional joins. Begin with `pivot_longer()`, `left_join()`, and
`anti_join()`; use another operation only when the reporting task requires it.
:::
