---
title: "Explanatory Visualization"
---

## Learning objectives

By the end of this chapter, you should be able to:

- design a chart around one defensible takeaway;
- order and label comparisons clearly;
- show units, dates, sources, and transformations;
- remove visual elements that do not serve the evidence; and
- write a headline and caption that match the calculation.

## Begin with a chart brief

Before coding, complete five lines:

1. **Audience:** Who needs to understand this?
2. **Question:** What exact comparison matters?
3. **Takeaway:** What does the analysis show?
4. **Chart form:** Why is this form appropriate?
5. **Caveat:** What must the caption or story explain?

We will turn the latest-year country ranking into an explanatory bar chart.

```{r}
library(tidyverse)
library(scales)

gpfg <- read_csv(
  "data/processed/gpfg_equities_latest.csv",
  show_col_types = FALSE
)

latest_year <- unique(gpfg$year)

country_summary <- gpfg |>
  summarise(
    market_value_usd = sum(market_value_usd),
    .by = country
  ) |>
  mutate(share_of_total = market_value_usd / sum(market_value_usd)) |>
  arrange(desc(market_value_usd))

top_10_countries <- country_summary |>
  slice_head(n = 10)

top_country <- top_10_countries |>
  slice_head(n = 1)
```

## Build a ranked bar chart

```{r}
country_chart <- top_10_countries |>
  ggplot(
    aes(
      x = market_value_usd,
      y = fct_reorder(country, market_value_usd)
    )
  ) +
  geom_col(fill = "#1769AA", width = 0.72) +
  geom_text(
    aes(label = percent(share_of_total, accuracy = 0.1)),
    hjust = -0.15,
    size = 3.4
  ) +
  scale_x_continuous(
    labels = label_dollar(scale = 1e-9, suffix = "B"),
    expand = expansion(mult = c(0, 0.16))
  ) +
  labs(
    title = paste0(
      top_country$country,
      " accounted for ",
      percent(top_country$share_of_total, accuracy = 0.1),
      " of reported equity value"
    ),
    subtitle = paste0("Ten largest investment markets in NBIM's ", latest_year, " year-end equity holdings"),
    x = "Market value in US dollars",
    y = NULL,
    caption = paste(
      "Labels show share of the full reported equity total.",
      "Source: Norges Bank Investment Management."
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0)
  )

country_chart
```

The bars show exact market-value comparisons; the labels answer the denominator
question. A rainbow palette would add categories that do not exist, so one
color is enough.

## Turn a trend into a clear line chart

```{r}
gpfg_panel <- readRDS(
  "data/processed/gpfg_equities_last_10_years.rds"
)

annual <- gpfg_panel |>
  summarise(
    market_value_nok = sum(market_value_nok),
    .by = year
  ) |>
  arrange(year) |>
  mutate(market_value_nok_trillions = market_value_nok / 1e12)

panel_years <- annual$year
start_year <- min(panel_years)
end_year <- max(panel_years)

endpoints <- annual |>
  filter(year %in% range(year))
```

```{r}
trend_chart <- ggplot(
  annual,
  aes(x = year, y = market_value_nok_trillions)
) +
  geom_line(color = "#1769AA", linewidth = 1.1) +
  geom_point(color = "#1769AA", size = 2) +
  geom_text(
    data = endpoints,
    aes(label = paste0(round(market_value_nok_trillions, 1), " tn")),
    vjust = -0.8,
    fontface = "bold"
  ) +
  scale_x_continuous(breaks = panel_years) +
  scale_y_continuous(
    labels = label_number(suffix = " tn"),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Reported equity market value more than tripled over the decade",
    subtitle = paste0("Year-end holdings in nominal Norwegian kroner, ", start_year, "–", end_year),
    x = NULL,
    y = "NOK trillions",
    caption = str_wrap(
      paste(
        "Values reflect prices, currencies, purchases, and sales; they are not investment returns.",
        "Source: Norges Bank Investment Management."
      ),
      width = 110
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0, size = 8.5),
    plot.caption.position = "plot"
  )

trend_chart
```

The title is descriptive rather than causal. The subtitle states the time,
snapshot basis, currency, and nominal units. The caption names the main
interpretive limitation.

## Use color and annotation with restraint

Color should do a job:

- highlight one comparison;
- distinguish a small number of series;
- encode a meaningful ordered scale; or
- signal a consistent category across several charts.

Avoid relying on red versus green alone, which is difficult for many readers.
Check that the chart remains understandable in grayscale and that labels carry
essential information.

Annotations should explain evidence, not repeat the axis. A useful annotation
may identify a documented policy change, classification break, or verified
event. Do not annotate a data jump with a causal explanation that the dataset
cannot establish.

## Export with reproducible dimensions

```{r}
#| eval: false
ggsave(
  file.path("outputs", paste0("country-holdings-", latest_year, ".png")),
  country_chart,
  width = 8,
  height = 5,
  dpi = 300
)
```

Choose dimensions for the publishing context. Always inspect the exported file
at its intended display size; labels that look fine in RStudio may be unreadable
on a phone.

## Publication checklist

Before publication, confirm:

- the title states one evidence-backed takeaway;
- the subtitle defines time, population, and measure;
- axes begin at a defensible point and disclose transformations;
- categories are sorted meaningfully;
- units and denominators are explicit;
- labels are readable and not clipped;
- color has a semantic purpose;
- the caption identifies the source and relevant caveat; and
- every number reconciles with a table produced by the analysis.

## Independent task

Take one exploratory chart from the previous chapter and turn it into a
publication chart. Submit:

1. the five-line chart brief;
2. the final chart;
3. the table used to verify its values;
4. alt text of no more than 100 words; and
5. a short note describing one design choice you rejected and why.
