---
title: "Visualization II"
---

## Learning objectives

By the end of this chapter, you should be able to:

- add color or fill only when it represents useful information;
- distinguish a fixed appearance from a data mapping;
- compare several time series;
- choose between grouped, stacked, and percentage-stacked bars;
- compare distributions with box plots;
- separate crowded groups with facets; and
- save a finished chart.

## When does a chart need more complexity?

Chapter 7 used a line, bars, a histogram, and a scatterplot to answer four
basic questions. An advanced chart should begin from one of those forms and
add something only because the reporting question requires it.

| New reporting need | Possible addition |
|---|---|
| Compare several groups over time | Map a group to color |
| Compare groups directly within categories | Grouped bars |
| Show how groups contribute to a total | Stacked bars |
| Compare composition rather than totals | Percentage-stacked bars |
| Compare several distributions | Box plots |
| Reduce overlap among many groups | Facets or small multiples |

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `08-data-visualization-ii.Rmd`. Reuse
`data/gpfg_5_years.csv` and save finished charts in `outputs/`.
:::

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_5_years.csv")

gpfg_latest <- gpfg |>
  filter(year == 2025)
```

## Map a column or set an appearance

This distinction matters throughout the chapter:

```r
geom_point(color = "#0072B2")
```

`color` is outside `aes()`, so every point receives the same fixed color. It
does not create a legend.

```r
geom_point(aes(color = region))
```

`color` is inside `aes()`, so values of `region` are mapped to different
colors. Because color now represents data, ggplot2 creates a legend.

The same rule applies to `fill`, `size`, and other visual properties.

## Compare several trends with color

### Step 1: state the question

> How did the total reported market value of NBIM's equity holdings in the
> United States, China, and Japan change from 2021 through 2025?

### Step 2: identify and prepare the columns

We need `year`, `country`, and `market_value_nok`. First choose a fixed set of
markets, then calculate one value per country and year:

```{r}
selected_countries <- c("United States", "China", "Japan")

country_trends <- gpfg |>
  filter(country %in% selected_countries) |>
  group_by(year, country) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  mutate(market_value_nok_billions = market_value_nok / 1000000000) |>
  arrange(country, year)

country_trends
```

One row represents one selected market in one year.
The `%in%` operator asks whether each `country` belongs to the
`selected_countries` vector created above.

### Step 3: plan the visual roles

| Visual role | Column | Reason |
|---|---|---|
| X-axis | `year` | Ordered time |
| Y-axis | `market_value_nok_billions` | Value being compared |
| Color | `country` | Separates the three time series |
| Size | None | Line position already represents value |

### Step 4: begin with the essential mappings

```{r}
country_trend_plot <- country_trends |>
  ggplot(
    aes(
      x = year,
      y = market_value_nok_billions,
      color = country
    )
  ) +
  geom_line()

country_trend_plot
```

Because `country` is inside `aes()`, the chart has a color legend. Color is
doing analytical work: it identifies which line belongs to which market.

### Step 5: add points and labels

```{r}
country_trend_plot <- country_trend_plot +
  geom_point() +
  labs(
    title = "Selected investment markets followed different paths",
    subtitle = "Reported year-end equity holdings, 2021–2025",
    x = "Year",
    y = "Market value in NOK billions",
    color = "Investment market",
    caption = "Source: Norges Bank Investment Management"
  )

country_trend_plot
```

The `color` entry in `labs()` gives the automatically created legend a clearer
title.

### Step 6: add a restrained theme and palette

```{r}
country_trend_plot <- country_trend_plot +
  scale_color_manual(
    values = c(
      "United States" = "#0072B2",
      "China" = "#D55E00",
      "Japan" = "#009E73"
    )
  ) +
  theme_minimal() +
  theme(legend.position = "top")

country_trend_plot
```

The manual scale gives every named market a consistent color. Add it only
after confirming that the basic multi-line chart works.

## Compare groups with grouped bars

### Question and analysis table

> How did the total market value of NBIM's reported equity holdings differ
> across three industries and three regions in 2025?

```{r}
region_industry <- gpfg_latest |>
  filter(
    region %in% c("Asia", "Europe", "North America"),
    industry %in% c("Technology", "Financials", "Industrials")
  ) |>
  group_by(industry, region) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  mutate(market_value_nok_billions = market_value_nok / 1000000000)

region_industry
```

The two conditions inside `filter()` are separated by a comma, so both must be
true: a row must belong to one of the selected regions and one of the selected
industries.

The visual roles are industry on x, value on y, and region mapped to fill.
Start with the default bars:

```{r}
region_industry |>
  ggplot(
    aes(
      x = industry,
      y = market_value_nok_billions,
      fill = region
    )
  ) +
  geom_col()
```

The default is stacked. This shows the combined height and each region's
contribution, but only the bottom segment begins at the same point, so the
other segments are harder to compare precisely.

For direct comparison, change one argument:

```{r}
region_industry |>
  ggplot(
    aes(
      x = industry,
      y = market_value_nok_billions,
      fill = region
    )
  ) +
  geom_col(position = "dodge")
```

`position = "dodge"` places the regions beside one another. The data and
aesthetic mappings have not changed; only the bar arrangement has changed.

Add labels and styling after choosing the arrangement that fits the question:

```{r}
region_industry |>
  ggplot(
    aes(
      x = industry,
      y = market_value_nok_billions,
      fill = region
    )
  ) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(
      "Asia" = "#009E73",
      "Europe" = "#E69F00",
      "North America" = "#0072B2"
    )
  ) +
  labs(
    title = "Regional values differed across selected industries",
    x = NULL,
    y = "Market value in NOK billions",
    fill = "Region",
    caption = "Three selected industries and regions; year-end 2025\nSource: Norges Bank Investment Management"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## Compare composition with percentage bars

A different question requires a different bar arrangement:

> Within each selected industry, what share of NBIM's reported total across
> the three selected regions came from each region?

Starting from the same table and mappings, change the position to `"fill"`:

```{r}
region_industry |>
  ggplot(
    aes(
      x = industry,
      y = market_value_nok_billions,
      fill = region
    )
  ) +
  geom_col(position = "fill") +
  scale_fill_manual(
    values = c(
      "Asia" = "#009E73",
      "Europe" = "#E69F00",
      "North America" = "#0072B2"
    )
  ) +
  labs(
    title = "Regional composition differed across selected industries",
    x = NULL,
    y = "Share of selected-region market value",
    fill = "Region",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

Every bar now has the same height. This makes composition easier to compare
but hides differences in total amounts. A percentage chart must not be used to
support a claim about which industry has the largest total.

## Compare distributions with box plots

Chapter 7's histogram described all 2025 holding values together. Now ask:

> Among companies in NBIM's 2025 data, how did the distribution of reported
> holding values differ across industries?

The source table already contains the two required columns:

| Visual role | Column |
|---|---|
| X-axis | `market_value_usd` |
| Y-axis | `industry` |

Begin without additional scales:

```{r}
industry_box <- gpfg_latest |>
  ggplot(
    aes(
      x = market_value_usd,
      y = industry
    )
  ) +
  geom_boxplot()

industry_box
```

The large values compress most boxes near zero. Because all displayed values
are positive, add a logarithmic x-axis:

```{r}
industry_box +
  scale_x_log10() +
  labs(
    title = "Holding values varied widely within every industry",
    x = "Market value in USD (log scale)",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

The line inside each box is the median. Points beyond the whiskers deserve
inspection, but they are not automatically errors. Always disclose a log
scale in the axis label.

## Add a third variable with facets

Return to the basic relationship question from Chapter 7:

> Among companies in NBIM's 2025 data, were higher ownership percentages
> associated with larger holding values, and did that pattern differ by region?

First select three regions so the comparison remains readable:

```{r}
selected_holdings <- gpfg_latest |>
  filter(region %in% c("Asia", "Europe", "North America"))
```

Begin with the familiar scatterplot, adding transparency because many points
overlap:

```{r}
relationship_plot <- selected_holdings |>
  ggplot(
    aes(
      x = market_value_usd,
      y = ownership_pct
    )
  ) +
  geom_point(alpha = 0.25)

relationship_plot
```

Now add one new layer, `facet_wrap()`, to create a panel for each region:

```{r}
relationship_plot +
  scale_x_log10() +
  facet_wrap(~ region) +
  labs(
    title = "The relationship can be inspected separately by region",
    x = "Market value in USD (log scale)",
    y = "Ownership (%)",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

The panels use the same axes, making their patterns easier to compare than
many overlapping colors in one panel.

## Save a finished chart

Before saving, verify that the chart matches the analysis table and that its
title, period, measure, unit, and source are clear.

```{r}
#| eval: false
ggsave("outputs/country-trends.png", country_trend_plot)
```

The filename and chart object are enough here. Width, height, and resolution
are optional controls for a later publication requirement.

::: {.callout-note title="Interactive charts are optional"}
Interaction can add tooltips, filters, and zooming, but it also introduces new
packages and HTML-specific behavior. First make a correct, readable static
chart. Interactive graphics can be considered later when interaction helps a
specific reporting task.
:::

## Practice

Choose one of the following:

1. Compare three industries across five years with colored lines.
2. Compare selected regions with grouped and stacked bars, then explain which
   arrangement better answers your question.
3. Add facets to a relationship plot and explain what became easier to see.

For every version, state the question, list the columns, identify the visual
roles, show the basic plot first, and explain every added layer.

## Takeaways

| Function or option | What it adds |
|---|---|
| `color = column` inside `aes()` | Maps a column to line or point color |
| `fill = column` inside `aes()` | Maps a column to bar fill |
| Fixed color outside `aes()` | Changes appearance without representing data |
| `position = "dodge"` | Places groups beside one another |
| `position = "fill"` | Converts stacked bars to equal-height composition bars |
| `geom_boxplot()` | Compares numeric distributions across groups |
| `scale_x_log10()` | Spreads positive values across a logarithmic axis |
| `facet_wrap()` | Creates a separate panel for each group |
| `scale_color_manual()` | Assigns deliberate colors to named groups |
| `ggsave()` | Saves a completed chart |

::: {.callout-tip title="Want more control?"}
The official [ggplot2 reference](https://ggplot2.tidyverse.org/reference/)
documents additional scales, legends, themes, annotations, and saving options.
Add them one at a time and only when they serve the reporting question.
:::
