---
title: "Visualization I"
---

::: {.callout-note title="Part 4: Data Visualization"}
Chapters 5 and 6 calculated and checked single-year and multi-year findings.
We will now turn those tables into charts. Every chart begins with a reporting
question and a plan; appearance comes later.
:::

## Learning objectives

By the end of this chapter, you should be able to:

- choose a chart based on a reporting question;
- identify which columns belong on the x- and y-axes;
- decide whether color or size needs to represent another variable;
- build a chart from a basic version to a labelled, styled version;
- explain what one visual mark represents; and
- distinguish a chart's data mappings from its appearance settings.

## A question-to-chart workflow

Use the same sequence every time:

1. State the reporting question.
2. Identify the columns needed to answer it.
3. Calculate and inspect the table that will be plotted.
4. Decide which column belongs on x, y, color, or size.
5. Make the simplest working plot.
6. Check what each bar, point, or line represents.
7. Add labels and a source.
8. Add color and a theme only when they improve communication.

The chart does not replace the analysis. It displays a result that should
already be visible in a checked table.

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `07-data-visualization-i.Rmd`. Reuse the combined
file created in Chapter 6. A ready-made
[`gpfg_5_years.csv`](https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_5_years.csv)
is available if needed.
:::

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_5_years.csv")

gpfg_latest <- gpfg |>
  filter(year == 2025)
```

## Show a trend with a line chart

We will develop the first chart slowly so that every addition is visible.

### Step 1: state the question

> How did the total reported market value of NBIM's equity holdings change
> from 2021 through 2025?

### Step 2: identify the columns

The source table contains many holdings per year. The question needs:

| Column | Role in the question |
|---|---|
| `year` | Time |
| `market_value_nok` | Measure to add within each year |

### Step 3: prepare and inspect the table

Reuse the group-and-summarise pattern from Chapter 6:

```{r}
annual_summary <- gpfg |>
  group_by(year) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(year) |>
  mutate(market_value_nok_trillions = market_value_nok / 1000000000000)

annual_summary
```

One row now represents one year. The value in trillions is the same measure
expressed in a unit that will be easier to label.

### Step 4: plan the visual roles

| Visual role | Column | Why? |
|---|---|---|
| X-axis | `year` | Time has a meaningful order |
| Y-axis | `market_value_nok_trillions` | The value changes vertically |
| Color | None | We are showing only one series |
| Size | None | The line position already represents the value |

Do not add a variable to color or size simply because ggplot2 allows it.

### Step 5: make the basic plot

`ggplot()` receives the table. `aes()` maps columns to visual roles, and
`geom_line()` draws the selected geometry:

```{r}
trend_plot <- annual_summary |>
  ggplot(
    aes(
      x = year,
      y = market_value_nok_trillions
    )
  ) +
  geom_line()

trend_plot
```

This is already a complete plot. One position on the line represents the
total reported value for one annual snapshot.

### Step 6: show the observed years

Add points as another layer. The earlier line remains unchanged:

```{r}
trend_plot <- trend_plot +
  geom_point()

trend_plot
```

The points remind readers that the data contain five observed annual values;
the line connects them.

### Step 7: add labels

```{r}
trend_plot <- trend_plot +
  labs(
    title = "Reported equity value increased from 2021 to 2025",
    subtitle = "Year-end holdings in nominal Norwegian kroner",
    x = "Year",
    y = "NOK trillions",
    caption = "Source: Norges Bank Investment Management"
  )

trend_plot
```

The labels state the period, measure, unit, and source. The title describes a
visible pattern without calling the change a return.

### Step 8: add a theme

```{r}
trend_plot <- trend_plot +
  theme_minimal()

trend_plot
```

`theme_minimal()` changes presentation, not the underlying values.

### Step 9: make deliberate color choices

For the finished version, use blue for the line and orange for the observed
points:

```{r}
trend_final <- annual_summary |>
  ggplot(
    aes(
      x = year,
      y = market_value_nok_trillions
    )
  ) +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#D55E00", size = 2.5) +
  labs(
    title = "Reported equity value increased from 2021 to 2025",
    subtitle = "Year-end holdings in nominal Norwegian kroner",
    x = "Year",
    y = "NOK trillions",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()

trend_final
```

Here `color` and `size` are outside `aes()`. They set a fixed appearance and
do not represent additional columns, so the chart does not need a legend.

::: {.callout-important title="A line does not explain the change"}
The values can change because of prices, exchange rates, purchases, sales, or
changes in the reported portfolio. The chart shows the pattern, not its cause.
:::

## Show a ranking with a bar chart

Now repeat the workflow more quickly.

### Question and columns

> Which investment markets accounted for the largest total market value of
> NBIM's reported equity holdings at the end of 2025?

| Column | Role |
|---|---|
| `country` | Category being compared |
| `market_value_nok` | Measure added within each country |

### Prepare the table

```{r}
top_countries <- gpfg_latest |>
  group_by(country) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(desc(market_value_nok)) |>
  slice_head(n = 10) |>
  mutate(market_value_nok_billions = market_value_nok / 1000000000)

top_countries
```

One row represents one investment market.

### Plan the visual roles

| Visual role | Column | Why? |
|---|---|---|
| X-axis | `market_value_nok_billions` | Bar length represents the amount |
| Y-axis | `country` | Horizontal bars leave room for labels |
| Color | None | The ranking does not require another group |
| Size | None | Bar length already represents magnitude |

### Begin with the basic plot

```{r}
country_plot <- top_countries |>
  ggplot(
    aes(
      x = market_value_nok_billions,
      y = country
    )
  ) +
  geom_col()

country_plot
```

`geom_col()` uses values that we already calculated. One bar represents one
market, and its length represents total market value.

### Order the categories

The table is sorted, but categorical axes have their own order.
`fct_reorder()` orders the country labels by the plotted value:

```{r}
country_plot <- top_countries |>
  ggplot(
    aes(
      x = market_value_nok_billions,
      y = fct_reorder(country, market_value_nok_billions)
    )
  ) +
  geom_col()

country_plot
```

### Add communication layers

```{r}
country_plot <- country_plot +
  labs(
    title = "The United States led reported equity value by a wide margin in 2025",
    x = "Market value in NOK billions",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()

country_plot
```

The first chart taught every stage separately. Here we can add familiar
label and theme layers together. For the finished version, repeat the checked
chart with one deliberate fixed fill color:

```{r}
country_final <- top_countries |>
  ggplot(
    aes(
      x = market_value_nok_billions,
      y = fct_reorder(country, market_value_nok_billions)
    )
  ) +
  geom_col(fill = "#007C83") +
  labs(
    title = "The United States led reported equity value by a wide margin in 2025",
    x = "Market value in NOK billions",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()

country_final
```

The fill is outside `aes()` because all bars belong to one ranking rather than
representing separate groups.

## Describe a distribution with a histogram

### Question and visual plan

> How were the reported USD values of NBIM's company holdings distributed in
> 2025?

This question uses the numeric `market_value_usd` column. A histogram maps it
to x, divides its values into ranges called bins, and counts the observations
in each range. It does not require a y column from the source table.

```{r}
ggplot(gpfg_latest, aes(x = market_value_usd)) +
  geom_histogram()
```

Most holdings are small relative to a few very large holdings. This is a
right-skewed distribution. The basic chart reveals the problem clearly even
though the smaller observations are crowded together. Chapter 8 will add a
log scale and compare distributions across groups.

Add labels and a simple fixed color only after reading the basic result:

```{r}
ggplot(gpfg_latest, aes(x = market_value_usd)) +
  geom_histogram(fill = "#0072B2", color = "white") +
  labs(
    title = "Most individual holdings were small relative to the largest",
    x = "Market value in USD",
    y = "Number of holding records",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

## Examine a relationship with a scatterplot

### Question and visual plan

> Among companies in NBIM's 2025 equity-holdings data, were higher ownership
> percentages associated with larger reported holding values?

| Visual role | Column |
|---|---|
| X-axis | `market_value_usd` |
| Y-axis | `ownership_pct` |
| Color | None initially |
| Size | None initially |

Both variables are numeric, so begin with a scatterplot:

```{r}
ggplot(
  gpfg_latest,
  aes(
    x = market_value_usd,
    y = ownership_pct
  )
) +
  geom_point()
```

One point represents one holding. A visible relationship does not show that
one variable caused the other. Chapter 8 will address overlapping points,
skewed values, and comparisons among regions.

## Choose a basic chart from the question

| Question | Useful starting chart |
|---|---|
| How did a measure change over ordered time? | Line chart |
| Which categories have the largest calculated values? | Bar chart with `geom_col()` |
| How is one numeric column distributed? | Histogram |
| Do two numeric columns vary together? | Scatterplot |

The starting chart should be simple enough that students can explain every
mapping and every mark.

## Practice

Choose either an industry comparison or a 2024 country ranking. Complete the
full workflow:

1. write the reporting question;
2. list the columns needed;
3. prepare and inspect the table;
4. decide what belongs on x and y;
5. make the basic plot;
6. explain what one mark represents;
7. add labels and a source; and
8. add one theme or fixed-color choice.

## Takeaways

| Function or idea | What it does |
|---|---|
| `ggplot(data)` | Starts a chart with a table |
| `aes()` | Maps data columns to visual roles |
| `geom_line()` | Connects values across an ordered axis |
| `geom_point()` | Draws observations as points |
| `geom_col()` | Draws bars from values already calculated |
| `geom_histogram()` | Displays the distribution of a numeric column |
| `fct_reorder()` | Orders category labels by a numeric value |
| `labs()` | Adds titles, axis labels, and a source |
| `theme_minimal()` | Applies a simple visual theme |
| Setting outside `aes()` | Gives every mark a fixed appearance |

::: {.callout-tip title="ggplot2 is a larger universe"}
The examples teach a small, reusable foundation. The official
[Introduction to ggplot2](https://ggplot2.tidyverse.org/articles/ggplot2.html),
[theme reference](https://ggplot2.tidyverse.org/reference/ggtheme.html), and
[ggplot2 book](https://ggplot2-book.org/) provide more options when a project
needs them.
:::
