---
title: "Visualize One Variable"
---

::: {.callout-note title="Part 4: Data Visualization"}
A chart should begin with a question. We will first describe one column, then
combine columns to visualize comparisons, relationships, and change. The final
chapter in this part introduces geographic questions and maps.
:::

## Learning objectives

By the end of this chapter, you should be able to:

- identify whether a variable is categorical or numeric;
- explain the basic parts of a `ggplot2` chart;
- count categories with a bar chart;
- examine a numeric distribution with a histogram and box plot; and
- recognize unusual values and strongly skewed distributions.

## Begin with one column

Describing one column is the simplest place to begin visual analysis. The
question depends on the kind of variable:

| Variable | Possible question | Useful chart |
|---|---|---|
| Categorical | How many holding records belong to each industry? | Bar chart |
| Numeric | How are holding values distributed? | Histogram |
| Numeric | What are the middle, spread, and unusual values? | Box plot |

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `06-one-variable-visualization.Rmd`. Reuse
`data/gpfg.csv`.
:::

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg.csv") |>
  mutate(market_value_usd_billions = market_value_usd / 1000000000)
```

## Build a chart with ggplot2

`ggplot2` is the tidyverse package for visualization. A basic chart has three
parts:

1. the data;
2. an aesthetic mapping created with `aes()`; and
3. a geometry that determines what is drawn.

`ggplot()` starts the chart. `aes()` identifies the column to display.
`geom_bar()` counts the rows in each category and draws the bars:

```{r}
ggplot(gpfg, aes(y = industry)) +
  geom_bar()
```

This is a complete chart. Beginners need only the data, the mapping, and the
geometry. The gray background, bar color, spacing, and axes are ggplot2
defaults; we do not need to control them before seeing whether the chart
answers the question.

The chart answers: “How many holding records belong to each industry?” It does
not show the number of unique companies or the total market value.

## Order a categorical bar chart

Alphabetical order is useful for finding a name. Ordering by frequency makes a
ranking easier to see. `fct_infreq()` orders categories by how often they
appear:

```{r}
ggplot(gpfg, aes(y = fct_infreq(industry))) +
  geom_bar()
```

The bars are horizontal because the category is mapped to the y-axis. This
gives long industry names more room.

## Improve the same chart with layers

Once the basic chart works, add only the layers that help a reader understand
it. The following version keeps the same data and calculation while adding
color, labels, and a cleaner theme:

```{r}
ggplot(gpfg, aes(y = fct_infreq(industry))) +
  geom_bar(fill = "#0072B2", width = 0.75) +
  labs(
    title = "Industrials appeared most often in the year-end holdings file",
    x = "Number of holding records",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.caption = element_text(color = "#6B7280"),
    plot.title.position = "plot",
    plot.caption.position = "plot"
  )
```

The `+` operator adds each layer to the chart. Here:

- `fill` changes the bar color and `width` changes bar thickness;
- `labs()` adds a title, axis labels, and source;
- `theme_minimal()` changes the overall appearance; and
- `theme()` adjusts a few individual display elements.

None of these appearance controls changes the count. Start with the simple
version, confirm the calculation, and then decide which additions improve the
communication.

::: {.callout-note title="Counting rows is not adding values"}
`geom_bar()` counts rows automatically. In the next chapter, we will prepare a
summary table and use `geom_col()` when the question concerns an amount such as
total market value.
:::

## Examine a numeric distribution

A **distribution** describes which values are common, how widely they vary,
and whether unusual values are present. A histogram divides a numeric column
into ranges called bins and counts the observations in each range:

```{r}
ggplot(gpfg, aes(x = market_value_usd_billions)) +
  geom_histogram(fill = "#0072B2", color = "white") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Most holdings are relatively small and a few are very large. This is a
**right-skewed distribution**: the long tail extends toward the larger values.

The default histogram is a starting point. `geom_histogram()` has optional
arguments for controlling the number and width of bins, but those choices can
wait until a reporting question requires more detail.

## Make a skewed distribution easier to see

`scale_x_log10()` changes the spacing of the x-axis so values that differ by
large ratios are easier to compare:

```{r}
ggplot(gpfg, aes(x = market_value_usd_billions)) +
  geom_histogram(fill = "#0072B2", color = "white") +
  scale_x_log10() +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

The source values have not changed, but the axis spacing has. Always tell
readers when an axis uses a log scale.

::: {.callout-warning title="Check before using a log scale"}
A log scale cannot display zero or negative values. Check whether those values
exist and decide how they should be handled before using the scale.
:::

## Summarize a numeric distribution with a box plot

A box plot gives a compact view of the middle and spread of a numeric column.
It also marks values far from the central part of the distribution:

```{r}
ggplot(gpfg, aes(x = market_value_usd_billions)) +
  geom_boxplot(
    fill = "#56B4E9",
    color = "#1F4E79",
    width = 0.35,
    outlier.color = "#D55E00"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

The line inside the box is the median. The box contains the middle half of the
values. Points beyond the whiskers deserve inspection, but they are not
automatically errors.

## Add clear labels

`labs()` adds information the reader needs, and `theme_minimal()` applies a
simple visual style:

```{r}
ggplot(gpfg, aes(x = market_value_usd_billions)) +
  geom_histogram(fill = "#0072B2", color = "white") +
  scale_x_log10() +
  labs(
    title = "Most reported equity holdings were small relative to the largest",
    x = "Market value in USD billions (log scale)",
    y = "Number of holding records",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.caption = element_text(color = "#6B7280"),
    plot.title.position = "plot",
    plot.caption.position = "plot"
  )
```

The labels name the measure, unit, source, and transformed axis.

::: {.callout-note title="A restrained visual style"}
The charts use a colorblind-aware blue (`#0072B2`) for the main data and an
orange (`#D55E00`) only when a value needs emphasis. A minimal theme, lighter
grid lines, and bold titles create hierarchy without adding decoration. The
hexadecimal color codes simply identify exact colors; you can reuse them
without memorizing how the codes work.
:::

::: {.callout-tip title="ggplot2 is a larger universe"}
The examples in this book teach a small, reusable foundation. ggplot2 can also
control scales, legends, annotations, facets, coordinates, fonts, and many
other design details. Explore the official
[Introduction to ggplot2](https://ggplot2.tidyverse.org/articles/ggplot2.html),
compare the [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html),
or consult the online [ggplot2 book](https://ggplot2-book.org/) when a project
needs more control. You do not need to learn all of these options at once.
:::

## Questions to ask about a one-variable chart

- What does one bar, bin, or point represent?
- Is the chart counting rows, unique entities, or an amount?
- Are missing, zero, or negative values absent?
- Is the axis transformed?
- Which unusual value needs source checking?
- What can this one column not explain?

## Practice

Using `gpfg`:

1. choose one categorical column and create a bar chart of its row counts;
2. choose one numeric column and create a histogram;
3. create a box plot of the same numeric column;
4. add a title, axis labels, and source; and
5. write one visible pattern and one question that requires another column.

## Takeaways

| Function | What it does |
|---|---|
| `ggplot()` | Starts a chart with data |
| `aes()` | Maps a column to an axis or another visual feature |
| `geom_bar()` | Counts rows in each category and draws bars |
| `fct_infreq()` | Orders categories by frequency |
| `geom_histogram()` | Shows the distribution of one numeric variable |
| `geom_boxplot()` | Summarizes the middle, spread, and unusual values |
| `scale_x_log10()` | Displays a numeric axis on a logarithmic scale |
| `labs()` | Adds titles, axis labels, and a source caption |
| `theme_minimal()` | Applies a simple visual theme |

::: {.callout-tip title="Want more control?"}
The official [ggplot2 reference](https://ggplot2.tidyverse.org/reference/)
documents individual functions and their optional arguments. Begin with the
simple layers above and add an option only when it helps answer the question.
:::
