---
title: "Visualize Multiple Variables"
---

## Learning objectives

By the end of this chapter, you should be able to:

- choose a chart that matches a comparison, relationship, or change question;
- make ordered, grouped, stacked, and percentage-stacked bar charts;
- use a dot plot as an alternative to bars;
- compare distributions with box plots;
- examine relationships with scatterplots;
- show change over time with line charts; and
- save a finished chart with `ggsave()`.

## Move beyond description

The previous chapter described one column. Questions involving two or more
columns allow us to compare categories, examine relationships, and follow
change over time:

| Question | Useful starting chart |
|---|---|
| Which categories have the largest values? | Ordered bar chart or dot plot |
| How do categories compare across a few groups? | Grouped bar chart |
| How is a total divided into parts? | Stacked bar chart |
| How do proportions differ? | Percentage-stacked bar chart |
| How do numeric distributions differ across groups? | Box plot |
| Are two numeric variables related? | Scatterplot |
| How does a measure change over time? | Line chart |

::: {.callout-tip title="Continue in the same project"}
Open `djr.Rproj` and create `07-multiple-variable-visualization.Rmd`. Reuse the
latest-year and multi-year GPFG files in `data/`. Save finished charts in
`outputs/`.
:::

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg.csv")
gpfg_panel <- read_csv("data/gpfg_equities_last_10_years.csv")
```

## Compare categories with an ordered bar chart

Ask:

> Which investment markets had the largest total reported market values at
> the end of 2025?

First prepare the table the chart will display:

```{r}
top_countries <- gpfg |>
  group_by(country) |>
  summarise(market_value_usd = sum(market_value_usd)) |>
  arrange(desc(market_value_usd)) |>
  slice_head(n = 10) |>
  mutate(market_value_usd_billions = market_value_usd / 1000000000)

top_countries
```

`geom_col()` draws bars from values that have already been calculated:

```{r}
country_bar <- top_countries |>
  ggplot(
    aes(
      x = market_value_usd_billions,
      y = fct_reorder(country, market_value_usd_billions)
    )
  ) +
  geom_col(fill = "#0072B2", width = 0.72) +
  labs(
    title = "The largest reported equity values were concentrated in a few markets",
    x = "Market value in USD billions",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.caption = element_text(color = "#6B7280"),
    plot.title.position = "plot",
    plot.caption.position = "plot"
  )

country_bar
```

The deep blue gives the book a consistent primary data color. The minimal theme
keeps the numeric grid lines needed for comparison while removing unnecessary
lines behind the country labels. Bold title text creates a clear reading order.

::: {.callout-note title="geom_bar() or geom_col()?"}
`geom_bar()` counts rows from one categorical column. `geom_col()` uses a
numeric value that has already been calculated. Use the function that matches
the question and the table supplied to `ggplot()`.
:::

## Use a dot plot for the same comparison

A dot plot can show the same ranking with less visual weight:

```{r}
top_countries |>
  ggplot(
    aes(
      x = market_value_usd_billions,
      y = fct_reorder(country, market_value_usd_billions)
    )
  ) +
  geom_point(color = "#0072B2", size = 3) +
  labs(
    title = "Ten investment markets with the largest reported equity values",
    x = "Market value in USD billions",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.caption = element_text(color = "#6B7280"),
    plot.title.position = "plot",
    plot.caption.position = "plot"
  )
```

Bars emphasize magnitude from a zero baseline. Dots are compact and make the
positions easy to compare, especially when many category labels are present.

## Compare categories across groups

To demonstrate grouped and stacked bars without overcrowding the chart, select
three regions and three industries:

```{r}
region_industry_counts <- gpfg |>
  filter(
    region %in% c("Asia", "Europe", "North America"),
    industry %in% c("Technology", "Financials", "Industrials")
  ) |>
  count(industry, region)

region_industry_counts
```

### Grouped bars

Grouped bars place the regions beside one another within each industry. They
are useful for direct comparisons across a small number of groups:

```{r}
ggplot(
  region_industry_counts,
  aes(x = industry, y = n, fill = region)
) +
  geom_col(position = "dodge", width = 0.72) +
  scale_fill_manual(
    values = c(
      "Asia" = "#009E73",
      "Europe" = "#E69F00",
      "North America" = "#0072B2"
    )
  ) +
  labs(
    title = "Holding-record counts differed across selected industries and regions",
    x = NULL,
    y = "Number of holding records",
    fill = "Region",
    caption = "Only three selected industries and regions are shown"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.title.position = "plot"
  )
```

`scale_fill_manual()` assigns the same color to each region in every grouped
and stacked example. Blue, orange, and green remain distinguishable for many
readers with common forms of color-vision deficiency. Color identifies groups;
it is not added merely as decoration.

### Stacked bars

Stacked bars emphasize the combined total and how each region contributes to
it:

```{r}
ggplot(
  region_industry_counts,
  aes(x = industry, y = n, fill = region)
) +
  geom_col(width = 0.72) +
  scale_fill_manual(
    values = c(
      "Asia" = "#009E73",
      "Europe" = "#E69F00",
      "North America" = "#0072B2"
    )
  ) +
  labs(
    title = "Selected regions contributed differently to each industry count",
    x = NULL,
    y = "Number of holding records",
    fill = "Region",
    caption = "Only three selected industries and regions are shown"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.title.position = "plot"
  )
```

Only the bottom segment shares a common baseline, so exact comparisons among
the other segments can be difficult. Use grouped bars when direct comparison
is more important than the combined total.

### Percentage-stacked bars

`position = "fill"` makes every bar the same height and displays composition:

```{r}
ggplot(
  region_industry_counts,
  aes(x = industry, y = n, fill = region)
) +
  geom_col(position = "fill", width = 0.72) +
  scale_fill_manual(
    values = c(
      "Asia" = "#009E73",
      "Europe" = "#E69F00",
      "North America" = "#0072B2"
    )
  ) +
  labs(
    title = "Regional composition varied across the selected industries",
    x = NULL,
    y = "Share of selected holding records",
    fill = "Region",
    caption = "Only three selected industries and regions are shown"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.title.position = "plot"
  )
```

This version makes proportions easier to compare but hides differences in the
total number of records. Choose between totals and percentages according to the
reporting question.

## Compare numeric distributions across groups

Adding `industry` to the one-variable box plot allows us to compare holding-
level distributions:

```{r}
ggplot(
  gpfg,
  aes(x = market_value_usd, y = industry)
) +
  geom_boxplot(
    fill = "#56B4E9",
    color = "#1F4E79",
    outlier.color = "#D55E00"
  ) +
  scale_x_log10() +
  labs(
    title = "Holding-level market values varied within every industry",
    x = "Market value in USD (log scale)",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.title.position = "plot"
  )
```

This chart compares individual holding distributions. It does not compare the
total market value of the industries.

## Examine a relationship with a scatterplot

Use a scatterplot when both variables are numeric:

```{r}
ggplot(
  gpfg,
  aes(x = market_value_usd, y = ownership_pct)
) +
  geom_point(color = "#0072B2", alpha = 0.25, size = 1.2) +
  scale_x_log10() +
  labs(
    title = "Market value and ownership percentage varied across holdings",
    x = "Market value in USD (log scale)",
    y = "Ownership (%)",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.title.position = "plot"
  )
```

Each point represents one holding. Look for direction, clusters, and unusual
observations. `alpha = 0.25` makes each point partly transparent, so dense
overlapping areas become visible instead of turning into one solid block. A
visible relationship does not show that one variable caused the other.

## Add a third variable with small multiples

`facet_wrap()` repeats the scatterplot for each region:

```{r}
ggplot(
  gpfg,
  aes(x = market_value_usd, y = ownership_pct)
) +
  geom_point(color = "#0072B2", alpha = 0.25, size = 0.9) +
  scale_x_log10() +
  facet_wrap(~ region) +
  labs(
    title = "The relationship can be inspected separately by region",
    x = "Market value in USD (log scale)",
    y = "Ownership (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.title.position = "plot"
  )
```

Because the panels use the same axes, their patterns can be compared. Faceting
is often clearer than placing many categories in one plotting area.

## Show change over time with a line chart

A line chart is suitable when the horizontal variable has a meaningful order,
usually time. Prepare one value per year:

```{r}
annual <- gpfg_panel |>
  group_by(year) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(year) |>
  mutate(market_value_nok_trillions = market_value_nok / 1000000000000)
```

`geom_line()` connects the annual values, while `geom_point()` keeps the
observed years visible:

```{r}
annual |>
  ggplot(aes(x = year, y = market_value_nok_trillions)) +
  geom_line(color = "#0072B2", linewidth = 1.1) +
  geom_point(color = "#D55E00", size = 2.6) +
  scale_x_continuous(
    breaks = c(2016, 2018, 2020, 2022, 2025)
  ) +
  labs(
    title = "Reported equity market value increased over the ten-year period",
    subtitle = "Year-end holdings in nominal Norwegian kroner",
    x = "Year",
    y = "NOK trillions",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", color = "#1F2933"),
    plot.subtitle = element_text(color = "#52606D"),
    plot.caption = element_text(color = "#6B7280"),
    plot.title.position = "plot",
    plot.caption.position = "plot"
  )
```

The line shows change in nominal year-end value. It does not show investment
returns or how much the fund purchased each year. Do not connect categories
with a line when they have no meaningful order.

## Finish and save a chart

Before publication, check:

- Does the chart answer one clear question?
- Does the title match the calculation?
- Are the population, period, measure, and unit clear?
- Are categories ordered meaningfully?
- Is any transformation disclosed?
- Does color distinguish meaningful groups?
- Is the source named?
- Do the plotted values agree with the analysis table?
- Is an important limitation stated in the chart or story?

`ggsave()` saves a chart object to a file:

```{r}
#| eval: false
ggsave("outputs/country-holdings.png", country_bar)
```

The filename and chart object are enough for this lesson. Width, height, and
resolution are optional controls for later publication requirements.

## Practice

Choose one question involving at least two columns. Create:

1. a precise reporting question;
2. the table needed for the chart;
3. one chart that matches the question;
4. a title, labels, and source; and
5. one sentence describing a limitation.

Then explain why another chart type in this chapter would be less suitable.

## Takeaways

| Function or chart | What it does |
|---|---|
| `geom_col()` | Draws bars from values already calculated |
| `position = "dodge"` | Places grouped bars beside one another |
| `position = "fill"` | Makes equal-height bars for comparing composition |
| `geom_point()` | Creates dot comparisons or scatterplots |
| `geom_boxplot()` | Compares numeric distributions across groups |
| `facet_wrap()` | Creates one panel for each category |
| `geom_line()` | Connects observations across an ordered axis |
| `fct_reorder()` | Orders categories by a numeric value |
| `ggsave()` | Saves a chart to a file |

::: {.callout-tip title="Want more control?"}
Use the official [ggplot2 reference](https://ggplot2.tidyverse.org/reference/)
when a question or publication format requires more control over scales,
labels, themes, or saved-image dimensions.
:::
