---
title: "Class 1: R and RStudio"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

By the end of class, you should be able to explain the roles of R, RStudio,
Projects, packages, and R Markdown—and knit this note to HTML.

## The five pieces

| Piece | Job |
|---|---|
| R | Runs code |
| RStudio | Provides a workspace for writing and running code |
| R Project | Keeps one analysis and its paths together |
| Package | Adds reusable functions |
| R Markdown | Combines notes, code, and results |

The relationship to remember: **R is the engine; RStudio is the workspace
around it.** Install R first, then RStudio.

## Live setup

1. In RStudio, choose **File → New Project → New Directory → New Project**.
2. Name the project `djr`.
3. Create two folders: `data` and `outputs`.
4. Save this file as `01-setup.Rmd` in the project folder.

Your project should look like this:

```text
djr/
├── data/
├── outputs/
├── 01-setup.Rmd
└── djr.Rproj
```

## Packages: install once, load each session

Run these installation commands in the **Console**, usually only once per
computer:

```r
install.packages("tidyverse")
install.packages("rmarkdown")
```

Put this loading command in a document whenever it uses tidyverse functions:

```{r}
library(tidyverse)
```

The tidyverse supplies tools for importing, cleaning, analysing, and
visualizing data. The rmarkdown package turns an `.Rmd` file into a finished
document.

## R Markdown anatomy

- YAML between `---` lines stores document information.
- Markdown is ordinary writing and simple formatting.
- R code goes inside code chunks.
- Knitting runs the chunks in order and creates a new output file.

Try a chunk:

```{r}
2 + 2
```

## Checkpoint

- Which program runs the code?
- Why open `djr.Rproj` before working?
- What is the difference between installing and loading a package?
- What three things can one R Markdown document contain?

## Before next class

Knit this document to HTML. Close RStudio, reopen `djr.Rproj`, and confirm that
the `data` and `outputs` folders appear in the Files pane.

