---
title: "Class 2: R Basics"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Build a small table step by step: values → vectors → tibble → column types →
functions and pipes.

```{r}
library(tidyverse)
```

## 1. Objects and comments

```{r}
# Store a value, then print it
score <- 85
score
```

Read `<-` as “gets.” Use `<-` for assignment in this course; `=` will usually
name an option inside a function.

## 2. Create related vectors

```{r}
student_names <- c("Ana", "Ben", "Chen")
student_ages <- c(20, 21, 22)
student_present <- c(TRUE, FALSE, TRUE)

student_ages + 1
```

The first item in each vector describes the same student, and all three
vectors have the same length.

## 3. Turn those vectors into a tibble

```{r}
students <- tibble(
  name = student_names,
  age = student_ages,
  present = student_present
)

students
```

One row is one student. One column is one variable. One cell is one value.

Read the labels below the column names:

- `<chr>`: text;
- `<dbl>`: numbers; and
- `<lgl>`: `TRUE` or `FALSE`.

## 4. Reuse a vector with functions

```{r}
sum(student_ages)
mean(student_ages)
median(student_ages)
min(student_ages)
max(student_ages)
```

A function is a named instruction. The value inside the parentheses is its
input, or argument.

## 5. Pipe an object into a function

```{r}
student_ages |>
  mean()
```

Read this top to bottom: start with `student_ages`, then calculate its mean.
The older `%>%` pipe is common in existing code, but this course uses `|>`.

## Try it

Create vectors named `book_titles`, `book_pages`, and `book_finished`. Use
them to make a three-row tibble named `books`. State what one row represents.

## Key code

| Code | Use |
|---|---|
| `<-` | Save a value as an object |
| `#` | Add a comment |
| `c()` | Combine values into a vector |
| `tibble()` | Create a table |
| `sum()`, `mean()`, `median()` | Summarize numbers |
| `|>` | Pass a result to the next step |

