---
title: "Class 3: Data Import"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Import one CSV from the project, confirm what arrived, and write a cleaning
plan. Do not clean the table yet.

## Before class

Save `gpfg_messy.csv` inside `djr/data/`. Open `djr.Rproj`, then open this
note from the project folder.

```{r}
library(tidyverse)
```

## Relative path

From the project folder, the file is located at:

```text
data/gpfg_messy.csv
```

This relative path can work on another computer if the project structure is
the same. Avoid paths tied to your personal Desktop or username.

## Import

```{r}
gpfg_raw <- read_csv("data/gpfg_messy.csv")
```

`read_csv()` is a readr function, available after loading the tidyverse. It
needs the file path; `<-` saves the imported tibble as `gpfg_raw`.

## Inspect

```{r}
glimpse(gpfg_raw)
```

Read the output in this order:

1. number of rows and columns;
2. column names;
3. column types such as `<chr>` and `<dbl>`; and
4. example values.

```{r}
gpfg_raw |>
  slice_head(n = 5)
```

The file imported successfully. The amounts and percentages are numeric, but
the reporting date arrived as text. A successful import only means R read the
file; it does not prove that every row and type is ready for analysis.

## Turn observations into questions

Complete this table during class:

| Check | What did we observe? | What might need cleaning? |
|---|---|---|
| Names |  |  |
| Types |  |  |
| Missing values |  |  |
| Repeated rows |  |  |
| Units and definitions |  |  |

## Exit ticket

- What should one row represent?
- Which columns already arrived as numbers?
- Which column should be a date?
- Why keep the downloaded file locally?

## Key code

| Code | Use |
|---|---|
| `read_csv("data/file.csv")` | Import a local CSV |
| `glimpse()` | Inspect names, types, size, and examples |
| `slice_head(n = 5)` | Preview five rows |
