---
title: "Week 3, Part 1: Data Import"
author: "Bin Chen"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

## Today’s workflow

```text
Understand the file → choose an import method → locate the file
→ import it → inspect columns and rows → write a cleaning plan
```

The goal is not simply to make a file appear in R. We need to understand what
arrived and decide whether it is ready for analysis.

## Before class

1. Open `djr.Rproj`.
2. Confirm that `gpfg_messy.csv` is inside the project's `data` folder.
3. Create or open an R Markdown document for Week 3.

`library(tidyverse)` loads the collection of packages used throughout this
course:

```{r}
library(tidyverse)
```

Today's file, `gpfg_messy.csv`, is adapted from the 2025 equity-holdings data
published by Norges Bank Investment Management. It is a CSV with a manageable
number of columns and four deliberately introduced teaching issues. One row is
intended to represent one reported equity holding at the end of 2025.

## 1. Dataset format: CSV

**CSV** means **comma-separated values**. It is a plain-text format for a
rectangular table:

```text
name,city,age
Ana,Hong Kong,20
Ben,Oslo,21
```

- the first line usually contains column names;
- each later line represents one row;
- commas separate the columns; and
- the file contains values, not spreadsheet formatting or formulas.

### Why use CSV?

- **Portable:** R, Python, spreadsheet software, and text editors can read it.
- **Simple:** the file contains the table without hidden formulas or multiple
  worksheets.
- **Easy to preserve:** a journalist can keep the exact file received from a
  source.
- **Suitable for reproducible work:** code can import the same saved file each
  time it runs.

CSV also has limits. It stores one table, does not preserve formatting, and
does not formally store column types. When R reads a CSV, it examines the
values and makes an informed guess about whether each column is text, a
number, a date, and so on. We therefore inspect the result after import.

## 2. Different ways to bring data into R

### A dataset supplied by a package

Some R packages include small datasets for learning or demonstrations. For
example, `mpg` is supplied with ggplot2. Package datasets are convenient for
practice, but a reporting project normally needs an external source.

### A public URL

A CSV can be read directly from a public web address. This is convenient for a
quick check. It is less reliable for a continuing project because the file or
address may change, the website may be unavailable, and knitting requires an
internet connection.

### A local file

For the main workflow, download the source file once, keep it unchanged in the
project, and import that local copy. The local copy preserves the evidence used
in the analysis. In your notes, also record the publisher, original URL,
download date, and any source documentation.

| Method | Best use | Main caution |
|---|---|---|
| Package dataset | Learning and examples | Usually not the reporting source |
| Public URL | Quick access to a stable public file | Requires internet; content may change |
| Local file | Reproducible reporting and class work | You must document where it came from |

## 3. readr and `read_csv()`

The **tidyverse** is a collection of R packages designed to work together.
**readr** is the tidyverse package for importing rectangular text files.
`read_csv()` is the readr function for CSV files.

```r
read_csv("path_to_the_file.csv")
```

A **function** is a named instruction. The function name is `read_csv`; the
file path inside the parentheses is the information the function needs.

Because we already ran `library(tidyverse)`, readr and `read_csv()` are
available. We do not also need `library(readr)`.

The three access patterns now look like this. The `::` in the first example
means “take `mpg` from the `ggplot2` package”:

```r
# Dataset supplied by a package
package_data <- ggplot2::mpg

# CSV at a public URL
web_data <- read_csv(
  "https://raw.githubusercontent.com/binchen19/djr/main/data/gpfg_messy.csv"
)

# CSV saved inside this RStudio Project
local_data <- read_csv("data/file.csv")
```

The next two examples use the same function; only the path changes. We will use
the local pattern for the live demonstration.

::: {.callout-note title="Why `read_csv()` rather than `read.csv()`?"}
`read.csv()` is a valid base R function. This course uses readr's
`read_csv()` because it returns a tidyverse tibble, reports the column types it
detected, and fits the workflow used in the following classes. Learning one
consistent method is enough for now.
:::

`read_csv()` has many optional arguments for unusual files. Begin with the
file path; add another argument only when the file requires it.

## 4. File paths and R Projects

A **path** tells R where a file is located. With `djr.Rproj` open, the RStudio
Project folder is the starting point:

```text
djr/
├── data/
│   └── gpfg_messy.csv
├── outputs/
├── djr.Rproj
└── week-03.Rmd
```

Starting from `djr/`, the path is:

```text
data/gpfg_messy.csv
```

This is a **relative path**: it describes the file's location relative to the
project folder.

Avoid a personal absolute path such as:

```text
/Users/your-name/Desktop/gpfg_messy.csv
```

A relative path is shorter and can work when the project is moved, shared, or
opened on another computer—as long as its folder structure stays the same.

## 5. Import the data

Read the CSV and assign the resulting tibble to `gpfg_raw`:

```{r}
gpfg_raw <- read_csv("data/gpfg_messy.csv")
```

Read this from right to left:

1. `read_csv()` reads the file at the stated path;
2. `<-` assigns the result; and
3. `gpfg_raw` is the R object that now stores the imported table.

The `_raw` suffix reminds us that this object reflects the imported source. We
have not cleaned it yet.

readr may print a message about rows, columns, and detected column types. This
is useful import information, not an error.

### Common import errors

| What you see | Likely cause | What to check |
|---|---|---|
| “file does not exist” or “cannot open file” | R cannot find the path | Is `djr.Rproj` open? Is the file inside `data/`? |
| A similar filename still fails | Spelling, capitals, spaces, or extension differ | Compare the path character by character with the Files pane |
| `read_csv()` cannot be found | The package was not loaded | Run `library(tidyverse)` first |
| `gpfg_raw` cannot be found | The import line did not run successfully | Run the chunks from the beginning and fix the first error |
| Everything appears in one column | The file may use another delimiter | Open the file as text and confirm how columns are separated |
| Unexpected column types | Values are inconsistent or formatted unusually | Inspect examples before deciding how to clean them |

Do not bypass a path problem by manually changing R's starting folder. Open
the correct RStudio Project and keep the file in its intended project folder.

## 6. Inspect columns and rows

Successful import means R read something. It does not mean the table is clean
or appropriate for the reporting question.

### Inspect the columns with `glimpse()`

`glimpse()` shows the table's size, column names, column types, and a few
example values:

```{r}
glimpse(gpfg_raw)
```

Read the output in this order:

1. **Rows and columns:** how large is the table?
2. **Column names:** what variables arrived?
3. **Column types:** does R show `<chr>`, `<dbl>`, or another type?
4. **Example values:** do the examples look plausible?

It displays the table vertically, making a wide dataset easier to inspect. It
shows examples from each column, not every value in the dataset.

The most common type labels in this lesson are:

| Label | Meaning | Example |
|---|---|---|
| `<chr>` | Character text | A company or country name |
| `<dbl>` | Numeric value | A market value or percentage |
| `<date>` | Calendar date | `2025-12-31` |
| `<lgl>` | Logical value | `TRUE` or `FALSE` |

### Inspect the first rows

`slice_head(n = 5)` returns the first five rows for display without changing
`gpfg_raw`:

```{r}
gpfg_raw |>
  slice_head(n = 5)
```

The named argument `n = 5` tells the function how many rows to return.

You may also see:

```{r}
head(gpfg_raw)
```

`head()` is a valid base R function and shows the first six rows by default.
The course normally uses `slice_head()` so the code remains in the tidyverse
style, but students should recognize both.

## 7. Common items to inspect

Use the source documentation and ask:

| Item | Question | What we notice in this file |
|---|---|---|
| Row meaning | What should one row represent? | One reported equity holding at the end of 2025 |
| Column names | Are they clear and easy to type? | Several contain capitals and spaces |
| Column types | Do the types match their meaning? | Amounts are numeric; `Report Date` is text |
| Dates | Is the order day-month-year, month-day-year, or year-month-day? | The source uses day-month-year |
| Missing values | Does `NA` mean unavailable, and are important fields missing? | A targeted check is needed |
| Repeated rows | Is a repeated row an error or a valid observation? | A targeted check is needed |
| Category values | Are spelling and capitalization consistent? | The supplied categories appear consistent |
| Units | What exactly does each number measure? | The dictionary distinguishes NOK, USD, and percentages |

### Turn observations into a cleaning plan

Do not clean everything automatically. Fix only problems supported by an
observation:

| Observation | Planned action |
|---|---|
| Four names are awkward | Rename those four columns |
| `Report Date` is character text | Convert it after confirming the date order |
| An important field may be missing | Display and interpret the affected row |
| A row may be repeated | Inspect the complete rows before removing anything |

Week 3, Part 2 applies this plan to the same `gpfg_raw` object.

## Checkpoint questions

1. What is stored—and not stored—in a CSV?
2. Why is a saved local file preferable for a continuing reporting project?
3. What is the relationship among tidyverse, readr, and `read_csv()`?
4. Why does `data/gpfg_messy.csv` work when `djr.Rproj` is open?
5. What does one row represent?
6. Which problems can `glimpse()` reveal, and which require targeted checks?

## Key code

| Code | Use |
|---|---|
| `library(tidyverse)` | Loads readr and the other core tidyverse packages |
| `read_csv("data/file.csv")` | Imports a local CSV as a tibble |
| `<-` | Assigns the imported table to an object |
| `glimpse()` | Inspects size, names, types, and example values |
| `slice_head(n = 5)` | Displays the first five rows |
| `head()` | Base R alternative that displays the first six rows by default |
