---
title: "Class 4: Data Cleaning"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Continue with the object imported in Class 3. We will solve four clear
problems: awkward names, a text date, an incomplete row, and a duplicate.

```{r}
library(tidyverse)

gpfg_raw <- read_csv("data/gpfg_messy.csv")
glimpse(gpfg_raw)
```

## 1. Rename selected columns

```{r}
gpfg <- gpfg_raw |>
  rename(
    year = Year,
    report_date = `Report Date`,
    company = `Company Name`,
    market_value_nok = `Market Value NOK`
  )
```

Read each line as `new_name = old_name`. Columns whose names were already
clear are left alone. `year` may display as `<dbl>`; it is already numeric and
does not need conversion for this work.

## 2. Convert the date

The source uses day-month-year order, such as `31/12/2025`:

```{r}
gpfg <- gpfg |>
  mutate(report_date = dmy(report_date))

glimpse(gpfg)
```

`mutate()` changes the column. `dmy()` converts day-month-year text to a real
date. Confirm the order from the source; never guess an ambiguous date.

## 3. Inspect the missing row

```{r}
gpfg |>
  filter(is.na(company))
```

This deliberately incomplete row has no company, country, industry, or value,
so it cannot represent a holding. After checking the whole row, remove it:

```{r}
gpfg <- gpfg |>
  filter(!is.na(company))
```

`!` means “not.” Do not automatically remove missing values in other data;
first determine what they mean.

## 4. Check a possible duplicate

```{r}
gpfg |>
  count(company, country, industry) |>
  filter(n > 1)
```

```{r}
gpfg |>
  filter(company == "3P Learning Ltd")
```

The two rows are completely identical, so keep one copy:

```{r}
gpfg <- gpfg |>
  distinct()
```

## 5. Verify and save

```{r}
glimpse(gpfg)

gpfg |>
  slice_head(n = 5)

write_csv(gpfg, "data/gpfg_2025.csv")
```

The result should have 7,201 rows and 11 columns. Keep the original messy CSV;
later classes use the new clean file.

## Exit ticket

- What evidence justified each cleaning step?
- What is the difference between renaming and changing a type?
- Why can we remove this missing row and this duplicate?
- Why should the raw CSV remain unchanged?

## Key code

| Code | Use |
|---|---|
| `rename(new = old)` | Rename selected columns |
| `mutate()` + `dmy()` | Convert a day-month-year column |
| `filter(is.na(...))` | Inspect missing values |
| `filter(!is.na(...))` | Keep non-missing rows |
| `count()` | Find repeated identifiers |
| `distinct()` | Remove a confirmed exact duplicate |
| `write_csv()` | Save the clean table |
