---
title: "Week 3, Part 2: Data Cleaning"
author: "Bin Chen"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

## From inspection to cleaning

**Data cleaning** means identifying, investigating, and correcting problems
that could make an analysis inaccurate or difficult to reproduce. Its goal is
not to make every value look neat. The goal is a table that is fit for the
reporting task: rows and columns have clear meanings, values have appropriate
types, missing and repeated records have been investigated, and every change
is recorded in code.

Keep the imported source unchanged. The cleaning code should create a separate
analysis-ready result that another person can reproduce and check.

Part 1 imported `gpfg_messy.csv` and produced a four-step plan. We will now:

1. rename four awkward columns;
2. convert one text column to a date;
3. inspect and handle one incomplete row;
4. inspect and remove one confirmed duplicate; and
5. verify and save a separate clean file.

Each R Markdown document must contain the code it needs to run from the
beginning, so start by recreating the imported object:

```{r}
library(tidyverse)

gpfg_raw <- read_csv("data/gpfg_messy.csv")
```

Keep `gpfg_raw` unchanged. We will store the cleaned result in a new object
called `gpfg`.

## 1. Rename selected columns

The inspection found four names that contain capitals and spaces. Use
`rename()` with `new_name = old_name`. Put backticks around an old name that
contains spaces:

```{r}
gpfg <- gpfg_raw |>
  rename(
    year = Year,
    report_date = `Report Date`,
    company = `Company Name`,
    market_value_nok = `Market Value NOK`
  )

glimpse(gpfg)
```

The new names follow `snake_case`: lowercase words separated by underscores.

The remaining names were already clear, so we did not change them. Cleaning
should respond to observed problems, not rewrite everything.

## 2. Correct the date type

The values in `report_date` look like `31/12/2025`. The source confirms that
the order is day, month, year.

Use `mutate(column = calculation)` to create or replace a column while keeping
the other rows and columns. Here `dmy()` supplies the calculation that
interprets the date order:

```{r}
gpfg <- gpfg |>
  mutate(report_date = dmy(report_date))

glimpse(gpfg)
```

The type should now be `<date>` rather than `<chr>`.

- `dmy()` reads day-month-year;
- `mdy()` reads month-day-year; and
- `ymd()` reads year-month-day.

Do not choose a date function by guessing. A value such as `03/04/2025` is
ambiguous without source documentation.

`year` may appear as `<dbl>`. It is already numeric and works for the tasks in
this course, so no conversion is needed merely to make the label look
different.

## 3. Inspect missing values

R represents missing information with `NA`. Missing is different from zero:

- `0` is a known value;
- `NA` means a value is not available.

### How `filter()` works

`filter()` keeps rows where a condition is `TRUE`:

```r
table |>
  filter(condition)
```

If we only print the result, the original table does not change. If we assign
the pipeline back with `<-`, the working object is updated to contain only the
kept rows.

| Condition | Meaning |
|---|---|
| `is.na(company)` | Company is missing |
| `!is.na(company)` | Company is not missing |
| `n > 1` | Count is greater than one |
| `company == "3P Learning Ltd"` | Company equals this exact text |

`!` means “not,” `>` means “greater than,” and `==` checks equality.

`is.na()` asks whether a value is missing. Use it inside `filter()` to display
rows with a missing company. Because the result is not assigned, `gpfg` does
not change:


```{r}
gpfg |>
  filter(is.na(company))
```

The result is deliberately incomplete: it has no company, country, industry,
or market value, so it cannot represent the holding described by the data
dictionary. To keep non-missing company rows, put `!` before the condition and
assign the result back to `gpfg`:

```{r}
gpfg <- gpfg |>
  filter(!is.na(company))
```

The working object now keeps only rows whose company is not missing.

Do not automatically remove every row containing `NA`. In another dataset, a
missing response may be valid, important, or impossible to recover. First ask
what the field means and whether the row can still answer the question.

## 4. Inspect possible duplicates

The intended row meaning is one reported holding at the end of 2025.
`count()` reports how often each combination of selected values appears. Use it
to flag possible repetition:

```{r}
gpfg |>
  count(company, country, industry) |>
  filter(n > 1)
```

The function creates `n`, the number of times each combination appears. It
flags a possible problem; it does not prove the records are identical.

Inspect the complete source rows:

```{r}
gpfg |>
  filter(company == "3P Learning Ltd")
```

Every displayed value is identical. We can therefore keep one copy of each
complete row:

```{r}
gpfg <- gpfg |>
  distinct()
```

A repeated company name alone would not be enough evidence. The same company
could appear in different markets, industries, or years.

## 5. Verify the result

Recheck the structure and preview:

```{r}
glimpse(gpfg)

gpfg |>
  slice_head(n = 5)
```

Now verify the two issues we removed:

```{r}
gpfg |>
  filter(is.na(company))
```

```{r}
gpfg |>
  count(company, country, industry) |>
  filter(n > 1)
```

Both results should be empty. The cleaned table should contain 7,201 rows and
11 columns.

## 6. Save without overwriting the source

`write_csv(table, "path")` saves a table as a CSV. Give it a new filename
rather than overwriting the imported source:

```{r}
write_csv(gpfg, "data/gpfg_2025.csv")
```

The two files have different jobs:

| File | Purpose |
|---|---|
| `gpfg_messy.csv` | Preserved imported source and evidence of the original problems |
| `gpfg_2025.csv` | Clean table used in later analysis |

Keeping the original makes the decisions reviewable and allows the cleaning
code to be rerun from the beginning.

## Complete Week 3 workflow

```text
CSV source
  → local file in an R Project
  → read_csv()
  → glimpse() and slice_head()
  → cleaning plan
  → rename(), mutate(), filter(), distinct()
  → verification
  → write_csv() to a new file
```

## Practice

Restart R and knit both Week 3 notes from the beginning. Then explain:

1. what one source row should represent;
2. the observation that justified each cleaning step;
3. the difference between renaming a column and changing its type;
4. why the missing row could be removed in this example;
5. why the repeated rows were confirmed before `distinct()`; and
6. the difference between printing a filtered result and assigning it; and
7. why the raw file was not overwritten.

## Key code

| Code | Use |
|---|---|
| `rename(new = old)` | Renames selected columns |
| `mutate()` | Creates or changes a column |
| `dmy()` | Reads day-month-year text as a date |
| `is.na()` | Identifies a missing value |
| `filter()` | Keeps rows where a condition is true; assignment saves the subset |
| `count()` | Counts repeated combinations |
| `distinct()` | Keeps each complete row once after verification |
| `write_csv()` | Saves a new clean CSV |
