---
title: "Class 5: Data Analysis"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Import five annual files, combine them, learn what the combined table contains,
and answer trend, description, ranking, and comparison questions.

## 1. Review import

```{r}
library(tidyverse)

gpfg_2021 <- read_csv("data/gpfg_2021.csv")
gpfg_2022 <- read_csv("data/gpfg_2022.csv")
gpfg_2023 <- read_csv("data/gpfg_2023.csv")
gpfg_2024 <- read_csv("data/gpfg_2024.csv")
gpfg_2025 <- read_csv("data/gpfg_2025.csv")
```

Before combining: compare `glimpse(gpfg_2021)` and `glimpse(gpfg_2025)`.
The names and types should match, and both files should contain `year`.

## 2. Stack the annual rows

```{r}
gpfg <- bind_rows(
  gpfg_2021,
  gpfg_2022,
  gpfg_2023,
  gpfg_2024,
  gpfg_2025
)

gpfg |>
  count(year)
```

`bind_rows()` is for the same columns and different observations. One row still
represents one reported holding at one year-end.

```{r}
write_csv(gpfg, "data/gpfg_5_years.csv")
```

## 3. Know the data before asking questions

```{r}
gpfg |>
  summarise(
    first_year = min(year),
    last_year = max(year),
    number_of_years = n_distinct(year)
  )
```

```{r}
gpfg |>
  distinct(region) |>
  arrange(region)
```

```{r}
gpfg |>
  count(country, sort = TRUE)
```

```{r}
gpfg |>
  summarise(
    missing_company = sum(is.na(company)),
    missing_value = sum(is.na(market_value_nok))
  )
```

Checkpoint: category values and record counts describe the file. They do not
yet answer which market has the greatest value.

## 4. Trend question

Question: How did total reported market value change from 2021 to 2025?

```{r}
annual_summary <- gpfg |>
  group_by(year) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(year)

annual_summary
```

After `summarise()`, one row represents one annual snapshot.

## 5. Describe the latest year

```{r}
gpfg_latest <- gpfg |>
  filter(year == max(year))

gpfg_latest |>
  summarise(
    total_nok = sum(market_value_nok),
    median_nok = median(market_value_nok),
    maximum_nok = max(market_value_nok)
  )
```

## 6. Rank individual holdings

```{r}
largest_holdings <- gpfg_latest |>
  select(company, country, industry, market_value_nok) |>
  arrange(desc(market_value_nok)) |>
  slice_head(n = 10)

largest_holdings
```

## 7. Compare markets

```{r}
country_summary <- gpfg_latest |>
  group_by(country) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(desc(market_value_nok))

country_summary |>
  slice_head(n = 10)
```

## Reporting language

Better: “The United States had the largest reported market value in NBIM's
2025 year-end equity-holdings file.”

Avoid language about purchases, flows, returns, or causes unless additional
evidence supports it.

## Exit ticket

- What does one row represent before and after the annual summary?
- Why did the five annual files need a `year` column before `bind_rows()`?
- Which result is a description, ranking, comparison, or trend?
- Write one supported finding and one follow-up reporting question.

