---
title: "Class 5: Analysis I"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Use the clean 2025 table to answer description, ranking, comparison, and
relationship questions. End by identifying the one major question this file
cannot answer: change over time.

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_2025.csv")
```

One row represents one reported equity holding at the end of 2025.

## Question workflow

1. State what one row represents.
2. Write a question with a measure, group, and period.
3. Identify the required columns.
4. Check their types, units, and missing values.
5. Calculate and inspect the result.
6. Write only what the evidence supports.

## 1. Know the table

```{r}
glimpse(gpfg)
```

```{r}
gpfg |>
  distinct(region) |>
  arrange(region)
```

```{r}
gpfg |>
  count(region) |>
  arrange(desc(n))
```

Record count describes rows; it does not measure their total value.

## 2. Description

Question: What were the total and typical values of individual holdings?

```{r}
holding_summary <- gpfg |>
  summarise(
    total_nok = sum(market_value_nok),
    mean_nok = mean(market_value_nok),
    median_nok = median(market_value_nok),
    minimum_nok = min(market_value_nok),
    maximum_nok = max(market_value_nok)
  )

holding_summary
```

Compare mean and median. A few large holdings can pull the mean upward.

## 3. Ranking

Question: Which ten individual holdings had the largest values?

```{r}
largest_holdings <- gpfg |>
  select(company, country, industry, market_value_nok) |>
  arrange(desc(market_value_nok)) |>
  slice_head(n = 10)

largest_holdings
```

## 4. Comparison

Question: Which investment markets had the largest total reported values?

```{r}
country_summary <- gpfg |>
  group_by(country) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(desc(market_value_nok))

country_summary |>
  slice_head(n = 10)
```

Before `summarise()`, one row is one holding. After it, one row is one
investment market.

```{r}
country_summary <- country_summary |>
  mutate(
    share_percent = market_value_nok / sum(market_value_nok) * 100
  )

country_summary |>
  slice_head(n = 10)
```

## 5. Prepare a relationship question

Question: Did market value and ownership percentage vary together?

```{r}
relationship_data <- gpfg |>
  select(company, market_value_usd, ownership_pct)

relationship_data |>
  slice_head(n = 5)
```

The columns exist, but thousands of printed rows are difficult to compare.
Visualization I will turn them into a scatterplot.

## Reporting language

Better: “The United States had the largest reported market value in NBIM's
2025 year-end equity-holdings file.”

Avoid language about purchases, flows, returns, or causes unless additional
evidence supports it.

## The limitation

The 2025 file cannot tell us whether a value rose or fell. A trend requires at
least two time points. Analysis II will add four earlier annual files.

## Exit ticket

- What does one row represent before and after the country summary?
- Which result is a description, ranking, or comparison?
- Why can count and amount produce different rankings?
- Write one supported finding and one trend question this file cannot answer.
