---
title: "Class 6: Analysis II"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Add the missing time dimension: import five annual snapshots, compare their
structures, combine their rows, check the result, and answer trend questions.

## 1. Import the annual files

```{r}
library(tidyverse)

gpfg_2021 <- read_csv("data/gpfg_2021.csv")
gpfg_2022 <- read_csv("data/gpfg_2022.csv")
gpfg_2023 <- read_csv("data/gpfg_2023.csv")
gpfg_2024 <- read_csv("data/gpfg_2024.csv")
gpfg_2025 <- read_csv("data/gpfg_2025.csv")
```

Before combining, compare the names and types:

```{r}
glimpse(gpfg_2021)
```

```{r}
glimpse(gpfg_2025)
```

Both files should describe the same variables and already contain `year`.

## 2. Stack the annual rows

```{r}
gpfg <- bind_rows(
  gpfg_2021,
  gpfg_2022,
  gpfg_2023,
  gpfg_2024,
  gpfg_2025
)
```

`bind_rows()` is for tables with the same columns and different observations.
One row still represents one holding at one year-end.

## 3. Check the combination

```{r}
gpfg |>
  count(year)
```

```{r}
gpfg |>
  filter(is.na(company) | is.na(market_value_nok))
```

```{r}
gpfg |>
  distinct(region) |>
  arrange(region)
```

Checkpoint: five years are present; record counts differ; important fields are
complete; and region labels remain understandable.

```{r}
write_csv(gpfg, "data/gpfg_5_years.csv")
```

## 4. Overall trend

Question: How did total reported market value change from 2021 to 2025?

```{r}
annual_summary <- gpfg |>
  group_by(year) |>
  summarise(
    holding_records = n(),
    market_value_nok = sum(market_value_nok)
  ) |>
  arrange(year)

annual_summary
```

After `summarise()`, one row represents one annual snapshot.

## 5. Comparative trend

Question: How did value change for the United States, China, and Japan?

```{r}
selected_countries <- c("United States", "China", "Japan")

country_trends <- gpfg |>
  filter(country %in% selected_countries) |>
  group_by(year, country) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(country, year)

country_trends
```

One row represents one selected market in one year. `%in%` checks whether each
country belongs to the selected vector.

## Reporting caution

The annual values are year-end holdings, not yearly purchases or returns.
Changes may reflect prices, exchange rates, purchases, sales, portfolio
coverage, or classifications. The table shows the pattern, not its cause.

## Exit ticket

- Why check annual structures before `bind_rows()`?
- What does one row represent before and after the annual summary?
- Why keep record count beside total value?
- Write one supported trend finding and one explanation requiring more
  reporting.
