---
title: "Class 7: Visualization I"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Move from a reporting question to a checked table and then to a basic chart.
Add one layer at a time so every change has a reason.

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_5_years.csv")

gpfg_latest <- gpfg |>
  filter(year == 2025)
```

## The routine

1. State the question.
2. Choose the needed columns.
3. Calculate and inspect the plotting table.
4. Assign x, y, and any other data mappings.
5. Make the simplest plot.
6. Add labels and style one layer at a time.

## 1. A trend question

Question: How did the total reported market value of NBIM's equity holdings
change from 2021 to 2025?

Needed columns: `year` and `market_value_nok`.

```{r}
annual_summary <- gpfg |>
  group_by(year) |>
  summarise(market_value_nok = sum(market_value_nok)) |>
  arrange(year)

annual_summary
```

Plan: x = time, y = annual value. No color or size variable is needed.

```{r}
ggplot(annual_summary, aes(x = year, y = market_value_nok)) +
  geom_line()
```

Add points so the five observations are visible:

```{r}
ggplot(annual_summary, aes(x = year, y = market_value_nok)) +
  geom_line() +
  geom_point()
```

Add meaning:

```{r}
ggplot(annual_summary, aes(x = year, y = market_value_nok)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Reported equity value increased over the five-year period",
    x = NULL,
    y = "Market value (NOK)",
    caption = "Source: Norges Bank Investment Management"
  )
```

Add restrained appearance last:

```{r}
ggplot(annual_summary, aes(x = year, y = market_value_nok)) +
  geom_line(color = "#007C83", linewidth = 1) +
  geom_point(color = "#007C83", size = 2.5) +
  labs(
    title = "Reported equity value increased over the five-year period",
    x = NULL,
    y = "Market value (NOK)",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

The fixed color is outside `aes()` because it does not represent another
column.

## 2. A ranking question

Question: Which ten investment markets accounted for the largest total market
value of NBIM's reported equity holdings in 2025?

```{r}
top_countries <- gpfg_latest |>
  group_by(country) |>
  summarise(market_value_usd = sum(market_value_usd)) |>
  arrange(desc(market_value_usd)) |>
  slice_head(n = 10)

top_countries
```

```{r}
ggplot(top_countries, aes(x = market_value_usd, y = fct_reorder(country, market_value_usd))) +
  geom_col()
```

```{r}
ggplot(top_countries, aes(x = market_value_usd, y = fct_reorder(country, market_value_usd))) +
  geom_col(fill = "#D95F59") +
  labs(
    title = "The United States led reported equity value by a wide margin in 2025",
    x = "Market value (USD)",
    y = NULL,
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

## Optional if time: relationship question

```{r}
ggplot(
  gpfg_latest,
  aes(x = market_value_usd, y = ownership_pct)
) +
  geom_point(alpha = 0.35) +
  labs(
    title = "Ownership and market value vary widely across holdings",
    x = "Market value (USD)",
    y = "Ownership (%)"
  ) +
  theme_minimal()
```

## Exit ticket

- Which table was checked before each plot?
- Why is `year` on x for the line chart?
- What did each added layer change?
- Why is the fixed bar color outside `aes()`?
