---
title: "Class 8: Visualization II"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Start from familiar charts and add complexity only when the question requires
another group, composition, distribution, or panel.

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_5_years.csv")

gpfg_latest <- gpfg |>
  filter(year == 2025)
```

## Mapping versus setting color

```r
geom_point(color = "#0072B2")
```

One fixed appearance: outside `aes()`.

```r
geom_point(aes(color = region))
```

Color represents a column: inside `aes()` and a legend is created.

## 1. Compare several trends

Question: How did the total reported market value of NBIM's equity holdings in
the United States, China, and Japan change from 2021 to 2025?

```{r}
selected_countries <- c("United States", "China", "Japan")

country_trends <- gpfg |>
  filter(country %in% selected_countries) |>
  group_by(year, country) |>
  summarise(market_value_nok = sum(market_value_nok))

country_trends
```

Plan: x = `year`, y = `market_value_nok`, color = `country`.

```{r}
ggplot(
  country_trends,
  aes(x = year, y = market_value_nok, color = country)
) +
  geom_line()
```

```{r}
ggplot(
  country_trends,
  aes(x = year, y = market_value_nok, color = country)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "United States" = "#0072B2",
    "China" = "#D55E00",
    "Japan" = "#009E73"
  )) +
  labs(
    title = "Three investment markets followed different paths",
    x = NULL,
    y = "Market value (NOK)",
    color = "Market",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

## 2. Grouped versus stacked bars

Question: Within the largest industries, how did the total market value of
NBIM's reported equity holdings compare across regions in 2025?

```{r}
selected_industries <- c("Technology", "Financials", "Industrials")

region_industry <- gpfg_latest |>
  filter(industry %in% selected_industries) |>
  group_by(industry, region) |>
  summarise(market_value_usd = sum(market_value_usd))

region_industry
```

Default bars stack the regional contributions:

```{r}
ggplot(
  region_industry,
  aes(x = industry, y = market_value_usd, fill = region)
) +
  geom_col()
```

Change only the position to compare regional bars directly:

```{r}
ggplot(
  region_industry,
  aes(x = industry, y = market_value_usd, fill = region)
) +
  geom_col(position = "dodge") +
  labs(
    title = "Regional values within four large industries",
    x = NULL,
    y = "Market value (USD)",
    fill = "Region",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

- Stacked: how groups contribute to a total.
- Grouped: direct comparison within each category.
- `position = "fill"`: compare proportions rather than total amounts.

## Optional if time: separate crowded groups

```{r}
ggplot(
  gpfg_latest,
  aes(x = market_value_usd, y = ownership_pct)
) +
  geom_point(alpha = 0.25) +
  facet_wrap(vars(region)) +
  scale_y_log10() +
  labs(
    title = "Ownership and value by region",
    x = "Market value (USD, log scale)",
    y = "Ownership (%)",
    caption = "Source: Norges Bank Investment Management"
  ) +
  theme_minimal()
```

## Exit ticket

- When does color belong inside `aes()`?
- Which question favors stacked bars? Which favors grouped bars?
- What new information did each advanced layer represent?
- Which chart would you keep for a news story, and why?
