---
title: "Class 9: Join and Reshape"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Answer a question that cannot be answered from either source table alone:

> How did the total reported market value of NBIM's equity holdings change by
> region from 2021 to 2025?

```{r}
library(tidyverse)

country_wide <- read_csv("data/gpfg_country_wide.csv")
country_lookup <- read_csv("data/gpfg_country_lookup.csv")
```

Inspect both tables. Ask what one row represents and which needed variable is
missing from each.

```{r}
glimpse(country_wide)
```

```{r}
glimpse(country_lookup)
```

## 1. Draw the two shapes

Wide source:

| country | 2021 | 2022 | 2023 | 2024 | 2025 |
|---|---:|---:|---:|---:|---:|

Long target:

| country | year | market_value_nok |
|---|---:|---:|

Long data make `year` available for grouping, filtering, and charting.

## 2. Pivot the year columns into rows

```{r}
country_year <- country_wide |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "market_value_nok"
  )

country_year
```

The old column names became text values. Correct the type:

```{r}
country_year <- country_year |>
  mutate(year = parse_integer(year))

glimpse(country_year)
```

One row now represents one country in one year.

```{r}
country_year |>
  group_by(year) |>
  summarise(missing_values = sum(is.na(market_value_nok)))
```

An empty cell in the wide source becomes `NA`, not a known zero.

## 3. Check the join key

`country` should be unique in the lookup table:

```{r}
country_lookup |>
  count(country) |>
  filter(n > 1)
```

Check countries without a region match:

```{r}
country_year |>
  distinct(country) |>
  anti_join(country_lookup, by = "country")
```

Do not skip these empty-looking results; they document that the proposed key
works.

## 4. Add region with a left join

```{r}
country_year <- country_year |>
  left_join(country_lookup, by = "country")

glimpse(country_year)
```

Checkpoint: one row should still represent one country-year. The join adds a
column, not more observations.

## 5. Answer the question

```{r}
region_year <- country_year |>
  group_by(year, region) |>
  summarise(market_value_nok = sum(market_value_nok, na.rm = TRUE)) |>
  arrange(year, desc(market_value_nok))

region_year
```

## Optional: wide output for a two-year table

```{r}
country_year |>
  filter(year %in% c(2024, 2025)) |>
  select(country, year, market_value_nok) |>
  pivot_wider(
    names_from = year,
    values_from = market_value_nok
  )
```

## Exit ticket

- Why did the data need to become long?
- What is the join key?
- Why check duplicate keys and unmatched keys before joining?
- What does one row represent before and after the regional summary?
