---
title: "Class 10: Mapping"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today

Question: Where was the market value of NBIM's reported equity holdings
concentrated at the end of 2025?

Use the Chapter 9 join workflow to connect one value per investment market to
world boundary coordinates, then build the map one layer at a time.

```{r}
library(tidyverse)

gpfg <- read_csv("data/gpfg_5_years.csv")
world <- read_csv("data/world_boundaries.csv")

gpfg_latest <- gpfg |>
  filter(year == 2025)
```

## 1. Prepare one value per place

```{r}
country_summary <- gpfg_latest |>
  group_by(country) |>
  summarise(market_value_usd = sum(market_value_usd)) |>
  mutate(market_value_usd_billions = market_value_usd / 1000000000)

country_summary |>
  arrange(desc(market_value_usd_billions)) |>
  slice_head(n = 5)
```

## 2. Make names match and check them

```{r}
country_summary <- country_summary |>
  mutate(
    map_name = recode(
      country,
      "Russia" = "Russian Federation",
      "South Korea" = "Republic of Korea",
      "Türkiye" = "Turkey"
    )
  )

map_names <- world |>
  distinct(name_long)

country_summary |>
  anti_join(map_names, by = c("map_name" = "name_long"))
```

## 3. Join values to boundaries

```{r}
world_holdings <- world |>
  left_join(
    country_summary,
    by = c("name_long" = "map_name")
  )
```

The market value repeats across coordinate rows. This is necessary to fill the
shape; do not sum the value after this join.

## 4. Basic map

```{r}
ggplot(
  world_holdings,
  aes(
    x = longitude,
    y = latitude,
    group = polygon_group,
    fill = market_value_usd_billions
  )
) +
  geom_polygon()
```

## 5. Add one improvement at a time

```{r}
ggplot(
  world_holdings,
  aes(
    x = longitude,
    y = latitude,
    group = polygon_group,
    fill = market_value_usd_billions
  )
) +
  geom_polygon(color = "white", linewidth = 0.1) +
  coord_quickmap(expand = FALSE) +
  scale_fill_gradient(
    low = "#DCEAF4",
    high = "#006D77",
    trans = "sqrt",
    na.value = "#E6E8EB"
  ) +
  labs(
    title = "Reported equity value was concentrated in a limited set of markets",
    subtitle = "Year-end 2025 market value by NBIM investment market",
    fill = "USD billions",
    caption = "Sources: Norges Bank Investment Management and Natural Earth\nGray indicates no matched value"
  ) +
  theme_void() +
  theme(
    plot.caption = element_text(hjust = 0),
    plot.margin = margin(12, 18, 12, 18)
  )
```

Gray means no matched value in the joined table, not necessarily zero. The
source field is an investment market, not necessarily company incorporation,
workers, or revenue.

## Exit ticket

- Why summarize before joining to a map?
- Why preserve the original country name?
- Why can a bar chart be better than a map for close comparisons?
- What does every color—including gray—mean?
