---
title: "Class 11: Billboard Hot 100"
author: "Bin Chen"
output:
  html_document:
    toc: true
    theme: cosmo
---

## Today's new challenge

One row is one song on one weekly chart—not one unique song. A title alone is
also not a reliable song identifier. We must define both the unit and the
meaning of “success” before ranking artists.

```{r}
library(tidyverse)

hot100 <- read_csv("data/hot100.csv", na = "-") |>
  rename(
    date = Date,
    song = Song,
    artist = Artist,
    rank = Rank,
    peak_position = `Peak Position`,
    weeks_in_chart = `Weeks in Charts`
  )

glimpse(hot100)
```

The source uses `-` when weeks are unavailable. `na = "-"` tells `read_csv()`
to store those entries as proper missing values instead of text.

## 1. Test whether title alone is unique

```{r}
hot100 |>
  distinct(artist, song) |>
  count(song, name = "number_of_artists") |>
  filter(number_of_artists > 1) |>
  arrange(desc(number_of_artists)) |>
  slice_head(n = 10)
```

For this analysis, define one song as an `artist`–`song` combination. State
that rule; remixes and changing credits can make it imperfect.

## 2. Define period and threshold

```{r}
chart_period <- hot100 |>
  filter(date >= "2000-01-01")

number_ones <- chart_period |>
  filter(rank == 1)
```

“Appeared in the Hot 100,” “reached the top 10,” and “reached number one” are
different thresholds.

## 3. Compare definitions of success

Distinct Hot 100 songs:

```{r}
artist_hot100_songs <- chart_period |>
  distinct(artist, song) |>
  count(artist, name = "hot100_songs") |>
  arrange(desc(hot100_songs))

artist_hot100_songs |>
  slice_head(n = 10)
```

Weekly number-one appearances:

```{r}
artist_weeks <- number_ones |>
  count(artist, name = "number_one_weeks") |>
  arrange(desc(number_one_weeks))

artist_weeks |>
  slice_head(n = 10)
```

Distinct number-one songs:

```{r}
artist_songs <- number_ones |>
  distinct(artist, song) |>
  count(artist, name = "number_one_songs") |>
  arrange(desc(number_one_songs))

artist_songs |>
  slice_head(n = 10)
```

The rankings answer different questions. The column name, chart title, and
finding must say whether the code counted weekly rows or distinct songs.

## 4. Keep artist and song together

```{r}
song_weeks <- number_ones |>
  count(artist, song, name = "number_one_weeks") |>
  arrange(desc(number_one_weeks))

song_weeks |>
  slice_head(n = 10)
```

Grouping only by title could combine unrelated songs with the same name.

## Try it

Choose a decade. Calculate distinct Hot 100 songs, weekly number-one
appearances, and distinct number-one songs by artist. Compare the top ten and
write one sentence explaining why the rankings differ.

## Reporting checklist

- What does one source row represent?
- What columns define a distinct entity?
- What period and threshold did you choose?
- What exactly does the metric count?
- Do artist-credit variations affect the result?
